---
title: "The data a tulpa model expects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The data a tulpa model expects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260529)
```

```{r load, message = FALSE}
library(tulpa)
```

## Expected structure

Every tulpa fit starts from two things: a formula and a data frame. The
formula names the response, the fixed effects, the grouping factors for
random effects, and any special structure (an offset, a spatial unit).
The data frame holds one row per observation, with one column for each
name the formula mentions. `tulpa()` walks the formula, pulls the matching
columns out of the data, and turns them into the numeric pieces an
inference backend consumes.

Two functions do that translation, and both are exported so you can call
them on their own. `tulpa_parse_formula()` reads the formula and reports
its parts. `tulpa_build_model_data()` takes that parsed object plus the
data and returns the design. Looking at their output is the fastest way to
see what tulpa thinks your model is before you commit to a fit.

The formula grammar follows lme4 and glmmTMB, so most mixed-model
formulas you already write carry over. Fixed effects sit on the right of
`~` as an additive chain. A random intercept is `(1 | g)`. A random
intercept and slope sharing a correlation is `(1 + x | g)`; the double bar
`(1 + x || g)` keeps the same two effects but drops the correlation
between them. An `offset(...)` term enters a known per-row shift on the
linear-predictor scale. A `spatial(col)` term names the column that maps
each row to a spatial unit, with the field structure itself arriving
through the `spatial=` argument.

Here is a formula that uses several of those at once. Print the parsed
object to read it back.

```{r parse-show}
pf <- tulpa_parse_formula(y ~ x + z + (1 | g) + (1 + x || site))
pf
```

The print method separates the response from the fixed formula and lists
each random-effect term in its canonical form. The bar operator is shown
as `|` or `||`, so a glance confirms whether a slope carries a
correlation. The fixed formula has had every random, latent, and special
term stripped out, which is exactly the formula used to build the
fixed-effects design matrix.

## Building from a data frame

A tidy data frame is the natural input. One row per observation, columns
typed as you intend them: numeric for continuous predictors, `factor` for
unordered categories, `ordered` for ranked ones. tulpa reads the type off
each column and builds the matching design piece, so a factor expands into
contrasts while a numeric column stays a single slope. Simulate a small
long-format frame with a numeric predictor, a numeric covariate, and two
grouping factors.

```{r make-data}
n  <- 200
df <- data.frame(
  y    = rpois(n, 3),
  x    = rnorm(n),
  z    = rnorm(n),
  g    = factor(sample(letters[1:6], n, replace = TRUE)),
  site = factor(sample(paste0("s", 1:4), n, replace = TRUE))
)
str(df)
```

Pass the parsed formula and the data to `tulpa_build_model_data()`. The
result is a list with the response vector, the fixed-effects design
matrix, an offset slot, the parsed random-effect terms, and a few count
fields the backends read.

```{r build}
bundle <- tulpa_build_model_data(pf, df)
names(bundle)
```

The design matrix `X` is the ordinary `model.matrix` output for the fixed
formula. It carries the intercept column unless you drop it, expands
factors into contrasts, and labels its columns.

```{r design}
dim(bundle$X)
colnames(bundle$X)
head(bundle$X, 3)
```

The response `y` comes from evaluating the left side of `~` against the
data, so a bare name, a transformation like `log(y)`, or a `cbind(...)`
pair all resolve through the same path. The count fields `n_obs`,
`n_fixed`, and `n_re_terms` are the sizes the backends need up front, and
`fixed_names` echoes the design columns.

```{r counts}
c(n_obs = bundle$n_obs, n_fixed = bundle$n_fixed,
  n_re_terms = bundle$n_re_terms)
```

When `tulpa()` runs a full fit it calls these two functions for you, then
attaches the design back onto the returned object as `$model_matrix` and
records `$fixed_names` and the parameter layout. Calling the builders
yourself is useful for checking that a factor has the levels you expect or
that an interaction expanded the way you wanted, before any inference runs.
The cost is a single function call, and the payoff is that a design problem
shows up as a printed matrix instead of a confusing fit failure several
steps later.

## Random-effect terms

Each random-effect term becomes one entry in `bundle$re_terms`, and the
entry carries everything a backend needs to place group-level effects: an
integer index mapping every row to its group, the number of groups, the
level labels, the coefficient count, and a slope matrix when the term has
slopes. Read the first term, the random intercept `(1 | g)`.

```{r re-intercept}
re1 <- bundle$re_terms[[1]]
re1[c("group_var", "n_groups", "n_coefs", "has_intercept")]
re1$levels
```

`group_idx` is the heart of it. Each observation gets an integer pointing
at one of the `n_groups` levels, and the order of `levels` fixes which
integer means which group. A random intercept has `n_coefs = 1` and no
slope matrix, since the design column is an implicit vector of ones.

```{r re-idx}
head(re1$group_idx, 12)
table(re1$group_idx)
```

The second term, `(1 + x || site)`, adds a random slope on `x`. Its
coefficient count rises to two (an intercept and the slope), the slope
column shows up in `slope_matrix`, and `correlated` is `FALSE` because the
double bar asked for independent effects.

```{r re-slope}
re2 <- bundle$re_terms[[2]]
re2[c("group_var", "n_groups", "n_coefs", "correlated")]
head(re2$slope_matrix)
```

Group levels come from factor levels, so the way you build a factor
controls the group ordering and which groups exist at all. If a level is
present in the factor but never observed, it still counts toward
`n_groups`, which inflates the number of group-level effects the model
carries with no data to inform them. To keep only the observed levels,
drop the unused ones with `droplevels()` before fitting. A character
column is coerced to a factor on the fly. That works, and it leaves the
level order up to `as.factor()`, which sorts alphabetically; setting the
factor yourself keeps that order under your control and makes the
`levels` slot read the way you expect.

The link between `group_idx` and `levels` is worth holding onto, because
it is how a fit reports group-level effects back to you. `ranef()` returns
one row per level in the order of `levels`, and a row of the design points
into that same vector through `group_idx`. When two factors share level
labels, keeping them as distinct columns rather than merging them avoids a
silent collision in the index.

The grouping side of a bar also accepts an expression rather than a bare
name, which covers crossed and nested designs without a reshape. Writing
`(1 | g:site)` crosses the two factors into one grouping, and tulpa
evaluates the interaction at build time instead of guessing column names.
Nested grouping `(1 | g/site)` expands into two terms, one for `g` and one
for the `g:site` combination, matching the lme4 convention. The parsed
object shows both terms, so you can confirm the expansion before fitting
rather than after.

```{r re-nested}
pn <- tulpa_parse_formula(y ~ x + (1 | g/site))
pn$n_re_terms
vapply(pn$random_effects, `[[`, character(1), "group_var")
```

## Binomial denominators and offsets

A binomial model needs to know the number of trials behind each success
count. tulpa offers two ways to supply that, and they suit different data
shapes. When your data frame has a column of successes and a column of
failures, pass both through `cbind()` on the response side. The builder
evaluates the call and returns a two-column matrix.

```{r binom-cbind}
db <- data.frame(x = rnorm(n))
prob   <- plogis(-0.2 + 0.8 * db$x)
db$succ <- rbinom(n, 15, prob)
db$fail <- 15 - db$succ

pf_b   <- tulpa_parse_formula(cbind(succ, fail) ~ x)
bun_b  <- tulpa_build_model_data(pf_b, db)
head(bun_b$y, 3)
```

When the denominators live in their own vector, keep the response as the
success count and hand the trials to `tulpa()` through `n_trials`, a vector
with one entry per row. This is the shape to reach for when every unit was
observed a known number of times and you would rather not carry a redundant
failure column. The two routes are equivalent: a single-column response of
successes paired with `n_trials`, or a two-column `cbind(succ, fail)`
response, describe the same model. Binary data is the special case where
every denominator is one, so a plain `0/1` response needs neither route.

```{r binom-ntrials}
db$tot <- 15L
fit_b <- tulpa(succ ~ x, data = db, family = "binomial",
               n_trials = db$tot, mode = "laplace")
coef(fit_b)

fit_cbind <- tulpa(cbind(succ, fail) ~ x, data = db, family = "binomial",
                   mode = "laplace")
max(abs(coef(fit_cbind) - coef(fit_b)))
```

An offset is a known additive term on the linear-predictor scale, fixed at
coefficient one rather than estimated, and exposure in a rate model is the
common case: a Poisson count observed over a varying window of time or
area enters as `offset(log(exposure))`, which turns the modelled mean into
a rate times exposure. The builder pulls the offset out of the design and
returns it in its own slot, so it never appears as a fitted column.

```{r offset}
dp <- data.frame(x = rnorm(n), expo = runif(n, 1, 8))
dp$y <- rpois(n, dp$expo * exp(0.1 + 0.5 * dp$x))

pf_o  <- tulpa_parse_formula(y ~ x + offset(log(expo)))
bun_o <- tulpa_build_model_data(pf_o, dp)
head(bun_o$offset, 4)
colnames(bun_o$X)
```

The offset column is `log(expo)` evaluated row by row, and `X` holds only
the intercept and `x`. The transformation inside `offset(...)` runs as
written, so the log is yours to apply, not something tulpa adds. Fit it as
a Poisson model and the slope recovers near its true value of 0.5, with the
exposure absorbed through the offset rather than estimated as a free
coefficient.

```{r offset-fit}
fit_o <- tulpa(y ~ x + offset(log(expo)), data = dp,
               family = "poisson", mode = "laplace")
coef(fit_o)
```

## Validation and errors

The builders fail early and name the problem, rather than passing a
half-formed design to a backend. The most common slip is a name in the
formula that has no matching column. Suppose the grouping factor is called
`g` in the data but the formula asks for `region`: the build stops and
quotes the offending name.

```{r err-group, error = TRUE}
bad <- tulpa_parse_formula(y ~ x + (1 | region))
tulpa_build_model_data(bad, df)
```

A response that cannot be found is reported the same way, against the
label that appeared on the left of `~`.

```{r err-response, error = TRUE}
tulpa_build_model_data(tulpa_parse_formula(count ~ x), df)
```

The special terms check their own shape at parse time. A `spatial(...)`
term wraps exactly one bare column name, so a call with the wrong arity is
caught before any data is touched, with a message that shows the intended
form.

```{r err-spatial, error = TRUE}
tulpa_parse_formula(y ~ spatial(x, g))
```

A spatial term also has a partner requirement that surfaces at fit time.
Naming a unit column with `spatial(col)` tells tulpa where each row sits,
but the field structure (the adjacency for an areal model, coordinates for
a continuous one) arrives through the `spatial=` argument. Leaving it out
stops the fit with guidance on what to pass.

```{r err-spatial-spec, error = TRUE}
tulpa(y ~ x + spatial(g), data = df, family = "poisson", mode = "laplace")
```

These checks share a habit worth leaning on. The formula declares intent,
the data supplies the columns, and the builder reconciles the two before
anything numeric happens. When a fit fails to start, parsing the formula
on its own and then building the model data by hand isolates whether the
trouble is in the formula grammar or in the columns the data does or does
not carry. Reading `bundle$re_terms` confirms the group counts; reading
`colnames(bundle$X)` confirms how factors and interactions expanded. Most
data-shape questions answer themselves once those two pieces are on screen.

## See also

- `?tulpa_parse_formula`, `?tulpa_build_model_data` for the function
  references.
- The getting-started vignette for the fit, extract, predict, and compare
  workflow once the data is in shape.
- The spatial and priors vignettes for the structures that ride alongside
  the formula through `spatial=` and the prior arguments.
