---
title: "EM + Laplace for latent-variable models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EM + Laplace for latent-variable models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r load, message = FALSE}
library(tulpa)
```

## What the driver is for

Some models have a latent variable that, once known, turns the fit into
an ordinary weighted GLM: occupancy (was the site occupied?), N-mixture
(what was the true count?), zero-inflation (is this a structural zero?).
`tulpa_em_laplace()` is the generic engine for that pattern. It does not
know your model -- you supply two callbacks and it runs the EM loop,
solving each M-step submodel with `tulpa_laplace()` and checking
convergence.

This is deliberately an engine door: model packages such as **tulpaObs**
build their occupancy and N-mixture fitters on top of it. The callbacks
below are the whole contract.

## The two callbacks

`e_step(fits, ...)` receives the current per-submodel fits and returns
the latent-variable posterior as weights:

```{r estep, eval = FALSE}
e_step <- function(fits, ...) {
  # Use the current fits to compute the responsibility of each observation
  # (e.g. the posterior probability that a zero is a Poisson zero, not a
  # structural one). Return them as a `weights` element.
  list(weights = responsibilities)
}
```

`m_step_encode(weights, ...)` turns those weights into one or more
weighted GLM submodels. Each block is a `list(y, n_trials, X, family,
offset)`; the driver fits each with `tulpa_laplace()` and threads the
per-block `family` and `offset` through automatically:

```{r mstep, eval = FALSE}
m_step_encode <- function(weights, ...) {
  list(
    lambda = list(y = counts, n_trials = 1L, X = X_abund,
                  family = "poisson", offset = log(weights)),
    pi     = list(y = z,      n_trials = 1L, X = X_zero,
                  family = "binomial", offset = NULL)
  )
}
```

## Running it, and the bias correction

```{r fit, eval = FALSE}
fit <- tulpa_em_laplace(
  e_step        = e_step,
  m_step_encode = m_step_encode,
  max_iter      = 30L,
  tol           = 1e-4
)
fit$fits        # the converged per-submodel Laplace fits
fit$n_iter
fit$converged
```

The EM point estimate is fast but its standard errors ignore the
uncertainty in the imputed latent variable. Two post-EM corrections
restore it, pooled by Rubin's rules:

```{r correct, eval = FALSE}
# Multiple imputation: refit on several latent draws and pool.
fit_mi <- tulpa_em_laplace(e_step, m_step_encode, correction = "mi")

# Warm-started Gibbs from the EM mode, then pool.
fit_gibbs <- tulpa_em_laplace(e_step, m_step_encode, correction = "gibbs")
```

An optional `beta_prior = list(mean, sd)` threads a Gaussian fixed-effect
prior into every M-step block and into the correction refits, and
`m_step_extra(fits, weights, ...)` updates non-linear-predictor
parameters (dispersions, mixing weights) between the M- and E-steps. See
`?tulpa_em_laplace`. For a Monte-Carlo E-step, use `tulpa_em_mc()`.
