---
title: "Inference modes: the three tiers and how to choose one"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inference modes: the three tiers and how to choose one}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260529)
```

```{r load, message = FALSE}
library(tulpa)
```

## Why the inference choice matters

A fitted model is two things at once: a set of parameter estimates and a
claim about how certain those estimates are. The estimates often agree
across methods. The uncertainty does not. A Laplace approximation, a
gradient sampler, and a variational fit can return the same posterior
mean for a slope and still disagree about the width of its credible
interval, the shape of its tail, and whether two parameters are
correlated. The number you report as a 95% interval depends on which
method produced it, and the methods do not all mean the same thing by
"95%".

Most fitting interfaces hide this. You call a fit function, a default
runs, and the output looks identical no matter which numerical engine did
the work. The default is usually a reasonable one. The problem is that it
is silent: when the default is wrong for your model, nothing in the output
tells you, and the failure shows up downstream as an interval that is too
narrow or a correlation that was never there.

tulpa makes the choice explicit. Every fit carries the mode it ran, the
tier that mode belongs to, and a recorded reason for the selection. The
tier carries the weight, because it pins down what the credible intervals
mean. Two backends sharing a tier offer the same epistemic warranty; a
backend one rung higher offers a weaker one. That warranty reads straight
off the fitted object, and `mode = "auto"` picks among tiers by a rule you
can inspect rather than by a buried heuristic.

This vignette walks through the tier system, the automatic selection
rule, and the backends you can drive from R. It fits one model three ways
and compares the estimates and the timing, so the cost-versus-guarantee
trade is concrete rather than abstract. It closes with practical guidance:
which mode to reach for, and at what point to move up a tier.

## The three tiers

tulpa sorts every inference backend into one of three tiers. The tier is
not a speed label. The tier states how correct the uncertainty a backend
reports actually is. `inference_mode_info()` prints the full
map, tier by tier, with the guarantee attached to each.

```{r mode-info}
inference_mode_info()
```

The output names three tiers. Read them as a ladder of promises.

**Tier 1, Exact.** A Tier 1 backend draws from the posterior and its
credible intervals are interpretable as posterior uncertainty, up to
Monte Carlo error that shrinks as you draw more samples. This is the
reference standard: if you want an interval you can quote without an
asterisk, a Tier 1 fit gives it to you. The cost is iteration. Every draw
is a likelihood evaluation, sometimes a gradient evaluation, and you need
many of them. MALA and the independence-MH samplers live here, along with
the Polya-Gamma Gibbs sampler for the families that admit it.

**Tier 2, Structured.** A Tier 2 backend is accurate conditional on an
explicit structural assumption. The Laplace approximation assumes the
posterior is Gaussian near its mode. Pathfinder fits a Gaussian along an
optimisation path. When the assumption holds, and for a latent Gaussian
model with enough data it usually does, the fit is fast and its intervals
are close to the exact ones. When the assumption fails, the failure is
predictable: a skewed posterior gets symmetrised, a heavy tail gets
clipped. You can see it coming from the model, and you can check it by
comparing against a Tier 1 fit. The promise is conditional, and the
condition is stated.

**Tier 3, Optimized.** A Tier 3 backend gives a point estimate and a
covariance from an optimisation objective with no general correctness
guarantee on the uncertainty. The mode is usually good. The spread is
often too small, the tails unreliable, the correlations approximate, and
the failure is typically silent. Generic variational inference sits here.
Tier 3 is optimisation, not sampling, and tulpa treats it as a deliberate
choice rather than a fallback, reachable only by asking for it by name.

The ladder runs from a promise you can quote, through a promise with a
stated condition, to a tool with no promise on its uncertainty. Moving up
a tier costs computation and buys a stronger guarantee. The point of the
system is that the guarantee is never traded away without a record of the
trade.

The design follows four rules, and they are worth stating because they
explain why the tier shows up on every fit. First, a mode changes the
meaning of the output, not just its runtime: an interval from the optimised
tier does not mean what an interval from the exact tier means, so the two
must be distinguishable. Second, the mode is always visible in the result.
Third, there is no silent upgrading or downgrading between tiers; if an
exact fit fails, the call errors rather than quietly substituting a
structured approximation and handing back numbers that look the same but
promise less. Fourth, adding a backend slots it into an existing tier and
never invents a new kind of promise, so the three guarantees above are the
whole vocabulary. Tier membership for each backend is derived from a single
registry, which is why `inference_mode_info()` and the field on a fit can
never drift apart.

## How `mode = "auto"` chooses

The default mode is `"auto"`. Its contract is one sentence: use the most
reliable method that is expected to finish for this model. Two parts of
that sentence carry weight. "Most reliable" means it prefers Tier 1 over
Tier 2 when both are feasible. "Expected to finish" means it steps down to
Tier 2 when a Tier 1 sampler would be too slow to be practical. The rule
never reaches Tier 3. The optimisation tier is never selected on your
behalf, because its silent under-coverage is exactly the failure an
automatic default should not introduce.

The decision is deterministic and depends on the model, not on chance. Two
properties drive it: structure and size. A latent prior block goes to
nested Laplace, the designed Tier 2 hot path for latent Gaussian structure.
A spatial field branches by its kind, with a binomial areal field handed to
the exact component-wise Polya-Gamma Gibbs sampler (Tier 1) and most other
fields sent to nested-Laplace integration of the spatial hyperparameter
(Tier 2). An ordinary model carrying neither spatial nor latent structure
heads for a full sampler, unless the dataset grows past tens of thousands
of rows, where the Laplace path takes over because a sampler would no
longer finish quickly enough to serve as a default.

Every fit records the reason. The `selection_reason` field is a short
string explaining the branch that was taken, and `backend` and
`inference_tier` report the result. Take a binomial areal model: twenty
regions on a ring, each adjacent to its two neighbours, with a latent
field that varies smoothly around the loop.

```{r auto-build}
n <- 400L; K <- 20L
W <- matrix(0, K, K)
for (i in 1:K) { j <- if (i < K) i + 1L else 1L; W[i, j] <- W[j, i] <- 1 }
x      <- rnorm(n)
region <- factor(sample(1:K, n, replace = TRUE))
field  <- as.numeric(scale(sin(2 * pi * (1:K) / K)))[region]
ds <- data.frame(y = rbinom(n, 1, plogis(-0.2 + 0.7 * x + field)),
                 x = x, region = region)
```

Fitting with `mode = "auto"` lets the rule pick. For a binomial ICAR field
it selects the exact Polya-Gamma Gibbs sampler, a Tier 1 backend, and says
so.

```{r auto-spatial}
fit_auto <- tulpa(y ~ x + spatial(region), data = ds, family = "binomial",
                  spatial = list(type = "icar", adjacency = W), mode = "auto")
c(backend = fit_auto$backend, tier = fit_auto$inference_tier)
fit_auto$selection_reason
```

Auto reached for the most reliable method, Tier 1, because for this model
class an exact sampler is both available and fast. A binomial likelihood
with an areal field admits the Polya-Gamma data augmentation, which turns
each conditional update into a Gaussian draw, and the field components
update one at a time. Those component-wise updates sidestep the
dimensionality that would slow a general gradient sampler on a field of
this size, so the exact route is also the practical one. The rule saw a
case where the strongest guarantee was affordable and took it. Now change
the situation to a large, plain Gaussian model where a full sampler would
be slow.

```{r auto-large}
nL <- 60000L
xL <- rnorm(nL)
dL <- data.frame(y = 0.5 + 1.2 * xL + rnorm(nL, sd = 0.8), x = xL)
fit_big <- tulpa(y ~ x, data = dL, family = "gaussian", mode = "auto",
                 phi = 0.8)
c(backend = fit_big$backend, tier = fit_big$inference_tier)
fit_big$selection_reason
```

Here auto stepped down to the Laplace path. The reason names the dataset
size: at sixty thousand rows the rule judges a full sampler too slow to be
the sensible default and takes the Tier 2 route instead, which for a
Gaussian likelihood is exact anyway. The estimates land on the truth.

```{r auto-large-coef}
coef(fit_big)
```

Two models, two different branches, each recorded. The same rule that
picked an exact sampler for the small spatial model picked the structured
approximation for the large plain one, and in both cases the choice traces
to a property of the model you can name in advance. That predictability is
the whole point. You are never left guessing which engine ran, and you can
reason about what auto will do before you call it. When you disagree with
the choice, override it by passing a tier (`"exact"` or `"structured"`) or
a backend name directly, and the fit records that you overrode it rather
than papering over the change.

## The R-callable backends

A backend is the concrete algorithm a mode resolves to. tulpa registers
more backends than it exposes from R: several ship a C++ kernel reachable
by model packages that link against tulpa, with no R entry point yet. The
`inference_mode_info()` map tags each one, `[R]` for callable from R and
`[C-ABI]` for the kernel-only ones. Asking for a C-ABI backend by name from
R errors with a message that names the missing entry point rather than
pretending to dispatch. This section covers the backends you can drive
from R today.

### laplace (Tier 2)

The workhorse. `mode = "laplace"` finds the posterior mode and the
curvature there, then reports a Gaussian centred at the mode with that
curvature as its precision. For a Gaussian likelihood this is exact and
returns immediately. For other families it is the Gaussian approximation
to the posterior, accurate when the posterior is close to Gaussian.

```{r laplace-fit}
set.seed(101)
g  <- factor(sample(1:12, n, replace = TRUE))
u  <- rnorm(12, sd = 0.6)
db <- data.frame(y = rbinom(n, 1, plogis(-0.3 + 1.0 * x + u[g])),
                 x = x, g = g)
fit_lap <- tulpa(y ~ x + (1 | g), data = db, family = "binomial",
                 mode = "laplace", sigma_re = 0.6)
coef(fit_lap)
```

The Laplace fit is deterministic, carries no Monte Carlo error, and its
`logLik()` is the approximate log marginal likelihood, the model evidence
that `compare_models()` uses. This is the first fit to run, every time, as
the fast sanity check and the comparison baseline.

### pathfinder (Tier 2)

Pathfinder runs a quasi-Newton optimiser toward the mode, fits a Gaussian
at the optimum from the inverse-Hessian estimate it accumulates along the
way, and returns draws from that Gaussian plus an ELBO score. It sits in
the same tier as Laplace: the output is a Gaussian approximation, not exact
samples. What it adds over a plain Laplace fit is a set of draws and a
diagnostic. The draws make it a drop-in where downstream code expects
samples, and the ELBO gauges how well the Gaussian fits.

```{r pathfinder-fit}
fit_pf <- tulpa(y ~ x + (1 | g), data = db, family = "binomial",
                mode = "pathfinder", sigma_re = 0.6,
                control = list(n_draws = 1000))
coef(fit_pf)
fit_pf$elbo
```

Pathfinder fits three situations well: you want draws but not the cost of
full MCMC, you need a warm start for a sampler, or you want a quick check on
whether the Laplace Gaussian is reasonable.

### mala (Tier 1)

The Metropolis-adjusted Langevin algorithm is a gradient sampler. Each
step proposes a move along the gradient of the log posterior and accepts or
rejects it with a Metropolis step, which makes the chain asymptotically
correct. Each iteration runs cheaper than full Hamiltonian Monte Carlo
because no leapfrog trajectory needs integrating, and mixing beats a random
walk because the drift term nudges proposals toward higher density. The
step size adapts during warmup toward an acceptance rate near 0.574.

```{r mala-fit}
fit_mala <- tulpa(y ~ x + (1 | g), data = db, family = "binomial",
                  mode = "mala", sigma_re = 0.6,
                  control = list(n_iter = 1500, warmup = 500))
coef(fit_mala)
fit_mala$mean_accept
```

MALA earns its keep when you want exact posterior moments at moderate
dimension and the posterior is not so badly scaled that a single step size
struggles across all directions. The acceptance rate is the diagnostic: far
below the target means the chain needs a smaller step or more warmup.

### imh_laplace (Tier 1)

Independence Metropolis-Hastings with a Laplace proposal is the cheapest
route to exact-tier draws when Laplace is almost right. It builds the
Laplace mode and precision once, then proposes from that Gaussian and
corrects with a Metropolis step. The proposal ignores
the current state, so the iterations are nearly free and the acceptance rate
doubles as a verdict on the Laplace approximation: high acceptance means
the posterior is close to the Gaussian, low acceptance means it is far and
Laplace was biased.

```{r imh-fit}
fit_imh <- tulpa(y ~ x + (1 | g), data = db, family = "binomial",
                 mode = "imh_laplace", sigma_re = 0.6,
                 control = list(n_iter = 1500, warmup = 500))
coef(fit_imh)
fit_imh$mean_accept
```

The acceptance here is high, which says the Laplace fit was already close.
Two jobs suit `imh_laplace`: debiasing a Laplace fit at low parameter
dimension, and repeated-fit workflows such as cross-validation where a full
sampler's startup cost would dominate.

### gibbs (Tier 1)

The Gibbs backend is the Polya-Gamma sampler for binomial and negative
binomial responses, including the binomial areal spatial models that
`mode = "auto"` selects it for. It samples the random-effect standard
deviation rather than conditioning on a fixed `sigma_re`, which is why the
spatial fit earlier reported a Gibbs backend. Its output carries the fixed
effects in `$beta` and the field and variance components alongside.

```{r gibbs-fields}
round(fit_auto$beta, 3)
```

Gibbs is the right call when the family is conjugate-friendly under the
Polya-Gamma scheme and you want an exact fit that samples the variance
rather than fixing it. For a binomial areal field, auto takes this path on
its own.

### re_cov_nested and agq (Tier 2)

Two further Tier 2 backends serve narrower roles. `re_cov_nested`
integrates a random-effect covariance matrix rather than conditioning on a
fixed standard deviation. A random-slope term such as `(1 + x | g)` has no
scalar `sigma_re` to condition on, so when a slope term is present the
Laplace path redirects to this backend automatically and integrates the
covariance with a CCD design and a PC or LKJ prior. You reach it not by
name but by writing a slope term and fitting at `mode = "laplace"`. The
adaptive Gauss-Hermite quadrature backend, `agq`, integrates a
random-intercept variance by quadrature and is callable through its fitter
`agq_fit()` for the single-grouping case rather than through `tulpa()`.

The seven R-callable backends span both exact tiers and the structured
tier. The next section puts three of them on the same model so the trade
between cost and guarantee is visible in numbers.

## Cost versus guarantee

The clearest way to see what a tier buys is to fit one model several ways
and lay the results side by side. Take the binomial random-intercept model
from above and fit it at Laplace, MALA, and Pathfinder, timing each with
`system.time()`.

```{r timing}
t_lap  <- system.time(
  f_lap  <- tulpa(y ~ x + (1 | g), data = db, family = "binomial",
                  mode = "laplace", sigma_re = 0.6))[["elapsed"]]
t_mala <- system.time(
  f_mala <- tulpa(y ~ x + (1 | g), data = db, family = "binomial",
                  mode = "mala", sigma_re = 0.6,
                  control = list(n_iter = 1500, warmup = 500)))[["elapsed"]]
t_pf   <- system.time(
  f_pf   <- tulpa(y ~ x + (1 | g), data = db, family = "binomial",
                  mode = "pathfinder", sigma_re = 0.6,
                  control = list(n_draws = 1000)))[["elapsed"]]
```

Collect the slope estimate, its standard error, the tier, and the elapsed
time into one table.

```{r timing-table}
slope_se <- function(f) summary(f)["x", "std.error"]
data.frame(
  backend  = c(f_lap$backend, f_mala$backend, f_pf$backend),
  tier     = c(f_lap$inference_tier, f_mala$inference_tier,
               f_pf$inference_tier),
  slope    = round(c(coef(f_lap)["x"], coef(f_mala)["x"],
                     coef(f_pf)["x"]), 3),
  slope_se = round(c(slope_se(f_lap), slope_se(f_mala),
                     slope_se(f_pf)), 3),
  seconds  = round(c(t_lap, t_mala, t_pf), 3)
)
```

Three patterns sit in this table. The slope estimates agree across all
three backends to within their standard errors. That is expected: the
posterior mean is the easy quantity, and every method finds it. The
standard errors agree too, because for this model the posterior is
near-Gaussian and the Tier 2 approximation is accurate. That agreement is
the signal that Laplace was safe here, confirmed independently by the high
IMH acceptance earlier. Timing is where the methods part ways. The
deterministic Laplace fit returns fastest. Pathfinder costs an optimisation
plus a draw step. MALA pays for its full chain of gradient evaluations, and
that cost is the visible price of the exact guarantee.

Read the table as a guarantee you bought, not effort you wasted. The
sampler purchased a promise this model did not need and another model
would. The agreement you see is itself the product of a near-Gaussian
posterior. Push the model toward a skewed or heavy-tailed posterior, by
shrinking the data to a few dozen rows, moving to a sparse binomial with
most outcomes zero, or adding a poorly identified variance component, and
the Tier 2 standard error would start to drift away from the Tier 1 one.
The samplers would track the true posterior spread through that drift,
while the Gaussian approximation would keep reporting the symmetric
interval its assumption forces on it. That is the moment the price of the
sampler turns into a reason to pay it.

The marginal likelihood ties the comparison back to model choice.
`compare_models()` reads `logLik()` from each fit, which on the Laplace
tier is the approximate log marginal likelihood.

```{r compare}
compare_models(laplace = f_lap, pathfinder = f_pf, criterion = "loglik")
```

The `glance()` accessor surfaces the per-fit diagnostics that matter for
the sampler tiers: the number of post-warmup draws, the acceptance rate,
and the divergence count.

```{r glance}
glance(f_mala)[c("n_samples", "mean_accept", "n_divergent")]
```

An acceptance rate inside the healthy band and no divergences say the MALA
chain mixed well, which is the prerequisite for trusting its intervals.
Without that check, a Tier 1 fit is only nominally exact: a stuck chain
reports a posterior it never explored.

## Choosing a mode

The tier system gives a default worth following and a small set of rules
for when to depart from it.

- **Start at `mode = "laplace"`.** The fit is deterministic, carries no
  Monte Carlo error, and returns in well under a second on the models in
  this vignette and in a fraction of a second per thousand rows on much
  larger ones. Use it for the first fit and for model comparison by marginal
  likelihood, and keep it as the baseline even when you intend to finish
  with a sampler.

- **Check Laplace before you trust its intervals.** The cheapest check is
  `mode = "imh_laplace"` on the same model: a high acceptance rate, above
  roughly 0.5, says the posterior is close to the Laplace Gaussian and the
  intervals are safe. An acceptance rate below about 0.1 says the posterior
  is far from Gaussian and the Laplace intervals are biased. The fit
  earlier accepted around 0.86, which is a clean pass.

- **Move to `mode = "mala"` for exact moments at moderate dimension.** When
  the posterior is visibly non-Gaussian, when imh_laplace acceptance is
  low, or when you simply need intervals you can quote without the
  Gaussian caveat, the gradient sampler is the next step. Budget a few
  thousand iterations with warmup at a third to a half of them, and watch
  the acceptance rate settle near 0.574 and the divergence count stay at
  zero.

- **Use `mode = "pathfinder"` when you want draws cheaply.** It costs more
  than Laplace and far less than MALA, returns a sample you can feed to
  downstream code, flags a poor Gaussian fit through its ELBO, and doubles
  as a warm start for a sampler.

- **Let `mode = "auto"` choose for structured models.** For a latent block
  or a spatial field, auto routes to the designed path: nested Laplace for
  latent and most spatial fields, exact Polya-Gamma Gibbs for binomial
  areal models, and the Laplace path for very large plain models past tens
  of thousands of rows. Read `selection_reason` to confirm the branch. On a
  small plain model with no spatial or latent structure auto defaults to the
  MALA gradient sampler (Tier 1); name `mode = "laplace"` instead when you
  want the fast deterministic fit.

- **Treat `mode = "optimized"` as a deliberate exception.** Tier 3 is opt-in
  because its uncertainty is unreliable and its failures are silent. It fits
  a narrow case: a fast point estimate when you have an independent handle on
  the uncertainty. Keep it off the default path, and off any interval you
  intend to report.

The thread through all of it is that the tier is visible and the choice is
recorded, so you are never reasoning about uncertainty you cannot account
for. Read `$backend` and `$inference_tier` off any fit to see what ran and
what its intervals promise, then read `$selection_reason` to see why that
backend was chosen; when the promise on a fit is not the one your analysis
needs, move up a tier and pay for the stronger guarantee on purpose.

## See also

- `inference_mode_info()` prints the full tier and backend map, with the
  R-callable and C-ABI-only backends tagged.
- The quickstart vignette covers the fitting workflow.
- The `tgmrf()` vignette runs the same tier ladder on a user-defined latent
  block, from nested Laplace through exact sampling.
- `?mala`, `?pathfinder`, `?imh_laplace` for the low-level sampler
  interfaces that drive a `log_posterior` closure directly.
