---
title: "Comparing tulpa models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing tulpa models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260529)
```

```{r load, message = FALSE}
library(tulpa)
```

## Why compare

You rarely fit one model. You fit a few: a plain trend, the trend
with a covariate, the same again with a grouping structure. Each
candidate encodes a different belief about what generated the data, and
the question that follows is always the same. Which structure does the
data support, and by how much?

tulpa answers that at the engine level with one number per fit: the log
marginal likelihood from a Laplace approximation. It is the model
evidence, the quantity that automatically rewards a better fit and
penalises wasted flexibility. This vignette walks through the criterion,
builds a nested ladder of models on data with a known generating
process, and reads the evidence as it climbs. It also says plainly where
the marginal likelihood stops and where WAIC and LOO take over, because
those two need machinery a base `tulpa()` fit does not carry.

The plan is concrete. Simulate Gaussian data with an intercept, a slope,
a group-level offset, and a second covariate that all genuinely shape
the response. Fit the ladder. Watch the evidence rise at each rung.
Replay the exercise on a binary outcome so the verdict holds across two
families instead of resting on a single example.

## Marginal likelihood from a Laplace fit

A Bayesian model assigns a probability to the data after the parameters
are integrated out:

$$
p(y \mid M) \;=\; \int p(y \mid \theta, M)\, p(\theta \mid M)\, d\theta .
$$

This is the marginal likelihood, also called the evidence. It sits in
the denominator of Bayes' theorem, and it grades a whole specification
rather than a single parameter value. A model whose prior mass piles up
where the observations actually fall scores high. A model that scatters
its prior across regions the data exclude pays a toll, since the integral
averages the likelihood against the prior and that wasted breadth drags
the average down. An extra coefficient earns its keep only when the
likelihood it unlocks outweighs the dilution it imposes. This trade is
Occam's razor, quantified: parsimony falls out of the arithmetic, and
you never tack on a manual complexity penalty.

The integral resists a closed form for most models. The Laplace
approximation tackles it through a tidy device: locate the posterior
mode, measure the curvature (the Hessian) there, and approximate the
integrand by a Gaussian bell centred on that peak. Under a Gaussian
likelihood paired with a Gaussian prior the bell is exact, so what comes
out is the genuine marginal likelihood. For heavier-tailed or skewed
families it becomes a second-order estimate, reliable whenever the
posterior stays roughly bell-shaped, which holds at most realistic
sample sizes.

When you fit with `mode = "laplace"`, tulpa computes this quantity and
stores it on the fit. `logLik()` returns it.

```{r first-evidence}
n  <- 500
x  <- rnorm(n)
y0 <- 0.5 + 1.2 * x + rnorm(n, sd = 0.8)
df <- data.frame(y = y0, x = x)

fit <- tulpa(y ~ x, data = df, family = "gaussian",
             mode = "laplace", phi = 0.8)
logLik(fit)
```

The `logLik` object carries two attributes worth knowing. `df` is the
number of fixed effects the fit estimated, and `nobs` is the number of
observations the evidence was computed on. Both matter for comparison:
an evidence number only means something against another fit on the same
data.

```{r evidence-attrs}
c(df = attr(logLik(fit), "df"), nobs = attr(logLik(fit), "nobs"))
```

A subtle point about the tiers. On the Laplace tier `logLik()` is the
log *marginal* likelihood, the evidence with parameters integrated out.
On a sampler tier (MALA, Gibbs, Pathfinder) the same accessor reports
the mean log-density across the draws, which is a different object and
not directly comparable to a marginal likelihood. For model comparison
by evidence, fit every candidate on the Laplace tier so the numbers
share a definition. That is also the fast path, which makes fitting a
whole ladder cheap.

## A nested model ladder

The clearest way to read evidence is to nest your candidates, each one
sitting inside the next, adding a single structural ingredient per step.
When an ingredient mirrors a real feature of the generating mechanism,
the evidence climbs once you include it. When it merely decorates, the
evidence stalls or slips. A ladder built on simulated truth converts the
criterion into something you verify rather than merely believe.

Simulate the truth first. The response has an intercept of 0.5, a slope
of 1.2 on `x`, a group-level offset with standard deviation 0.7 across
fifteen groups, and a second slope of 0.8 on `z`. Every term is real, so
a good criterion should reward each one as it enters.

```{r sim-ladder}
n  <- 500
x  <- rnorm(n)
z  <- rnorm(n)
g  <- factor(sample(1:15, n, replace = TRUE))
u  <- rnorm(15, sd = 0.7)
y  <- 0.5 + 1.2 * x + 0.8 * z + u[g] + rnorm(n, sd = 0.8)
dat <- data.frame(y = y, x = x, z = z, g = g)
```

Now the four rungs. The intercept-only model is the null reference. The
second rung adds the `x` slope. The third adds a random intercept on
`g`, conditioning on the random-effect standard deviation. The fourth
adds the second covariate `z`. Each fit is a single Laplace solve.

```{r fit-ladder}
m0 <- tulpa(y ~ 1, data = dat, family = "gaussian",
            mode = "laplace", phi = 0.8)
m1 <- tulpa(y ~ x, data = dat, family = "gaussian",
            mode = "laplace", phi = 0.8)
m2 <- tulpa(y ~ x + (1 | g), data = dat, family = "gaussian",
            mode = "laplace", sigma_re = 0.7, phi = 0.8)
m3 <- tulpa(y ~ x + z + (1 | g), data = dat, family = "gaussian",
            mode = "laplace", sigma_re = 0.7, phi = 0.8)
```

Pull the evidence from each and lay it side by side. The differences
between adjacent rungs are what to read: each is the evidence gained by
adding one true piece.

```{r ladder-evidence}
ll <- c(intercept = as.numeric(logLik(m0)),
        slope     = as.numeric(logLik(m1)),
        slope_re  = as.numeric(logLik(m2)),
        full      = as.numeric(logLik(m3)))
data.frame(model = names(ll), logLik = round(ll, 1),
           gain = c(NA, round(diff(ll), 1)))
```

Every step up the ladder lifts the evidence, and by a wide margin. The
`x` slope delivers the largest single jump, befitting its sizeable true
coefficient, while the random intercept and the second covariate each
contribute their own distinct signal that the evidence faithfully
registers. The ordering reproduces the generating recipe precisely. That
is exactly why running the comparison against a known truth pays off.

The gains here are enormous because every added term is genuinely
present and the sample is large. Read the *sign and size* of each gain,
not its precise value. A large positive gain says the term earns its
place. A gain near zero, or negative, would say the term adds structure
the data do not ask for.

## compare_models()

Computing `logLik()` per fit and assembling a table by hand works, and
`compare_models()` does it for you. Pass the fits as named arguments and
choose the criterion.

```{r compare}
compare_models(intercept = m0, slope = m1,
               slope_re = m2, full = m3, criterion = "loglik")
```

The table reports the model name, the parameter count, and the log
marginal likelihood. The names come straight from the arguments, so
label them in a way that reads as a ladder.

Read the `n_params` column with care. It counts fixed effects. The
random-intercept model `m2` shows the same count as the slope-only model
`m1`, because the random intercept is a structural term conditioned on a
fixed `sigma_re` rather than an estimated fixed effect. The evidence,
which integrates over the group offsets, still rewards the structure;
the column simply does not count it as a fixed parameter. When you read
the table, lean on the `logLik` differences and treat `n_params` as a
fixed-effect bookkeeping figure.

How large a difference counts as real? Evidence lives on the log scale,
so a gap of `d` means the winning model is `exp(d)` times more probable
given the data, before any preference between the models themselves. A
handful of log units already settles the matter; the leaps in this
ladder run into the hundreds. Narrow margins, a unit or two, warrant
caution, particularly where the added term strains the approximation,
say a sharply skewed non-Gaussian response whose posterior departs from
the bell shape.

The same call shape works for a binary response. Simulate a logistic
process with a real intercept, a real `x` slope, and a real `z` slope,
then fit the same three-rung ladder.

```{r binom-ladder}
set.seed(101)
xb  <- rnorm(n); zb <- rnorm(n)
yb  <- rbinom(n, 1, plogis(-0.4 + 1.1 * xb + 0.9 * zb))
dfb <- data.frame(y = yb, x = xb, z = zb)

b0 <- tulpa(y ~ 1,     data = dfb, family = "binomial", mode = "laplace")
b1 <- tulpa(y ~ x,     data = dfb, family = "binomial", mode = "laplace")
b2 <- tulpa(y ~ x + z, data = dfb, family = "binomial", mode = "laplace")
compare_models(intercept = b0, x = b1, xz = b2, criterion = "loglik")
```

Both covariates were in the generating process, and the evidence climbs
as each enters. For the binomial family the marginal likelihood is a
Laplace approximation rather than an exact integral, so the numbers
carry a small approximation error. With a well-behaved posterior, as
here, that error is far smaller than the gaps between models, and the
ranking is safe to read.

## Interpreting logLik differences

A worked reading of one gap makes the scale concrete. Take the binomial
ladder: adding `z` to the `x`-only model raised the evidence.

```{r interpret-gap}
gain <- as.numeric(logLik(b2)) - as.numeric(logLik(b1))
c(log_evidence_gain = round(gain, 2),
  evidence_ratio    = round(exp(gain), 1))
```

The evidence ratio is the Bayes factor for the larger model against the
smaller, assuming equal prior weight on the two. A ratio well above one
favours the richer model. The log scale is the comfortable place to work
because the numbers stay manageable and differences add up across nested
steps. Three rules of thumb help when you read a column of evidence:

- A difference of a fraction of a unit is noise. Treat the two models as
  tied and prefer the simpler one.
- A difference of a few units is meaningful support for the better
  model. The Bayes factor is in the tens or more.
- A difference of tens or hundreds, as in the Gaussian ladder, is
  decisive. The structure is clearly in the data.

Always compare on the same data, the same family, and the same tier. The
evidence is a property of a model *and* the data it saw; numbers from
fits on different rows, different responses, or different inference tiers
are not on a common scale.

## glance() per model

`glance()` gives a one-row summary of any single fit, which is handy for
folding several fits into a tidy frame next to the comparison. On a
Laplace fit it reports the fixed-effect count, the log marginal
likelihood, and a convergence flag; the sampler-only fields stay `NA`
because the Laplace tier carries no draws.

```{r glance-one}
glance(m3)
```

Stack the ladder into one frame by binding the per-model rows. This pairs
naturally with `compare_models()`: the comparison ranks, while `glance()`
carries the per-fit detail such as convergence.

```{r glance-stack}
do.call(rbind, Map(function(m, nm) cbind(model = nm, glance(m)),
                   list(m0, m1, m2, m3),
                   c("intercept", "slope", "slope_re", "full")))
```

The `converged` column is the one to scan before trusting any evidence
number. A Laplace fit that did not converge has not found the mode, so
its curvature, and the marginal likelihood that depends on it, are not
to be read. On the sampler tiers `glance()` also reports `mean_accept`
and the divergence count, which play the same gatekeeping role for those
fits.

## What WAIC and LOO need

The marginal likelihood scores a model on the data it was fit to. WAIC
and LOO estimate something different and complementary: how well the
model predicts data it has not seen. Both are built from the *pointwise*
log-likelihood, one value per observation per posterior draw, arranged
in a draws-by-observations matrix. From that matrix the loo package
computes the expected log predictive density and its information
criteria.

A base `tulpa()` fit does not carry that matrix. The Laplace tier stores
a mode and a curvature, not a stack of draws, and even the sampler tiers
store parameter draws rather than a per-observation log-likelihood. So
the engine's `waic` and `loo` adapters have nothing to read on a plain
fit, and they say so:

```{r waic-needs, error = TRUE}
loo::waic(b2)
```

The message names the missing piece. It is a pointwise log-likelihood,
the kind that only a full observation model produces. Model packages
built on this engine, such as tulpaObs and tulpaRatio, sample the
observation model and populate `$draws$log_lik`, plus `$draws$y_rep` for
posterior predictive checks. On those fits, `waic()`, `loo()`, and
`compare_models(..., criterion = "waic")` work directly. The matrix they
read is right there.

The split is deliberate. The engine owns the structural model and scores
it by evidence, which it can compute from the Laplace fit alone. The
model packages own the observation likelihood and the predictive
machinery that rides on it. If you need cross-validated predictive
comparison, fit through the relevant model package; for evidence-based
comparison of structural specifications, the engine's marginal
likelihood is the tool, and it needs nothing beyond the Laplace fit you
already have.

## Practical guidance

A short set of habits covers most comparison work:

- **Fit the whole ladder on the Laplace tier.** It is deterministic and
  fast, and it puts every candidate's evidence on one definition. Move a
  model to a sampler only when you have a separate reason to doubt the
  Gaussian posterior shape, and remember that a sampler fit's `logLik()`
  is a mean log-density, not a marginal likelihood.
- **Nest your models.** Add one structural piece per rung and read the
  gap each piece buys. A non-nested set of models can still be ranked by
  evidence, but nested steps make the *reason* for each gain legible.
- **Hold `sigma_re` and `phi` fixed across the ladder.** The evidence
  conditions on the values you pass. If you change `sigma_re` between
  rungs, the comparison mixes a structural change with a hyperparameter
  change and the gaps no longer mean one thing. To compare across
  random-effect variances, integrate them through the nested-Laplace
  path instead of fixing them by hand.
- **Check convergence before reading evidence.** Scan the `converged`
  column from `glance()`. A non-converged Laplace fit has no trustworthy
  curvature, and its marginal likelihood inherits that.
- **Match the criterion to the question.** Evidence answers "which
  structure does this data support". WAIC and LOO answer "which model
  forecasts fresh observations best". The two verdicts usually coincide;
  where they diverge, the mismatch flags overfitting, since a model can
  hug its training rows yet generalise poorly. For the predictive
  question, route the analysis through a model package that supplies the
  pointwise log-likelihood.

## Where to go next

- The marginal likelihood and the Laplace approximation in depth: the
  inference-modes and algorithm vignettes.
- Integrating random-effect variances rather than fixing them: the
  nested-Laplace path and the `tgmrf()` vignette.
- WAIC, LOO, posterior-predictive checks, and stacking on a full
  observation model: the model packages (tulpaObs, tulpaRatio) built on
  this engine.
