---
title: "Comparing tulpa models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing tulpa models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260529)
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r load, message = FALSE}
library(tulpa)
```

## Why compare

You rarely fit one model. You fit a few: a plain trend, the trend
with a covariate, the same again with a grouping structure. Each
candidate encodes a different belief about what generated the data, and
the question that follows is always the same. Which structure does the
data support, and by how much?

tulpa answers that at the engine level with one number per fit: the log
marginal likelihood from a Laplace approximation. It is the model
evidence, the quantity that automatically rewards a better fit and
penalises wasted flexibility. This vignette walks through the criterion,
builds a nested ladder of models on data with a known generating
process, and reads the evidence as it climbs. It also says plainly where
the marginal likelihood stops and where WAIC and LOO take over, and what
the engine reconstructs behind the scenes to compute those two on a
plain `tulpa()` fit.

The plan is concrete. Simulate Gaussian data with an intercept, a slope,
a group-level offset, and a second covariate that all genuinely shape
the response. Fit the ladder. Watch the evidence rise at each rung.
Replay the exercise on a binary outcome so the verdict holds across two
families instead of resting on a single example.

Three quantities do the work in this vignette, and they answer different
questions. The marginal likelihood, read off a Laplace fit with
`logLik()`, scores a whole specification against the data it saw. The
Bayes factor turns two such scores into a ratio, the odds one structure
gives the data over another. WAIC and LOO step outside that frame and
ask how well a model predicts data it has not yet met. The first two the
engine computes from a single deterministic solve; the third needs a
pointwise log-likelihood that only a full observation model supplies, so
it lives in the model packages. Knowing which number answers your
question saves a great deal of confused staring at tables.

## Marginal likelihood from a Laplace fit

A Bayesian model assigns a probability to the data after the parameters
are integrated out:

$$
p(y \mid M) \;=\; \int p(y \mid \theta, M)\, p(\theta \mid M)\, d\theta .
$$

This is the marginal likelihood, also called the evidence. It sits in
the denominator of Bayes' theorem, and it grades a whole specification
rather than a single parameter value. A model whose prior mass piles up
where the observations actually fall scores high. A model that scatters
its prior across regions the data exclude pays a toll, since the integral
averages the likelihood against the prior and that wasted breadth drags
the average down. An extra coefficient earns its keep only when the
likelihood it unlocks outweighs the dilution it imposes. This trade is
Occam's razor, quantified: parsimony falls out of the arithmetic, and
you never tack on a manual complexity penalty.

The integral resists a closed form for most models. The Laplace
approximation tackles it through a tidy device: locate the posterior
mode, measure the curvature (the Hessian) there, and approximate the
integrand by a Gaussian bell centred on that peak. Under a Gaussian
likelihood paired with a Gaussian prior the bell is exact, so what comes
out is the genuine marginal likelihood. For heavier-tailed or skewed
families it becomes a second-order estimate, reliable whenever the
posterior stays roughly bell-shaped, which holds at most realistic
sample sizes.

The penalty in the approximation is worth seeing explicitly, because it
is where Occam's razor enters the arithmetic. Write the log integrand as
`g(theta) = log p(y | theta) + log p(theta)`. The Laplace value is
`g(theta_hat) + (d/2) log(2 pi) - (1/2) log det H`, where `theta_hat` is
the mode, `H` the negative Hessian of `g` there, and `d` the parameter
count. The first piece rewards a good fit at the best parameter values.
The `log det H` piece is the cost of flexibility: a sharply peaked
posterior, the kind a well-identified extra parameter produces, has a
large determinant and pays for itself only if the fit it buys is large
enough. A parameter the data barely constrain leaves the posterior nearly
as broad as the prior, so the determinant term grows while the fit term
barely moves, and the evidence falls. That is the same trade you would
get from integrating the parameter out exactly, which is the point: the
penalty is not bolted on, it is what the integral does.

This is why the evidence is not the maximised likelihood. The maximised
likelihood only ever rises when you add a parameter, so ranking by it
always picks the largest model. The marginal likelihood can fall when a
parameter is added, and it does fall whenever the parameter earns less
fit than it costs in posterior concentration. A criterion that can move
in both directions is the one you want for choosing structure.

When you fit with `mode = "laplace"`, tulpa computes this quantity and
stores it on the fit. `logLik()` returns it.

```{r first-evidence}
n  <- 500
x  <- rnorm(n)
y0 <- 0.5 + 1.2 * x + rnorm(n, sd = 0.8)
df <- data.frame(y = y0, x = x)

fit <- tulpa(y ~ x, data = df, family = "gaussian",
             mode = "laplace", phi = 0.8)
logLik(fit)
```

The `logLik` object carries two attributes worth knowing. `df` is the
number of fixed effects the fit estimated, and `nobs` is the number of
observations the evidence was computed on. Both matter for comparison:
an evidence number only means something against another fit on the same
data.

```{r evidence-attrs}
c(df = attr(logLik(fit), "df"), nobs = attr(logLik(fit), "nobs"))
```

A subtle point about the tiers. On the Laplace tier `logLik()` is the
log *marginal* likelihood, the evidence with parameters integrated out.
On a sampler tier (MALA, Gibbs, Pathfinder) the same accessor reports
the mean log-density across the draws, which is a different object and
not directly comparable to a marginal likelihood. For model comparison
by evidence, fit every candidate on the Laplace tier so the numbers
share a definition. That is also the fast path, which makes fitting a
whole ladder cheap.

## A nested model ladder

The clearest way to read evidence is to nest your candidates, each one
sitting inside the next, adding a single structural ingredient per step.
When an ingredient mirrors a real feature of the generating mechanism,
the evidence climbs once you include it. When it merely decorates, the
evidence stalls or slips. A ladder built on simulated truth converts the
criterion into something you verify rather than merely believe.

Simulate the truth first. The response has an intercept of 0.5, a slope
of 1.2 on `x`, a group-level offset with standard deviation 0.7 across
fifteen groups, and a second slope of 0.8 on `z`. Every term is real, so
a good criterion should reward each one as it enters.

```{r sim-ladder}
n  <- 500
x  <- rnorm(n)
z  <- rnorm(n)
g  <- factor(sample(1:15, n, replace = TRUE))
u  <- rnorm(15, sd = 0.7)
y  <- 0.5 + 1.2 * x + 0.8 * z + u[g] + rnorm(n, sd = 0.8)
dat <- data.frame(y = y, x = x, z = z, g = g)
```

Now the four rungs. The intercept-only model is the null reference. The
second rung adds the `x` slope. The third adds a random intercept on
`g`, conditioning on the random-effect standard deviation. The fourth
adds the second covariate `z`. Each fit is a single Laplace solve.

```{r fit-ladder}
m0 <- tulpa(y ~ 1, data = dat, family = "gaussian",
            mode = "laplace", phi = 0.8)
m1 <- tulpa(y ~ x, data = dat, family = "gaussian",
            mode = "laplace", phi = 0.8)
m2 <- tulpa(y ~ x + (1 | g), data = dat, family = "gaussian",
            mode = "laplace", sigma_re = 0.7, phi = 0.8)
m3 <- tulpa(y ~ x + z + (1 | g), data = dat, family = "gaussian",
            mode = "laplace", sigma_re = 0.7, phi = 0.8)
```

Pull the evidence from each and lay it side by side. The differences
between adjacent rungs are what to read: each is the evidence gained by
adding one true piece.

```{r ladder-evidence}
ll <- c(intercept = as.numeric(logLik(m0)),
        slope     = as.numeric(logLik(m1)),
        slope_re  = as.numeric(logLik(m2)),
        full      = as.numeric(logLik(m3)))
data.frame(model = names(ll), logLik = round(ll, 1),
           gain = c(NA, round(diff(ll), 1)))
```

Every step up the ladder lifts the evidence, and by a wide margin. The
`x` slope delivers the largest single jump, befitting its sizeable true
coefficient, while the random intercept and the second covariate each
contribute their own distinct signal that the evidence faithfully
registers. The ordering reproduces the generating recipe precisely. That
is exactly why running the comparison against a known truth pays off.

The gains here are enormous because every added term is genuinely
present and the sample is large. Read the *sign and size* of each gain,
not its precise value. A large positive gain says the term earns its
place. A gain near zero, or negative, would say the term adds structure
the data do not ask for.

A plot of the ladder makes the climb easier to read than a column of
numbers. Each bar is one rung's evidence relative to the null, so the bar
heights are the cumulative gain and the steps between them are the
per-rung gains from the table.

```{r ladder-plot, eval = has_ggplot, fig.alt = "Bar chart of cumulative log evidence by ladder rung"}
library(ggplot2)
pd <- data.frame(model = factor(names(ll), levels = names(ll)),
                 rel = ll - ll[1])
ggplot(pd, aes(model, rel)) +
  geom_col(width = 0.6, fill = "#3a6ea5") +
  labs(x = NULL, y = "log evidence vs. null") +
  theme_minimal() +
  theme(panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA))
```

The first step, from intercept to slope, dominates the figure, which
matches the table: the `x` coefficient is large and the sample is wide
enough to pin it down. The random intercept and the second covariate add
visibly shorter steps, each still a clear lift. Plotting the cumulative
gain rather than the raw `logLik` keeps the axis readable, since the raw
numbers run to several hundred and would crowd the small steps off the
top of the panel.

To see the penalty bite, add a covariate that is pure noise, unrelated to
the response, and watch the evidence refuse to reward it. The full model
already holds the two real slopes and the random intercept; the extra
term `w` is independent of `y` by construction.

```{r noise-term}
dat$w <- rnorm(n)
m4 <- tulpa(y ~ x + z + w + (1 | g), data = dat, family = "gaussian",
            mode = "laplace", sigma_re = 0.7, phi = 0.8)
c(full = round(as.numeric(logLik(m3)), 2),
  plus_noise = round(as.numeric(logLik(m4)), 2),
  gain = round(as.numeric(logLik(m4)) - as.numeric(logLik(m3)), 2))
```

The noise covariate buys a gain near zero, often slightly negative once
the determinant penalty is paid. A maximised likelihood would have risen
here, since `w` can always soak up a sliver of residual variance at its
best-fitting coefficient. The marginal likelihood declines to pay,
because the fit `w` unlocks is smaller than the posterior concentration
it costs. That asymmetry is Occam's razor doing its job on a term you
know to be useless.

## compare_models()

Computing `logLik()` per fit and assembling a table by hand works, and
`compare_models()` does it for you. Pass the fits as named arguments and
choose the criterion.

```{r compare}
compare_models(intercept = m0, slope = m1,
               slope_re = m2, full = m3, criterion = "loglik")
```

The table reports the model name, the parameter count, and the log
marginal likelihood. The names come straight from the arguments, so
label them in a way that reads as a ladder.

Read the `n_params` column with care. It counts fixed effects. The
random-intercept model `m2` shows the same count as the slope-only model
`m1`, because the random intercept is a structural term conditioned on a
fixed `sigma_re` rather than an estimated fixed effect. The evidence,
which integrates over the group offsets, still rewards the structure;
the column simply does not count it as a fixed parameter. When you read
the table, lean on the `logLik` differences and treat `n_params` as a
fixed-effect bookkeeping figure.

How large a difference counts as real? Evidence lives on the log scale,
so a gap of `d` means the winning model is `exp(d)` times more probable
given the data, before any preference between the models themselves. A
handful of log units already settles the matter; the leaps in this
ladder run into the hundreds. Narrow margins, a unit or two, warrant
caution, particularly where the added term strains the approximation,
say a sharply skewed non-Gaussian response whose posterior departs from
the bell shape.

The same call shape works for a binary response. Simulate a logistic
process with a real intercept, a real `x` slope, and a real `z` slope,
then fit the same three-rung ladder.

```{r binom-ladder}
set.seed(101)
xb  <- rnorm(n); zb <- rnorm(n)
yb  <- rbinom(n, 1, plogis(-0.4 + 1.1 * xb + 0.9 * zb))
dfb <- data.frame(y = yb, x = xb, z = zb)

b0 <- tulpa(y ~ 1,     data = dfb, family = "binomial", mode = "laplace")
b1 <- tulpa(y ~ x,     data = dfb, family = "binomial", mode = "laplace")
b2 <- tulpa(y ~ x + z, data = dfb, family = "binomial", mode = "laplace")
compare_models(intercept = b0, x = b1, xz = b2, criterion = "loglik")
```

Both covariates were in the generating process, and the evidence climbs
as each enters. For the binomial family the marginal likelihood is a
Laplace approximation rather than an exact integral, so the numbers
carry a small approximation error. With a well-behaved posterior, as
here, that error is far smaller than the gaps between models, and the
ranking is safe to read.

### Non-nested candidates

Nesting makes each gain legible, but the evidence does not require it.
Two models that are not versions of each other, say one covariate against
a different covariate, still produce comparable marginal likelihoods as
long as they saw the same response on the same rows. The criterion ranks
them on the same scale; what you lose is the clean reading of one term's
contribution, since the difference now bundles together everything that
changed between the two specifications.

```{r non-nested}
mx <- tulpa(y ~ x, data = dat, family = "gaussian",
            mode = "laplace", phi = 0.8)
mz <- tulpa(y ~ z, data = dat, family = "gaussian",
            mode = "laplace", phi = 0.8)
compare_models(x_only = mx, z_only = mz, criterion = "loglik")
```

The `x`-only model wins, which fits the simulation: `x` carries the
larger true coefficient, so it explains more of the response than `z`
does on its own. A non-nested comparison answers "which of these
explains the data better" cleanly; it just does not decompose into
per-term gains the way a ladder does.

## Interpreting logLik differences

A worked reading of one gap makes the scale concrete. Take the binomial
ladder: adding `z` to the `x`-only model raised the evidence.

```{r interpret-gap}
gain <- as.numeric(logLik(b2)) - as.numeric(logLik(b1))
c(log_evidence_gain = round(gain, 2),
  evidence_ratio    = round(exp(gain), 1))
```

The evidence ratio is the Bayes factor for the larger model against the
smaller, assuming equal prior weight on the two. A ratio well above one
favours the richer model. The log scale is the comfortable place to work
because the numbers stay manageable and differences add up across nested
steps. Three rules of thumb help when you read a column of evidence:

- A difference of a fraction of a unit is noise. Treat the two models as
  tied and prefer the simpler one.
- A difference of a few units is meaningful support for the better
  model. The Bayes factor is in the tens or more.
- A difference of tens or hundreds, as in the Gaussian ladder, is
  decisive. The structure is clearly in the data.

Always compare on the same data, the same family, and the same tier. The
evidence is a property of a model *and* the data it saw; numbers from
fits on different rows, different responses, or different inference tiers
are not on a common scale.

A fuller set of numeric anchors, beyond the three above, helps when the
gaps are awkward. These follow the long-standing Jeffreys grading of
Bayes factors, read here as differences on the natural-log evidence
scale. A log gap below about 1 (Bayes factor under 3) is weak, barely
worth a preference. A gap between 1 and 3 (Bayes factor 3 to 20) is
positive support, enough to lean on the larger model. A gap between 3 and
5 (Bayes factor 20 to 150) is strong. Above 5 (Bayes factor past 150) the
support is decisive, and the Gaussian ladder above clears that bar by two
orders of magnitude. These bands are conventions, not laws, but they keep
you from over-reading a gap of half a unit or under-reading one of six.

## Bayes factors and a bridge-sampling cross-check

The Bayes factor between two models is the ratio of their marginal
likelihoods, so on the log scale it is the plain difference the table
already shows. For the binomial ladder, the factor for the two-covariate
model over the one-covariate model is the exponentiated gap, and the
posterior odds it implies depend on the prior odds you assign the models:

$$
\frac{p(M_2 \mid y)}{p(M_1 \mid y)}
  \;=\;
  \underbrace{\frac{p(y \mid M_2)}{p(y \mid M_1)}}_{\text{Bayes factor}}
  \;\times\;
  \frac{p(M_2)}{p(M_1)} .
$$

With equal prior weight on the two models the posterior odds equal the
Bayes factor, which is why the evidence ratio reads directly as relative
support.

A Laplace marginal likelihood is an approximation for non-Gaussian
families, so when a verdict is close it is worth a second estimate from a
different route. The engine ships `bridge_sampling()`, which estimates
the log marginal likelihood from posterior draws and an unnormalised
log-posterior, by the Meng-Wong / Gronau identity. It needs a fit that
carries draws, so fit the candidate on a sampler tier rather than the
Laplace tier, then hand it the draws and a function returning
`log p(theta, y)` at one parameter vector. The toy below estimates the
evidence of a one-parameter conjugate Gaussian, where the closed form is
known, so you can read the agreement directly.

```{r bridge}
y_obs   <- 1.5
log_post <- function(theta)
  dnorm(y_obs, theta, 1, log = TRUE) + dnorm(theta, 0, 10, log = TRUE)
post_sd  <- sqrt(100 / 101)
draws    <- matrix(rnorm(4000, y_obs * 100 / 101, post_sd), ncol = 1)
bs <- bridge_sampling(draws, log_post)
c(bridge = round(bs$log_marginal, 3),
  exact  = round(dnorm(y_obs, 0, sqrt(101), log = TRUE), 3))
```

The two numbers land within a few thousandths, which is the calibration
you want from a marginal-likelihood estimator on a problem with a known
answer. On a real model with no closed form, a bridge estimate that
tracks the Laplace value tells you the posterior is bell-shaped enough
for the Laplace evidence to be trusted; a bridge estimate that pulls away
from it flags a skewed or heavy-tailed posterior where the Laplace number
needs care. The `re_sd` field on the returned list is a rough relative
error, smaller meaning a tighter estimate. Bridge sampling is the
heavier of the two tools, since it wants a stack of draws and a callable
log-posterior, so reach for it as a check on a close call rather than as
the everyday criterion.

## glance() per model

`glance()` gives a one-row summary of any single fit, which is handy for
folding several fits into a tidy frame next to the comparison. On a
Laplace fit it reports the fixed-effect count, the log marginal
likelihood, and a convergence flag; the sampler-only fields stay `NA`
because the Laplace tier carries no draws.

```{r glance-one}
glance(m3)
```

Stack the ladder into one frame by binding the per-model rows. This pairs
naturally with `compare_models()`: the comparison ranks, while `glance()`
carries the per-fit detail such as convergence.

```{r glance-stack}
do.call(rbind, Map(function(m, nm) cbind(model = nm, glance(m)),
                   list(m0, m1, m2, m3),
                   c("intercept", "slope", "slope_re", "full")))
```

The `converged` column is the one to scan before trusting any evidence
number. A Laplace fit that did not converge has not found the mode, so
its curvature, and the marginal likelihood that depends on it, are not
to be read. On the sampler tiers `glance()` also reports `mean_accept`
and the divergence count, which play the same gatekeeping role for those
fits.

Where `glance()` summarises the whole fit in one row, `tidy()` opens up
the fixed effects of a single model, one row per coefficient with an
estimate, a standard error, and credible bounds. The two pair naturally:
the comparison and `glance()` rank specifications, then `tidy()` reads
off what the winner actually estimated.

```{r tidy-winner}
tidy(m3)
```

The slope on `x` sits near its true 1.2 and the slope on `z` near its
true 0.8, with the intercept near 0.5, so the model that won the ladder
also recovers the generating coefficients. A specification can rank first
on evidence and still hide a coefficient pinned far from where you
expected it, which is the kind of thing `tidy()` surfaces and a single
evidence number cannot.

## What WAIC and LOO need

The marginal likelihood scores a model on the data it was fit to. WAIC
and LOO estimate something different and complementary: how well the
model predicts data it has not seen. Both are built from the *pointwise*
log-likelihood, one value per observation per posterior draw, arranged
in a draws-by-observations matrix. From that matrix the loo package
computes the expected log predictive density and its information
criteria.

The pointwise structure is the part a fit summary does not hold, and
the reason is worth stating. WAIC estimates the out-of-sample
predictive density by summing the log pointwise predictive density over
observations and subtracting a variance-based penalty computed per
observation. LOO approximates leave-one-out cross-validation by
reweighting the existing draws for each held-out point, using
Pareto-smoothed importance sampling on the per-observation likelihood
ratios. Both need to see how the log-likelihood of each single
observation varies across the posterior draws, which is information a
collapsed summary throws away. A marginal likelihood is one scalar for
the whole dataset; it has already integrated over the observations and
the parameters together, so there is no per-observation, per-draw stack
left to read.

A base `tulpa()` fit does not store that matrix. The Laplace tier stores
a mode and a curvature, not a stack of draws, and even the sampler tiers
store parameter draws rather than a per-observation log-likelihood. So
an external tool that expects the stored matrix has nothing to read on a
plain fit, and it says so:

```{r waic-needs, error = TRUE}
loo::waic(b2)
```

The message names the missing piece: a stored pointwise log-likelihood.
Model packages built on this engine, such as tulpaObs and tulpaRatio,
sample the observation model and populate `$draws$log_lik`, plus
`$draws$y_rep` for posterior predictive checks, so external tools that
read that matrix work on their fits directly. The engine's own criteria
layer -- `compare_models(..., criterion = "waic")`, `tulpa_criteria()`,
and `model_average()` -- goes one step further for fits with a builtin
family: when the stored matrix is absent, it reconstructs the pointwise
log-likelihood from the linear-predictor posterior draws and the family
density, so WAIC and LOO comparison also work on plain `tulpa()` fits.
For the Laplace tier those draws come from the Gaussian approximation at
the mode, so the resulting WAIC inherits that approximation.

The split is deliberate. The engine owns the structural model and scores
it by evidence, which it can compute from the Laplace fit alone. The
model packages own the observation likelihood and the predictive
machinery that rides on it. If you need cross-validated predictive
comparison, fit through the relevant model package; for evidence-based
comparison of structural specifications, the engine's marginal
likelihood is the tool, and it needs nothing beyond the Laplace fit you
already have.

## model_average() for prediction

When no single model clearly wins, averaging their predictions can beat
committing to one. `model_average()` takes named fits, computes a weight
per model from the chosen criterion, and returns the weighted-average
fitted values alongside the weights and the comparison table. The fitted
values come from `fitted()` on each fit, so every candidate must carry
its design matrix, which a `tulpa()` fit does.

```{r model-average}
ma <- model_average(slope = m1, slope_re = m2, full = m3,
                    weights = "waic")
ma$weights
head(round(ma$averaged, 3))
```

The weights are genuine WAIC weights: `model_average()` reads the same
pointwise log-likelihood the criteria layer reconstructs from each fit's
linear-predictor draws, so predictive accuracy, not a uniform split,
decides each model's share. The `averaged` vector is the per-observation
blend you would use as a combined prediction; `ma$comparison` is the
table the weights were read from.

## When comparison is not meaningful

A few situations break the comparison quietly, returning numbers that
look fine and mean nothing. The first is a difference in data. The
evidence is a joint property of the model and the rows it saw, so two
fits on different subsets, different responses, or even the same response
after a transform are not on a shared scale. Dropping rows with a missing
covariate in one model but not another is the common version of this trap,
since the smaller dataset is genuinely a different `y`. Refit every
candidate on identical rows before lining up their evidence.

The second is a difference in family or link. A Gaussian fit and a
binomial fit assign probability to different sample spaces, a density in
one case and a mass function in the other, so their log marginal
likelihoods are not comparable even on the same `y`. The third is a
difference in tier. A `mode = "laplace"` fit reports a marginal
likelihood through `logLik()`; a sampler fit reports a mean log-density,
a different object entirely. Mixing the two in one `compare_models()`
call lines up two incompatible quantities under one column heading.

The fourth is subtler. When the Laplace approximation is poor for one
candidate but fine for another, the gap between them blends a real
structural difference with an approximation artefact. This bites hardest
on small samples with sharply skewed posteriors, where the bell-shaped
Gaussian around the mode misrepresents the integral. The bridge-sampling
cross-check above is the tool for catching it: if the bridge and Laplace
estimates agree for both models, the gap is structural; if they diverge
for one, treat that model's evidence as provisional. A high Pareto-k from
the engine's accuracy diagnostics on a nested-Laplace fit is the same
warning from a different angle.

## Practical guidance

A short set of habits covers most comparison work:

- **Fit the whole ladder on the Laplace tier.** It is deterministic and
  fast, and it puts every candidate's evidence on one definition. Move a
  model to a sampler only when you have a separate reason to doubt the
  Gaussian posterior shape, and remember that a sampler fit's `logLik()`
  is a mean log-density, not a marginal likelihood.
- **Nest your models.** Add one structural piece per rung and read the
  gap each piece buys. A non-nested set of models can still be ranked by
  evidence, but nested steps make the *reason* for each gain legible.
- **Hold `sigma_re` and `phi` fixed across the ladder.** The evidence
  conditions on the values you pass. If you change `sigma_re` between
  rungs, the comparison mixes a structural change with a hyperparameter
  change and the gaps no longer mean one thing. To compare across
  random-effect variances, integrate them through the nested-Laplace
  path instead of fixing them by hand.
- **Know what conditioning on a hyperparameter costs.** When you pass
  `sigma_re = 0.7` or `phi = 0.8`, the marginal likelihood is the
  evidence given that value, not integrated over it. The Gaussian
  random-effect offsets and the fixed effects are integrated out; the
  variance and the residual scale are held. That is exactly the right
  object for a ladder where you want each rung to differ only in
  structure, and it is the wrong object if the hyperparameter itself is
  the thing in question. A fit at `sigma_re = 0.7` cannot tell you
  whether `0.5` or `1.0` fits better, because both were excluded by the
  conditioning. For that question the nested-Laplace path places a grid
  over the hyperparameter and integrates, and `logLik()` on that fit
  returns the log-sum-exp over the grid, the genuinely integrated
  evidence rather than a conditional one. The same caution covers `phi`:
  a Gaussian ladder run at one `phi` ranks structures cleanly, but it
  says nothing about which residual scale the data prefer.
- **Check convergence before reading evidence.** Scan the `converged`
  column from `glance()`. A non-converged Laplace fit has no trustworthy
  curvature, and its marginal likelihood inherits that.
- **Match the criterion to the question.** Evidence answers "which
  structure does this data support". WAIC and LOO answer "which model
  forecasts fresh observations best". The two verdicts usually coincide;
  where they diverge, the mismatch flags overfitting, since a model can
  hug its training rows yet generalise poorly. For the predictive
  question, route the analysis through a model package that supplies the
  pointwise log-likelihood.

## Where to go next

- The marginal likelihood and the Laplace approximation in depth: the
  inference-modes and algorithm vignettes.
- Integrating random-effect variances rather than fixing them: the
  nested-Laplace path and the `tgmrf()` vignette.
- WAIC, LOO, posterior-predictive checks, and stacking on a full
  observation model: the model packages (tulpaObs, tulpaRatio) built on
  this engine.
