---
title: "Specifying and checking priors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specifying and checking priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260529)
```

```{r load, message = FALSE}
library(tulpa)
```

## Why priors get their own object

Every Bayesian fit carries priors, whether you write them down or not. tulpa
keeps them in one place: a `tulpa_priors` object that names a prior for each
kind of parameter the engine knows about. You build it with `tulpa_priors()`,
fill the slots with `prior_*()` constructors, and hand the result to the
functions that need it. The same object drives prior predictive simulation,
so the priors you check are the priors you fit with.

This vignette walks through the prior families, the defaults, how to move a
fixed-effect prior and watch the posterior follow, and how to look at what a
prior implies about the data before any of it touches the likelihood. The
closing section turns that into a short set of working rules.

## Prior families available

A prior in tulpa is a small tagged list built by one of seven constructors.
Each one carries a distribution name and its parameters, and prints in a
compact form. The seven cover three jobs: location parameters that range over
the whole real line, scale parameters that must stay positive, and proportions
bounded in the unit interval.

### Location: the normal prior

`prior_normal(mean, sd)` is the workhorse for fixed effects, which live on the
link scale and can take either sign. The default is `prior_normal(0, 2.5)`.

```{r prior-normal}
prior_normal(0, 2.5)
prior_normal(0, 1)
```

The mean sets where the prior is centred and the standard deviation sets how
far it reaches. On a logit or log link, a standard deviation of 2.5 already
covers a wide range of multiplicative effects, so this is weakly informative
rather than flat.

### Scale: priors for standard deviations

Random-effect standard deviations, dispersion parameters, and any other
positive quantity need a prior with support on the positive half-line. tulpa
offers four.

```{r prior-scale}
prior_half_normal(1)
prior_half_cauchy(2.5)
prior_gamma(2, 0.1)
prior_exponential(1)
```

`prior_half_normal(sd)` folds a mean-zero normal at the origin: light tails,
firm shrinkage toward small values. `prior_half_cauchy(scale)` has the same
mode at zero but much heavier tails, so it tolerates the occasional large
standard deviation that a half-normal would fight. The half-Cauchy is a common
default for variance parameters precisely because it stays gentle in the tail.

`prior_gamma(shape, rate)` is the flexible positive prior. Its mean is
`shape / rate` and its variance is `shape / rate^2`, both printed alongside the
parameters. With `shape = 2, rate = 0.1` the mean sits at 20, a deliberately
loose prior for a quantity you expect to be large. `prior_exponential(rate)`
is the one-parameter special case (a gamma with `shape = 1`); its mean is
`1 / rate`.

### The PC prior

`prior_pc(U, alpha)` builds a penalised-complexity prior, the recommended
choice for variance components. You give it a value `U` you would consider
large and a small tail probability `alpha`, and it encodes the statement
`P(parameter > U) = alpha`.

```{r prior-pc}
prior_pc(U = 1, alpha = 0.01)
prior_pc(U = 0.5, alpha = 0.05)
```

The print-out shows both the readable statement and the exponential it reduces
to: the rate is `-log(alpha) / U`. The penalised-complexity idea, from Simpson
et al. (2017), treats a model component as a deviation from a simpler base
model (here, a standard deviation of zero, meaning the component switches off).
The prior puts an exponential penalty on the distance from that base model on
a natural scale, which translates to an exponential prior on the standard
deviation. Two properties follow. The prior shrinks toward the simpler model
unless the data pull it away, and its single tuning knob is a tail probability
you can reason about directly. Saying "a standard deviation above 1 should
happen one percent of the time" is easier to defend than picking gamma shape
and rate by feel.

The shrinkage matters most when groups are few. A handful of groups carry
little information about their shared variance, and a flat or heavy prior on
that variance lets the random effect overfit, soaking up noise that belongs in
the residual. A PC prior leans the other way by default, switching the
component off unless the groups disagree enough to keep it on. That is the
behaviour you want from a regulariser: invisible when the data speak, firm when
they do not.

### Proportions: the beta prior

`prior_beta(alpha, beta)` covers parameters bounded in `(0, 1)`, such as an
AR(1) correlation rescaled to the unit interval or the BYM2 mixing proportion.

```{r prior-beta}
prior_beta(1, 1)
prior_beta(2, 2)
prior_beta(5, 2)
```

`prior_beta(1, 1)` is uniform on the interval. `prior_beta(2, 2)` is symmetric
and peaked at 0.5. Pushing the first shape above the second, as in
`prior_beta(5, 2)`, tilts the mass toward 1, which suits a parameter you
believe is high (strong temporal autocorrelation, say). The printed mean,
`alpha / (alpha + beta)`, gives a quick read on where the prior sits.

### What each family is for

A short map from parameter to natural prior family:

| Parameter | Lives on | Natural priors |
|---|---|---|
| Fixed effects (`beta`) | whole real line | `prior_normal()` |
| Random-effect SD (`sigma`) | positive | `prior_pc()`, `prior_half_normal()`, `prior_half_cauchy()` |
| Dispersion / precision (`phi`) | positive | `prior_pc()`, `prior_gamma()`, `prior_exponential()` |
| Temporal AR(1) (`rho_temporal`) | `(0, 1)` | `prior_beta()` |
| Spatial mixing (`rho_spatial`) | `(0, 1)` | `prior_beta()` |

Matching a positive prior to a positive parameter is on you: the constructors
accept any combination, so the table is the convention to follow rather than a
rule the code checks. A normal prior paired with a standard deviation slot will
build without complaint and then misbehave at fit time, so treat the support of
each parameter as the first thing to get right.

## Default priors

`tulpa_priors()` with no arguments returns the package defaults. Print it to
see all five slots at once.

```{r defaults}
tulpa_priors()
```

The defaults read as follows. Fixed effects get `prior_normal(0, 2.5)`, wide
enough to let coefficients roam across the plausible link-scale range without
being flat. Random-effect standard deviations get a PC prior with
`P(sigma > 1) = 0.01`, which favours smaller variance components and guards
against the variance running off when groups are few. The dispersion `phi`
gets a PC prior with `P(phi > 10) = 0.01`. Temporal correlation defaults to
`prior_beta(2, 2)`, symmetric around 0.5, and the spatial mixing proportion to
`prior_beta(1, 1)`, a flat unit prior with no lean toward structured or
unstructured variation.

`priors_default()` prints the same defaults with a sentence of interpretation
and the customisation entry point for each slot.

```{r defaults-annotated}
priors_default()
```

Change one slot by naming it; the rest stay at their defaults.

```{r defaults-custom}
tulpa_priors(
  beta  = prior_normal(0, 1),
  sigma = prior_pc(U = 0.5, alpha = 0.01)
)
```

Because each slot validates its argument, passing something that is not a
`tulpa_prior` object fails early with a clear message rather than surfacing
deep inside a fit.

## Priors on fixed effects

The engine's Laplace path takes a Gaussian prior on the fixed effects through
the `beta_prior` argument of `tulpa()`, written as `list(mean =, sd =)`. The
`sd` is required; `mean` defaults to 0. Each may be a scalar applied to every
coefficient or a vector with one entry per coefficient. Leaving `beta_prior`
at `NULL` keeps a weak built-in prior that barely moves the fit.

A prior earns its keep when the data are thin. Simulate a small Gaussian
dataset with a slope of 1.2, the kind of sample where the likelihood alone
leaves real uncertainty.

```{r fe-sim}
n  <- 25
x  <- rnorm(n)
y  <- 0.5 + 1.2 * x + rnorm(n, sd = 1.5)
df <- data.frame(y = y, x = x)
```

Fit it once with the default weak prior. The slope estimate is driven by the
likelihood.

```{r fe-weak}
fit_weak <- tulpa(y ~ x, data = df, family = "gaussian",
                  mode = "laplace", phi = 1.5^2)
coef(fit_weak)
```

Now impose a tight prior that says the slope is near zero: mean 0 and a small
standard deviation. The intercept keeps a loose prior, so only the slope is
pulled.

```{r fe-tight}
fit_tight <- tulpa(y ~ x, data = df, family = "gaussian",
                   mode = "laplace", phi = 1.5^2,
                   beta_prior = list(mean = c(0, 0), sd = c(10, 0.2)))
coef(fit_tight)
```

The slope under the tight prior sits well below its weakly informed value: the
prior at zero and the likelihood at the data strike a compromise, and the small
prior standard deviation gives the prior most of the weight. Put the two fits
side by side.

```{r fe-compare}
data.frame(
  term  = names(coef(fit_weak)),
  weak  = round(coef(fit_weak), 3),
  tight = round(coef(fit_tight), 3)
)
```

This is regularisation made explicit. Shrinking a coefficient toward a value
accepts a little bias in exchange for lower variance, which is worth doing when
the data cannot pin the coefficient down on their own. The credible intervals
tell the same story from the uncertainty side.

```{r fe-confint}
confint(fit_weak)["x", ]
confint(fit_tight)["x", ]
```

The tight prior narrows the slope's interval and shifts it toward zero. A
vector `sd` lets you regularise some coefficients hard while leaving others
free, which is the usual pattern: pin nuisance terms, let the effects of
interest follow the data.

The fixed-effect prior threads through the plain Laplace and sampler paths.
Some structured paths (spatial fields, the SPDE integrator) carry their own
built-in fixed-effect prior and will tell you so if you pass `beta_prior`
there.

## Prior predictive checks

A prior is a statement about parameters, but its consequences land on the data
scale, often somewhere surprising. A prior that looks mild on a log link can
imply counts in the millions. The way to catch this is a prior predictive
check: draw parameters from the priors, push them through the linear predictor,
simulate responses, and look at the responses.

`prior_predict()` does the drawing. It needs a formula, a `tulpa_family` that
supplies a simulator, the data (for its design and dimensions, not its
response), and the priors. The family is a thin object built by
`tulpa_family()`: a name plus a `simulate_fn(eta, params, n_obs, ...)` that
turns a linear predictor into a response. Build a Poisson family whose
simulator exponentiates the linear predictor.

```{r pp-family}
pois <- tulpa_family(
  name = "poisson",
  simulate_fn = function(eta, params, n_obs, ...) rpois(n_obs, exp(eta[[1]]))
)
```

Use a small covariate frame; the response column can be a placeholder because
the priors generate the response.

```{r pp-data}
dat <- data.frame(y = rep(0, 60), x = rnorm(60))
```

Draw from a deliberately vague prior on the fixed effects. A standard deviation
of 5 on a log link is enormous: it lets the intercept and slope wander far
enough to produce astronomically large counts.

```{r pp-vague}
pp_vague <- prior_predict(
  y ~ x, family = pois, data = dat, n_draws = 200,
  priors = tulpa_priors(beta = prior_normal(0, 5)), seed = 1
)
pp_vague
```

The returned object holds one simulated dataset per draw in `pp_vague$y`. Look
at the largest count any draw produced.

```{r pp-vague-range}
max(vapply(pp_vague$y, max, numeric(1)))
```

That number is far past anything a real count process would produce. The prior
is too vague: it places real mass on data the model should consider impossible.
Tighten the fixed-effect prior to a standard deviation of 1 and redraw.

```{r pp-sensible}
pp_ok <- prior_predict(
  y ~ x, family = pois, data = dat, n_draws = 200,
  priors = tulpa_priors(beta = prior_normal(0, 1)), seed = 1
)
max(vapply(pp_ok$y, max, numeric(1)))
```

The largest simulated count now sits in a range a count model can take
seriously. Quantiles across all draws sharpen the contrast.

```{r pp-quantiles}
vague_all <- unlist(pp_vague$y)
ok_all    <- unlist(pp_ok$y)
rbind(
  vague    = quantile(vague_all, c(0.5, 0.9, 0.99)),
  sensible = quantile(ok_all,    c(0.5, 0.9, 0.99))
)
```

The median is modest under both priors, but the upper tail under the vague
prior is orders of magnitude heavier. The vague prior sits at the right centre
and fails on its width: the damage hides in the tail, where a glance at the
median would miss it entirely.

`plot()` overlays the simulated datasets so you can see the spread directly.
Capping the counts keeps the vague-prior plot readable, since a handful of
draws would otherwise stretch the axis past everything else.

```{r pp-plot, fig.alt = "Prior predictive draws under a vague Normal(0, 5) prior"}
pp_capped <- pp_vague
pp_capped$y <- lapply(pp_vague$y, function(yi) pmin(yi, 200))
plot(pp_capped, max_draws = 40)
```

```{r pp-plot-ok, fig.alt = "Prior predictive draws under a sensible Normal(0, 1) prior"}
plot(pp_ok, max_draws = 40)
```

The sensible-prior draws cluster in a believable band; the vague-prior draws,
even capped, spread across the full range. Reading the `$y` numbers and the
overlay together is the check: if the prior predictive covers values the
response could never take, the prior is too vague, and the families above let
you pull it in. The same machinery extends to any family by swapping the
`simulate_fn`, and to random-effect spread by setting the `sigma` prior, which
`prior_predict()` draws and feeds into the group effects.

## Choosing priors

A working order of operations for setting priors in tulpa:

- **Start from the defaults and change what you have reason to change.**
  `tulpa_priors()` is weakly informative across the board: a wide normal on
  fixed effects, PC priors that shrink variance components, symmetric priors on
  correlations. For a first fit on a healthy dataset, the defaults are a
  reasonable place to stand.

- **Reach for PC priors on variance components.** `prior_pc(U, alpha)` lets you
  state the prior as a tail probability, which is the parameter you can defend
  in a methods section. Set `U` to a standard deviation you would call large on
  the link scale and `alpha` to a small probability of exceeding it. When you
  expect the occasional genuinely large standard deviation, reach for
  `prior_half_cauchy()`: its heavier tail tolerates that case where the firm
  PC penalty would fight it.

- **Make fixed-effect priors carry weight only when the data are thin.** With
  hundreds of informative rows the likelihood dominates and `beta_prior` barely
  moves the fit, but with few rows, a poorly separated predictor, or a
  coefficient you want to regularise, a tight `beta_prior` becomes the lever the
  worked example demonstrated, and a vector `sd` lets you pull selected
  coefficients hard while leaving the rest free.

- **Run a prior predictive check before trusting a prior on the link scale.**
  Link-scale intuition is unreliable. A standard deviation that feels mild can
  imply impossible data once exponentiated, and the failure shows up in the
  upper tail rather than the centre, so a summary mean will not warn you.
  `prior_predict()` plus a one-line `tulpa_family()` simulator turns the
  question into numbers you can read off `$y` and an overlay you can scan;
  tighten until the simulated data look like data the process could produce.

- **Match the prior family to the parameter's support.** Normal for the real
  line, a positive prior (PC, half-normal, half-Cauchy, gamma, exponential) for
  standard deviations and dispersions, beta for proportions; the constructors
  accept any pairing, so a normal on a variance slot will build cleanly and then
  bite at fit time. The map in the families section is the safe default to copy.

- **Keep the prior object you checked.** A single `tulpa_priors` object feeds
  both `prior_predict()` and the fit, so a prior that survives the predictive
  check is the prior you go on to fit with, with no second specification drifting
  out of sync behind it.

## See also

- `?tulpa_priors`, `?prior_normal`, `?prior_pc` and the other `prior_*`
  constructor pages.
- `?prior_predict` and `?tulpa_family` for prior predictive simulation.
- The getting-started vignette for fitting, prediction, and model comparison.
- Simpson, D., Rue, H., Riebler, A., Martins, T. G., & Sorbye, S. H. (2017).
  Penalising model component complexity. *Statistical Science*, 32(1), 1-28.
