---
title: "Getting started with tulpa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with tulpa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260529)
```

```{r load, message = FALSE}
library(tulpa)
```

## What tulpa fits

tulpa fits Bayesian hierarchical models from a formula. You write
`y ~ x + (1 | g)`, choose a family, and `tulpa()` parses the formula,
builds the design, picks an inference backend, and returns a fitted
object you can summarise, predict from, and compare.

The engine carries the structural pieces that hierarchical models reuse:
fixed effects, random intercepts and slopes, spatial fields (SPDE, NNGP,
HSGP, ICAR, BYM2), temporal structure, and user-defined latent blocks. It
also carries a tier system for *how* the posterior is computed: a fast
Laplace approximation, a gradient sampler, or a Pathfinder fit, each with
a stated epistemic guarantee rather than a silent default.

The observation likelihood is the one piece tulpa keeps deliberately
small: five built-in families (`gaussian`, `binomial`, `poisson`,
`neg_binomial_2`, `beta`). Model packages plug their own likelihoods into the same engine
and add the predictive machinery (posterior-predictive checks, WAIC, LOO)
on top. This vignette stays on the engine: simulate, fit, extract,
predict, compare.

## A first model

Start with a Gaussian response and known coefficients, so the fit has a
truth to recover. The intercept is 0.5 and the slope is 1.2.

```{r sim-gaussian}
n  <- 400
x  <- rnorm(n)
y  <- 0.5 + 1.2 * x + rnorm(n, sd = 0.8)
df <- data.frame(y = y, x = x)
```

Fit with `tulpa()`. `mode = "laplace"` is the Tier-2 Gaussian
approximation: it finds the posterior mode and the curvature there, which
for a Gaussian likelihood is exact and returns immediately.

```{r fit-gaussian}
fit <- tulpa(y ~ x, data = df, family = "gaussian",
             mode = "laplace", phi = 0.8)
coef(fit)
```

The coefficients land near the truth. `summary()` adds the posterior
standard error and the 2.5%/97.5% credible bounds for each fixed effect.

```{r summary-gaussian}
summary(fit)
```

`confint()` returns the same bounds as a matrix, and `vcov()` the
fixed-effect covariance. Both read the fixed-effect block of the Hessian
that the Laplace fit already computed, so there is no extra sampling step.

```{r confint-gaussian}
confint(fit)
```

The `phi` argument is the family's dispersion: residual variance for
`gaussian`, the size parameter for `neg_binomial_2`, the precision for
`beta`. The Laplace and sampler paths both condition on it; integrating it
is the job of the nested-Laplace / EM layer (see the inference vignettes).

## Random effects

A random intercept enters through the `(1 | g)` term. Simulate twelve
groups with their own offsets, then fit conditioning on the random-effect
standard deviation `sigma_re`.

```{r sim-re}
g  <- factor(sample(1:12, n, replace = TRUE))
u  <- rnorm(12, sd = 0.6)
df$y <- 0.5 + 1.2 * df$x + u[g] + rnorm(n, sd = 0.8)
df$g <- g

fit_re <- tulpa(y ~ x + (1 | g), data = df, family = "gaussian",
                mode = "laplace", sigma_re = 0.6, phi = 0.8)
coef(fit_re)
```

`coef()` reports the fixed effects; the group-level effects come from
`ranef()`, one row per group with the posterior mean (BLUP).

```{r ranef}
head(ranef(fit_re), 4)
```

The fit conditions on `sigma_re` rather than estimating it. Passing one
value applies it to the term; with several random-effect terms you pass
one value per term. To integrate the random-effect variance instead of
fixing it, use the nested-Laplace path; for random *slopes*
(`(1 + x | g)`) the engine integrates the covariance automatically.

## Inference tiers

The same model can be fit at different tiers. `mode` selects the backend,
and each backend states what its credible intervals mean. `inference_mode_info()`
prints the full map.

```{r modes, eval = FALSE}
inference_mode_info()
```

The Tier-2 Laplace fit above conditions on a Gaussian posterior shape.
Tier 1 draws from the posterior directly. MALA is a gradient sampler;
Pathfinder is a variational fit that returns draws from a Gaussian
approximation found along an optimisation path. Both are callable from R.

```{r tiers}
df_b <- df
df_b$y <- rbinom(n, 1, plogis(-0.3 + 1.0 * df$x + u[df$g]))

fit_lap <- tulpa(y ~ x + (1 | g), data = df_b, family = "binomial",
                 mode = "laplace", sigma_re = 0.6)
fit_mala <- tulpa(y ~ x + (1 | g), data = df_b, family = "binomial",
                  mode = "mala", sigma_re = 0.6,
                  control = list(n_iter = 1500, warmup = 500))
```

The two tiers agree on the fixed effects to within their uncertainty. The
Laplace fit is deterministic and immediate; the sampler carries Monte
Carlo error and stays agnostic about the posterior shape.

```{r tiers-compare}
data.frame(
  term    = names(coef(fit_lap)),
  laplace = round(coef(fit_lap), 3),
  mala    = round(coef(fit_mala), 3)
)
```

A fit reports which backend ran and at which tier through the object's
`backend` and `inference_tier` fields, and `selection_reason` records why
`mode = "auto"` chose what it did.

```{r tier-fields}
c(backend = fit_mala$backend, tier = fit_mala$inference_tier)
```

The tiers encode an epistemic guarantee, not just a speed trade-off. Tier 1
(exact) draws from the posterior and its credible intervals are interpretable
as posterior uncertainty. Tier 2 (structured) is correct *conditional* on its
structural assumption (a Gaussian posterior shape for Laplace). `mode = "auto"`
picks between Tier 1 and Tier 2 and never silently drops to the optimisation
tier. The R-callable backends:

| Mode | Tier | What it does | Reach for it when |
|---|---|---|---|
| `"laplace"` | 2 | mode + curvature | a fast, deterministic fit; the default sanity check |
| `"pathfinder"` | 2 | variational draws along an optimisation path | you want draws but not full MCMC |
| `"mala"` | 1 | gradient-based Metropolis sampler | exact posterior moments, moderate dimension |
| `"imh_laplace"` | 1 | Laplace proposal with independence-MH correction | debiasing a Laplace fit at low parameter dimension |
| `"gibbs"` | 1 | Polya-Gamma sampler (binomial / negbin) | a conjugate-style exact fit for those families |

`inference_mode_info()` lists every backend, including the ones reachable only
from model packages through the C interface.

## Prediction

`predict()` gives the fixed-effect prediction at new covariate values.
Random effects are held at their population mean of zero, so this is the
marginal trend. `se.fit = TRUE` adds the link-scale standard error and
credible bounds from the fixed-effect covariance.

```{r predict}
nd <- data.frame(x = seq(-2, 2, length.out = 50))
pr <- predict(fit_re, newdata = nd, se.fit = TRUE)
head(pr, 3)
```

Plot the trend with its credible band.

```{r predict-plot, fig.alt = "Predicted response across x with a 95% credible band"}
plot(nd$x, pr$fit, type = "l", lwd = 2, xlab = "x", ylab = "predicted y")
polygon(c(nd$x, rev(nd$x)), c(pr$lower, rev(pr$upper)),
        col = adjustcolor("steelblue", 0.25), border = NA)
lines(nd$x, pr$fit, lwd = 2)
```

For a non-Gaussian family, `type = "response"` maps the prediction
through the inverse link, and the credible bounds stay inside the
response range (probabilities in `[0, 1]` for `binomial`, positive means
for `poisson`).

```{r predict-response}
pr_p <- predict(fit_lap, newdata = nd, type = "response", se.fit = TRUE)
range(pr_p$fit)
```

## Comparing models

`compare_models()` ranks fits by a shared criterion. With
`criterion = "loglik"` it reports `logLik()` for each model, which for a
Laplace fit is the approximate log *marginal* likelihood, the model
evidence, the right quantity for comparing nested specifications.

```{r compare}
m0 <- tulpa(y ~ 1,     data = df, family = "gaussian", mode = "laplace", phi = 0.8)
m1 <- tulpa(y ~ x,     data = df, family = "gaussian", mode = "laplace", phi = 0.8)
m2 <- tulpa(y ~ x + (1 | g), data = df, family = "gaussian",
            mode = "laplace", sigma_re = 0.6, phi = 0.8)
compare_models(intercept = m0, slope = m1, slope_re = m2, criterion = "loglik")
```

Adding the slope lifts the evidence sharply; adding the random intercept
lifts it further, matching how the data were generated. WAIC and LOO are
available for fits that carry a pointwise log-likelihood, which the model
packages provide; the engine's own comparison runs on the marginal
likelihood and the joint log-density.

## Other families

The same call shape covers counts, proportions, and binary data; only
`family` (and any dispersion `phi`) changes.

```{r families}
# Poisson counts
dp <- data.frame(y = rpois(n, exp(0.2 + 0.6 * x)), x = x)
coef(tulpa(y ~ x, data = dp, family = "poisson", mode = "laplace"))

# Beta proportions in (0, 1), precision phi
mu <- plogis(0.1 + 0.8 * x)
db <- data.frame(y = rbeta(n, mu * 8, (1 - mu) * 8), x = x)
coef(tulpa(y ~ x, data = db, family = "beta", mode = "laplace", phi = 8))
```

## Prior predictive checks

Before fitting, it helps to see what the priors imply about the data.
`prior_predict()` draws parameters from their priors, pushes them through
the linear predictor, and simulates responses. The simulator comes from a
`tulpa_family` object; model packages supply rich ones, and a one-line
Poisson simulator is enough here.

```{r prior-pred}
fam <- tulpa_family(
  "poisson",
  function(eta, params, n_obs, ...) rpois(n_obs, exp(eta[[1]]))
)
pp <- prior_predict(y ~ x, family = fam, data = dp, n_draws = 50,
                    priors = tulpa_priors())
pp
```

`pp$y` is a list with one simulated dataset per prior draw, and
`plot(pp)` overlays their densities. When the prior predictive covers
ranges the response could never take, the priors are too vague; the priors
vignette tightens them with `prior_pc()`, `prior_normal()`, and friends.

## A spatial field

Areal data carry a neighbourhood structure. Pass the adjacency through
`spatial=` and name the per-observation unit with a `spatial(col)` term;
`type = "icar"` puts an intrinsic conditional-autoregressive field on the
units. Here twenty regions sit on a ring, each adjacent to its two
neighbours.

```{r spatial-build}
K <- 20
W <- matrix(0, K, K)
for (i in 1:K) { j <- if (i < K) i + 1 else 1; W[i, j] <- W[j, i] <- 1 }

region <- factor(sample(1:K, n, replace = TRUE))
field  <- as.numeric(scale(sin(2 * pi * (1:K) / K)))[region]
ds <- data.frame(y = rbinom(n, 1, plogis(-0.2 + 0.7 * x + field)),
                 x = x, region = region)
```

```{r spatial-fit}
fit_sp <- tulpa(y ~ x + spatial(region), data = ds, family = "binomial",
                spatial = list(type = "icar", adjacency = W),
                mode = "laplace")
coef(fit_sp)
```

The slope recovers while the ICAR field absorbs the regional structure.
Continuous fields use a coordinate spec instead of an adjacency:
`spatial_gp(~ lon + lat)` for an NNGP field, `spatial_hsgp(~ lon + lat)`
for a Hilbert-space GP, `spatial_spde(~ lon + lat, data)` for a Matern
SPDE. The dedicated spatial vignette walks through each.

## User-defined latent blocks

When the structure you need is not a built-in field, `tgmrf()` lets a
script define its own Gaussian Markov random field (a precision matrix
and a hyperprior) and plug it in as a `latent(...)` term that reaches
every inference tier. The `tgmrf()` vignette builds a periodic AR(1) block
end to end.

## Extracting results

The fitted object answers the standard accessors. They work the same way
on a Laplace fit and a sampler fit; the Laplace tier reports the Gaussian
approximation, the sampler tier reports empirical posterior summaries.

| Accessor | Returns |
|---|---|
| `coef()` | named fixed-effect estimates |
| `summary()` | fixed effects: estimate, SE, credible bounds |
| `confint()` | credible intervals (matrix) |
| `vcov()` | fixed-effect covariance |
| `ranef()` | group-level random effects |
| `tidy()` | one-row-per-term data frame (broom style) |
| `glance()` | one-row model summary |
| `predict()`, `fitted()` | fixed-effect prediction / in-sample mean |
| `logLik()` | log marginal likelihood (Laplace) or mean log-density |
| `plot()` | fixed-effect densities (or trace for sampler fits) |

```{r tidy}
tidy(fit_re)
```

## Practical guidance

A few rules of thumb for everyday use:

- **Start at Tier 2.** `mode = "laplace"` is deterministic and returns in
  well under a second on tens of thousands of rows. Use it for the first
  look and for model comparison by marginal likelihood. Move to a sampler
  (`mode = "mala"`) when you need exact posterior moments or the posterior
  is visibly non-Gaussian.
- **`sigma_re` is conditioned, not estimated, on the fixed-`sigma_re`
  paths.** With a random intercept the fit assumes the value you pass. If
  you do not have one, fit at a couple of values (say 0.3 and 1.0) and
  check whether the fixed effects move; if they do, integrate the variance
  through the nested-Laplace path rather than guessing.
- **Sample size.** A random-effect term needs enough groups to inform its
  variance: with fewer than about 5--8 groups, the group-level estimates
  are dominated by the prior, and a fixed effect for the grouping is often
  the honest choice.
- **Watch convergence on the sampler tiers.** `glance()` reports
  `mean_accept` and the divergence count; an acceptance rate far from
  0.4--0.8, or any divergences, means the chain needs more warmup or a
  smaller step size (`control$epsilon`).
- **Reach for a model package when the likelihood is not built in.** The
  five families here cover the common cases. Zero-inflation, hurdles,
  detection/occupancy, ratio responses, and posterior-predictive checking
  live in the packages built on this engine, not in `tulpa()` itself.

## Where to go next

- Inference details and the tier guarantees: the inference-modes and
  algorithm vignettes.
- Spatial and temporal fields, spatially varying coefficients: the
  spatial, temporal, and SVC vignettes.
- Priors and prior predictive checks: the priors vignette and
  `prior_predict()`.
- Posterior-predictive checks, WAIC, and LOO on a full observation model:
  the model packages (tulpaObs, tulpaRatio) that build on this engine.
```
