---
title: "Reliability of nested approximations: reading the outer Pareto-k"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reliability of nested approximations: reading the outer Pareto-k}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260619)
```

```{r load, message = FALSE}
library(tulpa)
```

## What the diagnostic is for

A nested-Laplace fit answers a hard integral with two approximations stacked
on top of each other. The inner one is a Laplace approximation: at each value
of the hyperparameters it replaces the latent posterior with a Gaussian at its
mode. The outer one is an integration: it sweeps the hyperparameters on a grid,
weights each cell by its marginal likelihood, and sums. The fit you read back
is the result of both. Either one can be wrong, and when the answer is off the
estimates and intervals still come back looking entirely ordinary.

A sampler has a standard answer to this problem. Run several chains, and Rhat
tells you whether they agree and the effective sample size tells you how much
independent information you bought. Those numbers are not the posterior; they
are a verdict on whether the machinery that produced it converged. A nested
approximation draws no chains, so Rhat is vacuous for it, but it needs the same
kind of verdict. The outer Pareto-k-hat is that verdict. It is the accuracy
gate for a deterministic fit, the counterpart to what Rhat and ESS are for a
sampled one.

The one thing to fix before anything else is what the number actually checks.
The outer k-hat scores the **outer integration**: whether the Gaussian grid the
integrator places over the hyperparameters faithfully covers the hyperparameter
posterior it is integrating. It does not see the inner Laplace. A fit can carry
a clean k-hat and still be biased if the inner Gaussian was a poor fit to a
skewed latent conditional, because the k-hat was never measuring that. Read
`pareto_k < 0.5` as "the integration is reliable", never as "the posterior is
correct". The distinction decides how you act on the number, and the last
section comes back to it.

## A fit to diagnose

The diagnostic lives on the joint nested-Laplace backend, the path for models
whose arms share a latent field and integrate one hyperparameter posterior
between them. Build a two-arm example: a presence arm with a binomial
likelihood and a positive-value arm with a Gaussian one, both loading on a
shared smooth field over a chain of spatial units.

```{r build}
S     <- 25L
nb    <- lapply(seq_len(S), function(s) setdiff(c(s - 1L, s + 1L), c(0L, S + 1L)))
nn    <- lengths(nb)
field <- as.numeric(scale(cumsum(rnorm(S, 0, 0.4))))

mk_arm <- function(m, fam) {
  si  <- sample(S, m, replace = TRUE)
  x   <- rnorm(m)
  lin <- 0.2 + 0.5 * x + 0.8 * field[si]
  y   <- if (fam == "binomial") rbinom(m, 1L, plogis(lin)) else lin + rnorm(m, 0, 0.5)
  list(y = as.numeric(y), n_trials = rep(1L, m), X = cbind(1, x),
       spatial_idx = si, family = fam, phi = if (fam == "gaussian") 0.5 else 1)
}

prior <- list(type = "icar", n_spatial_units = S,
              adj_row_ptr = c(0L, cumsum(nn)), adj_col_idx = unlist(nb) - 1L,
              n_neighbors = nn, sigma_grid = c(0.3, 0.7, 1.4))
```

Fit the two arms together. The diagnostic runs by default, so the k-hat comes
back on the fitted object with no extra request.

```{r fit}
fit <- tulpa_nested_laplace_joint(
  responses = list(occ = mk_arm(200L, "binomial"),
                   pos = mk_arm(200L, "gaussian")),
  prior     = prior,
  control   = list(diagnose_draws = 400L))

c(pareto_k = round(fit$pareto_k, 3),
  is_ess   = round(fit$pareto_k_is_ess, 1))
fit$pareto_k_scope
```

The scope string names what was scored: the Gaussian proposal the integrator
placed over the joint hyperparameter posterior. The k-hat reads below the
boundaries discussed next, so the integration over the shared field amplitude
and the per-arm scales is reliable, and the importance-sampling effective
sample size confirms the proposal covered the target without a few draws
carrying all the weight.

## What the number means

The k-hat is the shape parameter of a generalized-Pareto distribution fitted to
the upper tail of the importance weights, and it estimates how many moments
that weight distribution has. The reading splits into three bands.

- **Good, `k < 0.5`.** The weights have finite variance. A central limit
  theorem applies, the importance-sampling estimate converges at the usual
  rate, and the integration can be trusted.
- **Ok, `0.5 <= k < 0.7`.** The weights have a finite mean but infinite
  variance. The estimate still converges, slower, and the smoothing keeps it
  usable. The result is acceptable but not as sharp as the good band.
- **Unreliable, `k >= 0.7`.** The proposal cannot be corrected to the target
  by importance weighting. A skewed or heavy-tailed hyperparameter posterior
  is being asked to fit inside a Gaussian grid that does not cover it, and a
  handful of draws dominate the sum.

The two boundaries are configurable through `control$k_conf_bands`, defaulting
to `c(0.5, 0.7)`, and the band the point estimate falls in is the first thing
to read.

```{r band}
band <- cut(fit$pareto_k, c(-Inf, 0.5, 0.7, Inf),
            labels = c("good", "ok", "unreliable"))
band
```

A negative k-hat, as in this fit, sits in the good band with room to spare: the
weights have an even lighter tail than finite variance requires, which is the
easy case a well-resolved grid over a near-Gaussian hyperparameter posterior
produces.

## The boundary moves with the draw count

The `0.7` upper boundary is not a universal constant. The reliability of an
importance-sampling estimate at a given k-hat depends on how many draws were
used to form it. The draw count sets how slowly a heavy-tailed weight
distribution converges, so the boundary below which the estimate can still be
trusted tightens when there are few draws and relaxes toward `0.7` only when
there are many. The usable boundary is `min(1 - 1/log10(S), 0.7)`, with `S` the
number of importance draws (Vehtari, Simpson, Gelman, Yao and Gabry, 2024). The
diagnostic draws `S = diagnose_draws` points, each one independent and drawn from
a single Gaussian proposal, so the nominal draw count is the right `S` and the
rule applies directly.

```{r threshold}
k_usable <- function(S) pmin(1 - 1 / log10(S), 0.7)
S_grid   <- c(100, 200, 400, 1000, 2154, 4000)
data.frame(draws = S_grid, usable_boundary = round(k_usable(S_grid), 3))
```

A few hundred draws place the boundary nearer `0.6` than `0.7`, and it reaches
the familiar `0.7` only past roughly two thousand draws. The engine applies this
by default: `pareto_k_band_confident` tests the bootstrap interval against
`c(0.5, min(1 - 1/log10(S), 0.7))` at the realised draw count, so the good cut
stays at `0.5` and the usable cut tracks `S`. A k-hat that looks acceptable at
`0.65` from four hundred draws is already past the usable boundary for that draw
count, and the flag reflects it. Pass `control$k_conf_bands` a fixed vector
(e.g. `c(0.5, 0.7)`) to override the size-dependent default. The practical
consequence is conservative either way: a fit near the boundary at a modest draw
count deserves more draws before it is trusted, which is what the next section
does.

## One k-hat is a noisy reading

The k-hat is a single number, but it is an estimate: a generalized-Pareto shape
fitted to a few hundred tail weights carries real sampling error. The true outer
k-hat is fixed for a given fit and proposal; what a finite draw count leaves
uncertain is the estimate of it, not the quantity. Were you to redraw the
importance sample the k-hat would move, and acting on which side of `0.5` a
single estimate happens to land is reading noise.

The fit reports the spread alongside the point. By default it bootstraps the
diagnostic's own importance ratios and returns a confidence interval for the
k-hat together with a flag for whether that interval stays inside a single
reliability band.

```{r bootstrap}
c(k       = round(fit$pareto_k, 3),
  ci_low  = round(fit$pareto_k_ci_low, 3),
  ci_high = round(fit$pareto_k_ci_high, 3))
fit$pareto_k_band_confident
```

`pareto_k_band_confident` is the number to act on, not the point estimate. It
is `TRUE` only when the whole interval sits in one band, so it answers the
question you actually have: not "what is the k-hat" but "do I know which band
this fit is in". A point estimate of `0.45` with an interval spanning `0.3` to
`0.6` is not confidently good; it might be merely ok, and the flag says so.

The bootstrap cannot invent tail information. It measures how unstable the
current tail estimate is, given the draws in hand. To make the interval itself
tighter you need more actual tail ratios, which means raising `diagnose_draws`,
not the bootstrap count.

## Tightening a borderline reading

When the band is not confident the interval straddles a boundary, and the fix is
more tail information. The precision knob is `diagnose_draws`: each draw is an
off-grid Laplace re-solve, and more of them give the generalized-Pareto fit more
tail ratios, a tighter k-hat, and a narrower bootstrap interval. Refit the same
model with more draws and read the interval again.

```{r more-draws}
fit_hi <- tulpa_nested_laplace_joint(
  responses = list(occ = mk_arm(200L, "binomial"),
                   pos = mk_arm(200L, "gaussian")),
  prior     = prior,
  control   = list(diagnose_draws = 1500L))

c(k              = round(fit_hi$pareto_k, 3),
  ci_low         = round(fit_hi$pareto_k_ci_low, 3),
  ci_high        = round(fit_hi$pareto_k_ci_high, 3),
  band_confident = fit_hi$pareto_k_band_confident)
```

Raising `diagnose_draws` is the lever for both halves of the reading at once. It
tightens the interval around the estimate, and because the usable boundary
itself relaxes toward `0.7` as the draw count grows, it widens the band the
estimate has to clear. Raising `k_bootstrap` does neither: it re-uses the tail
ratios already in hand, so it sharpens the description of the current estimate
without adding tail information. When more draws still leave the interval across
a boundary, the fit is genuinely on the line, and the next rung of the ladder is
the answer.

## The reliability ladder

The bands, the band-confidence flag, and the draw-count lever are rungs on a
single ladder. Each rung either certifies the fit or hands you the reason to
climb to the next.

1. **Report.** Read `pareto_k` and `pareto_k_band_confident`. A confident good
   reading ends here: the integration is reliable and nothing more is needed.
2. **Resolve the band.** When the band is not confident, raise `diagnose_draws`
   to add tail information until the bootstrap interval clears a boundary, or
   until more draws stop moving it and the fit is genuinely on the line. This
   rung pins down which band you are in; it does not change the fit.
3. **Refine the integration.** When the band resolves to ok or unreliable
   because the hyperparameter posterior is skewed or multi-modal within the
   grid, the fix is to refine the integration so the proposal covers it, then
   re-diagnose. This rung changes the fit. It is described in the roadmap below.
4. **Debias.** When refinement cannot reach the requested band, or when the
   suspected error is the inner Laplace the outer k-hat cannot see, escalate to
   an exact debias: a Gibbs or gradient-sampler correction that replaces the
   deterministic approximation with sampling on the directions where it was
   biased.

The ladder is the engine's design philosophy made operational. A cheap
deterministic approximation carries the body of the work, a diagnostic says
where it is trustworthy, and exact sampling is reserved for the residual
directions where the approximation falls short. The k-hat is the gate between
the rungs. The companion escalation for a random-effect covariance fit is shown
in the inference-modes vignette, where a k-hat past the boundary routes the fit
to `control = list(re_cov = "gibbs")` and an exact inverse-Wishart draw.

## Where this is heading

The ladder is currently driven by reading the flag and choosing the next rung
by hand. A forthcoming front door collapses it to a single statement of intent.
You name the reliability you require, and the engine climbs the ladder for you,
stopping when the validated k-hat interval is confidently inside the band you
asked for or reporting plainly that it could not get there.

```{r roadmap, eval = FALSE}
fit <- tulpa_nested_laplace_joint(
  responses = arms, prior = prior,
  control = list(
    k_quality = "good",      # require a confident good band, not a lucky point
    k_refine  = "mixture"    # refine the integration when the band falls short
  ))

fit$k_quality_requested      # "good"
fit$k_quality_reached        # TRUE or FALSE, never silently downgraded
fit$k_quality_best           # the band actually achieved, e.g. "ok"
fit$k_quality_reason         # why it stopped: resolved, cap, or budget
```

The contract is that the request is a target, not a promise. `k_quality =
"good"` means accept a fit only when the k-hat interval is confidently below
the good boundary; `k_quality = "ok"` lowers the bar to the usable boundary.
When the engine reaches the requested band it says so, and when it cannot it
returns the best band it did reach with the reason it stopped, rather than
handing back a fit that quietly fell short. The same honesty the band-confidence
flag already enforces, applied to the whole loop.

## See also

- `tulpa_nested_laplace_joint()` documents the full `control` surface for the
  diagnostic: `diagnose_k`, `diagnose_draws`, `k_bootstrap`, `k_tail_points`,
  `k_conf_bands`, and `k_threads`.
- `tulpa_psis()` is the native Pareto-smoothed importance-sampling core the
  diagnostic is built on, callable directly on a vector of log importance
  ratios.
- `mcmc_diagnostics()` is the chain-fit counterpart, the Rhat and ESS surface
  for sampled fits.
- The inference-modes vignette runs the reliability ladder from a Laplace fit
  through an exact debias, and shows the k-hat on a random-effect covariance
  fit.
```
