---
title: "Validating calibration: simulation-based calibration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating calibration: simulation-based calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7, fig.height = 5
)
set.seed(20260828)
```

```{r cran-gate, include = FALSE}
# The fits below are evaluated when the article is built locally, in CI and for
# the pkgdown site (all of which set NOT_CRAN). Each experiment here runs a few
# hundred fits, which does not fit inside CRAN's ten-minute budget for the whole
# check, so there the code is shown without being run.
EVAL_FITS <- identical(Sys.getenv("NOT_CRAN"), "true")
knitr::opts_chunk$set(eval = EVAL_FITS)
```

```{r load, message = FALSE}
library(tulpa)
```

## What the instrument reads

`diagnostics()` scores one fit. It asks whether the machinery that produced
that posterior behaved: whether the chains mixed, whether the outer grid
covered the hyperparameter posterior, whether the inner Gaussian was a
reasonable shape for the latent conditional. Those are verdicts on the
procedure at one data set.

Calibration is a different question, and no single fit can answer it. An
inference algorithm is calibrated when its posteriors are self-consistent
across the whole generative model: simulate a truth, simulate data at that
truth, fit, and the truth should land at a uniformly distributed position
inside the posterior it produced. Repeat that many times and those positions
have a distribution you can test.

The usual way to check this is coverage: count how often a 95% interval
contains the truth. That reads one point of the marginal CDF, or two if you
also run a 50% interval. It cannot say whether a posterior is biased,
over-dispersed, under-dispersed or asymmetric, and it is weak enough that two
genuinely different reads of the same fit can score identically on it.
`sbc()` reads the whole CDF.

The measurement is the probability integral transform. For a scored quantity
whose true value is `theta_0` and whose reported posterior CDF is `F`, take
`u = F(theta_0)`. Under exact inference `u` is exactly Uniform(0, 1), so the
entire empirical CDF of the `u` values across simulations is the test
statistic (Talts et al. 2018).

## A model to validate

A balanced Gaussian random-intercept model: six regions of four observations
each, an intercept and one covariate, a random intercept per region whose
standard deviation `sigma` is drawn from a seven-point grid.

The design is small on purpose. The `sigma` posterior then spreads across four
or five of the seven grid cells, which is the regime where the outer-grid
mixture posterior for a fixed effect is farthest from the Gaussian matching
its first two moments.

```{r fixture}
GRID <- exp(seq(log(0.2), log(1.5), length.out = 7))
PHI  <- 0.7
BETA <- c(-0.2, 0.7)

simulate_one <- function(seed) {
  set.seed(seed)
  sigma  <- GRID[sample.int(length(GRID), 1L)]
  region <- rep(seq_len(6L), each = 4L)
  X      <- cbind(1, rnorm(24L))
  u      <- rnorm(6L, 0, sigma)
  list(y      = as.numeric(X %*% BETA) + u[region] + rnorm(24L, 0, PHI),
       X      = X,
       region = region,
       theta  = c(beta1 = BETA[1], beta2 = BETA[2], sigma = sigma))
}

fit_one <- function(d, diagnose = FALSE) {
  tulpa_nested_laplace(
    y = d$y, n_trials = rep(1L, length(d$y)), X = d$X,
    prior = list(list(type = "iid", obs_idx = d$region,
                      n_units = max(d$region), sigma_grid = GRID)),
    family = "gaussian", phi = PHI,
    control = list(n_threads = 1L, keep_grid_hessians = TRUE,
                   auto_recenter = FALSE, progress = FALSE,
                   diagnose_k = diagnose, diagnose_skew = diagnose))
}
```

`simulate_one()` has to be a pure function of its seed, and it returns `theta`,
the named vector of true values the PIT is taken against. `auto_recenter =
FALSE` is not a speed knob: the argument that the `sigma` PIT is uniform needs
the fitted grid to be the same seven points the truth was drawn from, and a
recentred grid breaks it.

The single-fit reliability diagnostics are off inside the loop. Each one costs
its own batch of inner solves per fit, and the last section is where they get
read, on one fit rather than two hundred.

## Reporting a fit in a shape `sbc()` can read

A fitter callback returns a named list of **arms**, each a named list over
scored quantities, each entry one of the predictive shapes in
`?sbc_predictive`. Everything downstream dispatches on the shape's `kind` tag,
so a backend reporting something new is one shape rather than a parallel
scorer.

Two of them cover a nested-Laplace fit. What the outer grid defines for a
fixed effect is a Gaussian mixture over the cells, and
`tulpa_posterior_draws()` realizes it: each draw picks a cell by its weight,
then samples that cell's inner Gaussian. `coef()` and `vcov()` report that
same mixture's first two moments, so building an `sbc_normal()` from them is
the collapsed read of the identical posterior. For the hyperparameter the grid
defines a distribution on a finite support, which is `sbc_discrete()`.

```{r arms}
arms <- function(d) {
  fit <- fit_one(d)
  m   <- coef(fit)
  se  <- sqrt(diag(vcov(fit)))
  D   <- tulpa_posterior_draws(fit, n = 2000)
  w   <- fit$weights / sum(fit$weights)

  list(
    mixture = list(
      beta1 = sbc_draws(D[, 1]),
      beta2 = sbc_draws(D[, 2]),
      sigma = sbc_discrete(as.numeric(fit$theta_grid), w)),
    collapsed = list(
      beta1 = sbc_normal(m[1], se[1]),
      beta2 = sbc_normal(m[2], se[2])),
    narrow = list(
      beta1 = sbc_normal(m[1], se[1] / 1.25),
      beta2 = sbc_normal(m[2], se[2] / 1.25)))
}
```

Every arm reads off **one solve per seed**, so an arm-to-arm difference
carries no fit-to-fit noise. `narrow` is a deliberately mis-scaled posterior,
the same moments with the standard deviation divided by 1.25. A calibration
harness that cannot fail is worth nothing, so the run below has to put it
outside the band.

## The prior-predictive experiment

```{r prior-predictive}
res <- sbc("prior_predictive",
           simulator  = simulate_one,
           fitter     = arms,
           n_sim      = 200L,
           flat_prior = c("beta1", "beta2"))
res
```

`flat_prior` is a guard doing its job. Ordinary SBC draws the truth from the
prior, so an improper prior has nothing to draw from, and the nested-Laplace
door puts no prior on the fixed effects. `sbc()` probes the simulator, sees
that `beta1` and `beta2` do not move across simulations, and refuses to score
them unless the caller names them. Naming them asserts that their flat prior
leaves the PIT uniform by a structural argument; the assertion is checked in
both directions and travels on the result in `res$premises`.

## Three reads of the same PIT sample

`ks` and `p_unif` read the raw PIT ECDF. `inside_folded` and `p_folded` read
`2 |u - 1/2|`, which is also uniform under correct inference. The fold is
where a symmetric error shows: a posterior that is too narrow pushes PIT mass
towards both ends at once, which cancels in the raw ECDF and accumulates in
the folded one. `crps` is the continuous ranked probability score, closed form
for a Gaussian mixture, so the nested tier's own posterior is scored with no
Monte Carlo.

`inside` is the verdict against a **simultaneous** band. A pointwise binomial
band is not one: at n = 100, holding each order statistic inside its own 95%
interval holds all of them together only 44.71% of the time. The band here is
calibrated by bisection against the exact crossing probability of the uniform
order statistics, so an ECDF excursion anywhere along the curve is a
0.05-level event.

Discrete quantities are randomized within their atom,
`u = F(theta^-) + V P(theta)`, so `sigma` on its seven-point grid and a
continuous coefficient share one uniform reference and one band. Reading a
rank against a continuous uniform is the classic silent SBC bug, and it is not
something you can opt into here.

## The proper score, paired

`summary(baseline = )` pairs the CRPS seed by seed against one arm.

```{r crps}
summary(res, baseline = "mixture")
```

`delta` with its `t` is the proper-score verdict; a negative `delta` is the
arm scoring better. `p_sign` is more powerful at detecting that two arms
differ at all, but the sign test is not a proper score, so it cannot rank
them.

Expect the mixture and collapsed arms to score close together. They carry the
same first two moments, and the CRPS integrates the whole squared CDF
difference, which those moments dominate. Where they separate is on the ECDF
reads above, which is why both instruments are worth running.

The CRPS is a proper posterior score here only because the truth is drawn
afresh each simulation. Hold the truth fixed across seeds and the
CRPS-optimal forecast is a point mass at it, so a sharper wrong posterior
wins. That is enforced rather than documented: a fixed-truth sweep is not
offered as an experiment, and `summary(baseline = )` refuses to rank one.

## The picture

```{r plot-raw, fig.alt = "PIT ECDF difference from uniform against the simultaneous band"}
plot(res, arm = c("mixture", "narrow"), quantity = "beta2")
```

```{r plot-folded, fig.alt = "Folded PIT ECDF difference from uniform against the simultaneous band"}
plot(res, arm = c("mixture", "narrow"), quantity = "beta2", folded = TRUE)
```

The plot draws the ECDF **difference** from uniform, so a calibrated read is a
flat line at zero inside the band. The under-dispersed arm bows away from it,
and the bow is larger folded than raw.

## Calibration conditional on an observed data set

The experiment above averages over the prior. A user fitting their own data
asks something narrower: is the inference reliable in the posterior geometry
*this* data set produces. The prior average can miss a defect confined to a
small region of parameter space, and it can flag one the observed data rules
out.

`experiment = "posterior"` answers the narrow question (Sailynoja et al. 2026,
Algorithm 2). Draw `theta'` from `pi(theta | y_obs)`, simulate a replicate at
`theta'`, and take the PIT under the **augmented** posterior
`pi(theta | y, y_obs)`. That is ordinary SBC with `pi(theta | y_obs)` in the
role of the prior, so the same band, the same folded read and the same proper
score all carry over, and it needs no proper prior at all.

It takes six callbacks.

```{r posterior-model}
d_obs <- simulate_one(99L)

model <- list(
  data_obs = d_obs,

  fit = function(data) fit_one(data),

  draw_theta = function(fit, seed) {
    set.seed(seed)
    b <- tulpa_posterior_draws(fit, n = 1L)
    k <- attr(b, "cells")[1]
    c(beta1 = unname(b[1, 1]), beta2 = unname(b[1, 2]),
      sigma = as.numeric(fit$theta_grid)[k])
  },

  simulate = function(theta, seed) {
    set.seed(seed)
    region <- rep(seq_len(6L), each = 4L)
    X      <- cbind(1, rnorm(24L))
    u      <- rnorm(6L, 0, theta[["sigma"]])
    list(y = as.numeric(X %*% theta[c("beta1", "beta2")]) +
           u[region] + rnorm(24L, 0, PHI),
         X = X, region = region)
  },

  pool = function(obs, rep) list(
    y      = c(obs$y, rep$y),
    X      = rbind(obs$X, rep$X),
    region = as.integer(c(obs$region, rep$region + max(obs$region)))),

  arms = function(fit, data) {
    m  <- coef(fit)
    se <- sqrt(diag(vcov(fit)))
    D  <- tulpa_posterior_draws(fit, n = 2000)
    list(
      mixture = list(
        beta1 = sbc_draws(D[, 1]),
        beta2 = sbc_draws(D[, 2]),
        sigma = sbc_discrete(as.numeric(fit$theta_grid),
                             fit$weights / sum(fit$weights))),
      narrow = list(
        beta1 = sbc_normal(m[1], se[1] / 1.25),
        beta2 = sbc_normal(m[2], se[2] / 1.25)))
  },

  group_ids = function(data) data$region)
```

```{r posterior-run}
pres <- sbc("posterior", model = model, n_sim = 200L)
pres
```

The driver hands `draw_theta` and `simulate` different seeds, so
`set.seed(seed)` at the top of each is the correct fixture. Giving both the
same seed makes the replicate's noise a function of the truth, which is not
`p(y | theta')`, and it shows up as a non-uniform PIT with nothing wrong in
the inference under test.

## Two premises, and what checks them

Each premise silently turns the construction into something that is not SBC,
so each has a guard, and each guard's conclusion travels on the result.

```{r premises}
str(pres$premises)
```

**The augmented posterior conditions on both data sets.** Fitting the
replicate alone is ordinary SBC under a hand-made prior. `sbc()` refuses a
`pool()` returning no more than the replicate, or no more than the observed
data.

**The replicate is conditionally independent of `y_obs` given theta.** The
nested tier integrates the random effects out, so `theta` carries no per-group
value, and a replicate drawn on the *same* regions couples the two data sets
through the group effects theta does not describe. `simulate()` above draws
six fresh regions and `pool()` offsets their labels past the observed ones.
Supply `group_ids` and the observable half of that is verified, the labels
being disjoint. Omit it and the result records the premise as unverified
rather than assumed. The other half, that the replicate's group effects came
from the prior rather than conditionally on `y_obs`, is not visible from
outside the callback and is not claimed.

A third requirement is a property of `draw_theta` rather than a guard.
`theta'` has to be a **joint** draw from `pi(theta | y_obs)`. Sampling `beta`
from `tulpa_posterior_draws()` and `sigma` independently from the grid weights
gives the right two marginals and the wrong joint, and the `sigma` read leaves
the band when you do it. Reading `attr(b, "cells")` takes the coefficient and
the hyperparameter from the same cell, which is what makes the draw joint.

## Where this sits beside the single-fit band

`diagnostics()` on a nested fit reports the outer Pareto-k-hat, the inner
skewness estimate `gamma_3` and the inner importance k-hat, combined into one
reliability band. That band runs on the fit you already have; `sbc()` costs a
few hundred fits. So the band is what you read routinely, and this is what you
reach for when the answer has to hold up.

The cheap one is not a compressed version of the expensive one, and the two
disagree in both directions. Measured over fifteen configurations at 1000
simulations (`dev_notes/issue339/`): a Poisson configuration carrying an outer
k-hat of 0.196 with both inner scores in the `good` band, the cleanest verdict
the band can give, fails calibration at `p = 2.3e-13` on its intercept. A
binomial configuration carrying an outer k-hat of 1.413, well past the 0.7
escalation threshold, passes at `p = 0.17`.

So read the shipped band as a screen. A rejection from `sbc()` is the strong
statement: on that same measurement the false-positive rate is 0.0117 against
a nominal 0.05, and power reaches 80% at roughly a 10% dispersion error or a
0.14-SD location bias, which makes a pass the weaker one.

`diagnostics()` reads both at once when you hand it the calibration result.

```{r combined}
fit <- fit_one(d_obs, diagnose = TRUE)
diagnostics(fit, sbc = res)
```

## See also

- `?sbc` for the full callback contract, the guards, and what each column of
  the report is.
- `?sbc_predictive` for the predictive shapes, and which one a backend should
  report.
- `vignette("reliability-pareto-k")` for the single-fit reliability band this
  screens against: the outer Pareto-k-hat, `gamma_3`, and the combined
  verdict.
- `?tulpa_posterior_draws` for the mixture sampler the fixed-effect arms are
  built on, and `?tulpa_psis` for the Pareto-smoothing core the k-hat uses.

## References

Talts, S., Betancourt, M., Simpson, D., Vehtari, A. and Gelman, A. (2018).
Validating Bayesian inference algorithms with simulation-based calibration.
arXiv:1804.06788.

Sailynoja, T., Schmitt, M., Buerkner, P.-C. and Vehtari, A. (2026). Posterior
SBC: simulation-based calibration checking conditional on data. *Statistics
and Computing* **36**, 78. doi:10.1007/s11222-026-10825-9.

Gneiting, T. and Raftery, A. E. (2007). Strictly proper scoring rules,
prediction, and estimation. *Journal of the American Statistical Association*
**102**, 359-378.
