---
title: "Spatial fields in tulpa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial fields in tulpa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260529)
```

```{r load, message = FALSE}
library(tulpa)
```

## Why a spatial field

Regions that sit next to each other tend to look alike. Two adjacent
counties share weather, soil, road networks, and the people who move
between them, so their disease counts, yields, or survey rates carry a
similar unmeasured signal. A regression that ignores this treats each
region as an independent draw, and that assumption fails in two ways at
once.

The first failure is in the residuals. Neighbouring regions leave
correlated leftovers, the model's effective sample size is smaller than
the row count suggests, and the standard errors come out too small. You
can see this in any map of model residuals: when nearby regions share the
sign of their leftover, an inference procedure that assumes independence
counts each of those regions as fresh information when, statistically,
they are partly the same observation seen twice. Confidence intervals
shrink below their nominal width, and a slope that is really uncertain
gets reported as sharp.

The second failure is in the coefficients themselves. When a covariate is
itself spatially patterned, say elevation or median income, it competes
with the unmeasured regional signal for the same variance. The fitted
slope then absorbs part of that signal and drifts away from the effect you
meant to estimate. This is the spatial face of confounding by an omitted
variable: the covariate and the missing regional driver vary together
across space, and a model with no place to put the regional driver has no
choice but to load it onto the covariate. The bias stays hidden in plain
sight. The slope looks plausible, the fit looks clean, and the number is
wrong.

Both failures share one cause. The model has nowhere to absorb the part of
the response that depends on *where* an observation sits rather than on its
covariates. A spatial field is the place. You add one random effect per
region, tie neighbouring effects together so the field stays smooth, and
let it soak up the regional signal the covariates do not explain. The
slope is then read against a background that already accounts for location.
The standard errors widen to their honest size because the random effect,
not the covariate, now carries the spatial autocorrelation.

The smoothing is the subtle part. A free random effect per region with no
ties between neighbours would absorb the signal, but it would also absorb
the covariate effect you are trying to measure, because with enough free
parameters it can mimic almost anything. The neighbour ties are what keep
the surface smooth and let the covariate keep its own variance. How strong
those ties should be is a parameter in its own right, and a fit that fixes
it by hand is making a choice the data could make better. tulpa builds the
field from an adjacency matrix, integrates the smoothing strength over a
grid rather than fixing it, and reports a standard error that carries the
field's own uncertainty into the fixed effects.

This vignette stays on the areal path, where regions are discrete units
joined by an adjacency graph. We simulate a known field on a ring of
regions, fit it, pull the field back out per region, check that it
matches the truth, look at how the smoothing precision is recovered, and
compare a spatial fit against a non-spatial one. Continuous fields, where
the signal lives on raw coordinates, get a short note at the end with
pointers to their own vignettes.

## The model

Write the response for observation $i$ in region $r_i$ through a linear
predictor on the link scale:

$$
g(\mu_i) \;=\; \beta_0 + x_i^\top \beta + \phi_{r_i},
$$

where $\beta_0$ is the intercept, $\beta$ the covariate slopes, and
$\phi_{r}$ the spatial effect for region $r$. The field $\phi$ carries the
regional signal; everything spatially structured that $x$ misses lands
there.

The intrinsic conditional autoregressive (ICAR) prior ties the field to
the adjacency graph. Each region's effect is centred on the mean of its
neighbours:

$$
\phi_r \mid \phi_{-r} \;\sim\; \mathcal{N}\!\left(
  \frac{1}{n_r}\sum_{s \sim r} \phi_s,\;\; \frac{1}{\tau\, n_r}\right),
$$

with $s \sim r$ meaning $s$ is a neighbour of $r$, $n_r$ the number of
neighbours, and $\tau$ a precision that sets how tightly the field is
smoothed. The conditional form above is convenient for intuition, since it says each
region pulls toward its neighbours' average. The joint distribution it
implies is a Gaussian Markov random field with precision matrix

$$
Q \;=\; \tau\,(D - W),
$$

where $W$ is the adjacency matrix and $D$ the diagonal of neighbour
counts. The off-diagonal entries of $Q$ are nonzero only between
neighbours, which is what makes $Q$ sparse and the model cheap to work
with even on thousands of regions. That sparsity is the Markov property of
the random effect written as a matrix, not a numerical convenience tacked
on afterward. A region is conditionally independent of every
non-neighbour given its neighbours, and that conditional independence is
exactly a zero in $Q$.

The matrix $D - W$ is the graph Laplacian. Its smallest eigenvalue is
zero, with the constant vector as eigenvector, so $Q$ is rank-deficient by
one. The practical consequence is that the field is pinned only up to an
additive constant: shifting every $\phi_r$ by the same amount does not
change the joint density. The overall level is left for the intercept to
identify, and the field carries only the *contrasts* between regions, the
part that actually describes spatial pattern. This is what makes the ICAR
an intrinsic prior, improper on its own but proper once the data and a
sum-to-zero handling fix the level.

The precision $\tau$ sets the smoothing strength. A large $\tau$ inflates
the penalty on neighbour differences, forces neighbours to agree, and
flattens the field toward the constant. A small $\tau$ relaxes the
penalty and lets each region drift toward an independent random effect.
The value that fits the data is rarely known in advance, and choosing it
wrong has a real cost: too much smoothing erases genuine pattern, too
little lets it chase noise and steal variance from the covariates.
Integrating $\tau$ rather than fixing it is the point of the nested path
used below.

BYM2 splits the field into a spatially structured ICAR part and an
unstructured per-region part:

$$
\phi_r \;=\; \frac{1}{\sqrt{\tau}}\left(\sqrt{\lambda}\,u_r^{\star}
  + \sqrt{1-\lambda}\,v_r\right),
$$

where $u^{\star}$ is a scaled ICAR field, $v$ an IID per-region effect,
and $\lambda \in [0,1]$ a mixing parameter for the share of variance that
is spatially structured. As $\lambda \to 1$ the field is pure ICAR; as
$\lambda \to 0$ it collapses to independent noise. The ICAR component is
scaled following Riebler et al. (2016) so that $\lambda$ reads on a
comparable scale from one graph to the next, which a raw ICAR variance
does not. Both `icar` and `bym2` route through the same machinery in
tulpa; the choice between them is about interpretation, and it comes back
in the practical guidance below.

A proper CAR sits between the two. It keeps the single field of the ICAR
but replaces the fixed unit autocorrelation with an estimated parameter
$\rho$, so the precision becomes $Q = \tau\,(D - \rho W)$. With
$\rho$ free in its valid range the field can range from near-independence
at small $\rho$ to near-ICAR as $\rho$ approaches one, and the precision
is proper rather than rank-deficient, which incidentally tames the
intercept confounding. The cost is one more parameter to integrate and a
field that no longer reduces to the clean graph Laplacian. For most
smoothing tasks ICAR or BYM2 is the practical starting point, with proper
CAR held in reserve for when the autocorrelation strength is itself a
quantity of interest.

## Simulating data

A simulation with a known field is the only honest way to check a spatial
fit. Real data never reveal the truth, so a recovered field that merely
looks reasonable proves nothing. Here we set the field by hand, generate
the response from it, and ask the fit to find it again. The test has
teeth: the response is binary, the data are spread thin across many
regions, and the field carries structure no intercept can fake.

Put twenty-five regions on a ring, each adjacent to its two neighbours.
The ring is the smallest graph that is connected, regular, and has an
obvious smooth field on it. A regular graph keeps every region with the
same neighbour count, so no region is privileged by the topology, and a
ring's single loop gives the field one clean wavelength to recover.

```{r adjacency}
K <- 25
W <- matrix(0, K, K)
for (i in 1:K) {
  j <- if (i < K) i + 1 else 1
  W[i, j] <- W[j, i] <- 1
}
```

The true field is one smooth wave around the ring, centred and scaled so
its values run from roughly $-1.5$ to $1.5$. A wave is a fair test on two
counts. It is smooth, so an ICAR prior that rewards agreement between
neighbours should track it well rather than fight it. And it has real
structure, a peak on one side of the ring and a trough on the other, that
no single intercept can capture, so the random effect has to do genuine
work. A pattern of pure noise would be the opposite test, one an ICAR
prior is designed to smooth away, and recovering it would say little.

```{r truth}
field_true <- as.numeric(scale(sin(2 * pi * (1:K) / K)))
beta0 <- -0.2
beta1 <-  0.8
```

Draw `n` observations, assign each to a region, and build a binomial
response on the logit scale. The intercept is $-0.2$, the covariate slope
$0.8$, and the region's field value adds to the linear predictor.

```{r simulate}
n      <- 800
x      <- rnorm(n)
region <- sample(1:K, n, replace = TRUE)
eta    <- beta0 + beta1 * x + field_true[region]
y      <- rbinom(n, size = 1, prob = plogis(eta))
ds     <- data.frame(y = y, x = x, region = factor(region))
```

The slope we want back is `beta1`; the field we want back is
`field_true`. Eight hundred observations spread across twenty-five
regions average to thirty-two rows each, but the sampling is random, so
several regions land with far fewer. Those thin regions are where
neighbour borrowing earns its keep: a region with a handful of binary
trials cannot estimate its own effect, yet its neighbours can lend it
theirs, and the smoothing is the channel that lending flows through.

## Fitting

An areal field needs two pieces that travel together. The `spatial=`
argument carries the structure, a list with the field `type` and the
adjacency matrix. A `spatial(region)` term in the formula names the column
that maps each row to its region. Drop either piece and `tulpa()` stops
with a message rather than guessing.

```{r fit}
fit <- tulpa(y ~ x + spatial(region), data = ds, family = "binomial",
             spatial = list(type = "icar", adjacency = W),
             mode = "nested_laplace")
fit$backend
```

`mode = "nested_laplace"` runs the Tier-2 path that integrates the field's
precision $\tau$ over a grid rather than fixing it. The path lays a grid
of $\tau$ values, runs an inner Laplace solve of the random effect and
fixed coefficients at every value, weights each grid point by its marginal
likelihood, and pools the answers. This integration turns a single guessed
smoothing strength into a posterior over smoothing strengths, which is why
the coefficient errors below inherit the field's uncertainty instead of
pretending $\tau$ is known exactly.

For a binomial areal field, `mode = "auto"` makes a different choice. It
picks the Polya-Gamma Gibbs sampler, a Tier-1 exact fit that draws from
the posterior directly and is the natural match for a binomial response.
That sampler is correct and often what you want, but it does not expose
the integration grid, so the per-region effects and the smoothing
hyperparameter are not laid out on the returned object the way the nested
path arranges them. This vignette requests `nested_laplace` explicitly so
the marginalised field and $\tau$ summary are available to plot and check.
For a routine binomial areal model where only the slopes matter,
`mode = "auto"` is the lighter call.

The fixed effects come back from `coef()`.

```{r coef}
coef(fit)
```

The slope lands near the simulated $0.8$. With the regional wave absorbed
into the field, the covariate reports its own effect rather than a blend
of effect and location. `summary()` adds the posterior standard
error and the 2.5% / 97.5% credible bounds.

These standard errors are grid-marginalised, and the distinction matters.
At a single fixed $\tau$ the Laplace solve gives a within-point covariance,
the uncertainty in the slope conditional on that smoothing strength. Such
a conditional error understates the truth because it behaves as though
$\tau$ were known. The law of total variance repairs this: the marginal
variance of the slope equals the average within-point variance plus the
spread of the within-point means across the grid. tulpa retains the
per-grid covariance (`control$keep_grid_hessians` defaults to on for
precisely this reason) and combines both terms, so the reported error
absorbs the cost of not knowing how smooth the surface is.

```{r summary}
summary(fit)
```

### Extracting the field per region

The slopes are only half the story. Often the field itself is the
quantity of interest, the smoothed map you wanted in the first place, and
the nested fit carries it on the object. `$modes` is an `n_grid` by
`n_latent` matrix. Each row is one grid point's inner-Laplace mode, and
the columns run `[fixed effects, field]`, so the first `n_fixed` columns
are the intercept and slopes and the rest are the per-region effects.
`$weights` are the integration weights over the grid, one per row, and
they sum to one after normalising.

The grid-marginalised value for region $u$ is the weighted average of its
own column across grid points: take column `n_fixed + u`, weight each
grid point's entry by its integration weight, and sum. This is the same
marginalisation the slopes get, applied to the latent field instead of
the fixed effects.

```{r field-extract}
w     <- fit$weights / sum(fit$weights)
nf    <- fit$n_fixed
field_hat <- vapply(seq_len(K), function(u) sum(w * fit$modes[, nf + u]),
                    numeric(1))
round(head(field_hat), 3)
```

That recipe is the same one the temporal path uses for a random-walk
field, since an RW1 chain is an ICAR on a line: time $t$ neighbours $t-1$
and $t+1$, the chain adjacency's Laplacian is the RW1 precision, and the
field-extraction step reads the same `$modes` columns. One marginalisation
serves both areal maps and time series.

## Checking the fit

Recovery means two things here: the field traces the truth, and the
smoothing precision sits where the data put it.

Plot the fitted values against the simulated ones. The intrinsic prior
identifies the field only up to an additive constant, so centre both
sequences before comparing.

```{r field-plot, fig.alt = "Estimated spatial field per region against the simulated truth, near the identity line"}
plot(field_true - mean(field_true),
     field_hat  - mean(field_hat),
     xlab = "true field (centred)", ylab = "estimated field (centred)",
     pch = 19, col = "steelblue")
abline(0, 1, lwd = 2, col = "grey40")
```

```{r field-cor}
cor(field_true, field_hat)
```

The points fall along the identity line and the correlation is high. The
fit has recovered the wave from binary data spread thin across
twenty-five regions, which is exactly what neighbour borrowing buys. A
region with only a few trials does not stand alone; its estimate is pulled
toward the smooth curve its neighbours trace, and the binary noise that
would swamp an isolated estimate averages out along the curve. The
centring matters because the intrinsic prior fixes the effect only up to a
constant, so the truth and the estimate can differ by an overall level
even when their shapes match perfectly. Subtracting each mean removes that
free constant and leaves the contrasts, which are what the model actually
identifies.

The precision $\tau$ is summarised on the object. `$theta_mean` is the
posterior mean over the grid, on the natural precision scale, and
`$theta_ci_lo` / `$theta_ci_hi` give a credible interval.

```{r tau}
c(tau_mean = fit$theta_mean,
  lo = fit$theta_ci_lo, hi = fit$theta_ci_hi)
```

A larger $\tau$ would mean a stiffer penalty on neighbour differences and
a flatter surface; in the limit of very large $\tau$ the effect collapses
to a constant and the spatial structure vanishes. The interval here sits
well away from that limit, so the data are telling you the pattern has
real wiggle worth keeping. Reading $\tau$ alongside the recovered map is a
useful habit. An apparently structured map whose $\tau$ interval drifts
toward large values warns that the structure may be the prior talking
rather than the data; a flat map with a small $\tau$ is the data declining
to find any pattern at all.

## Interpreting the fit

Two points govern how to read a nested areal fit.

The first is uncertainty. The slope's standard error in `summary()` is not
the conditional error you would get by fixing $\tau$ at one value. It is
marginalised over the grid of $\tau$ values, so it already includes the
cost of not knowing how smooth the field is. `confint()` returns the same
bounds as a matrix, and `vcov()` the fixed-effect covariance.

```{r confint}
confint(fit)
```

The second point is the intercept, and it follows directly from the
rank-deficiency of the ICAR precision. Under an intrinsic field the
overall level of $\phi$ and the intercept $\beta_0$ are confounded:
shifting every $\phi_r$ up by a constant and the intercept down by the
same constant leaves the linear predictor, and therefore the likelihood,
unchanged. The two parameters trade off along a flat ridge, and no amount
of data can pin them both. tulpa handles this by letting the intercept
float against a near-flat prior, which shows up as a very wide intercept
standard error in the summary above. The slopes are unaffected, because a
covariate's effect does not live on that ridge. Read the slopes, which are
identified, and treat the intercept as a nuisance level rather than a
quantity to interpret.

```{r intercept-se}
summary(fit)["(Intercept)", "std.error"]
```

The wide value here is by design, not a fitting failure. A proper CAR or a
BYM2 field with its IID part tamed will narrow it, at the cost of an extra
estimated parameter. Whenever you report results from an intrinsic model,
say so plainly: the intercept is a floating reference, the slopes are the
inferential payload, and a reader who treats the intercept as an absolute
baseline rate will be misled. This caveat travels with every disease-map
or species-distribution table that smooths over an adjacency graph.

## Prediction

`predict()` gives the fixed-effect prediction at new covariate values.
The spatial field is held at its population level, so the result is the
marginal covariate trend with the regional signal averaged out. This is
the right default for a question like "what does the covariate do on
average across the map", as opposed to "what happens in region 7", which
needs that region's field value added back. `se.fit = TRUE` adds the
link-scale standard error and credible bounds, drawn from the same
grid-marginalised fixed-effect covariance the summary uses.

```{r predict}
nd <- data.frame(x = seq(-2.5, 2.5, length.out = 50))
pr <- predict(fit, newdata = nd, type = "link", se.fit = TRUE)
head(pr, 3)
```

Plot the trend on the link scale with its credible band.

```{r predict-plot, fig.alt = "Predicted logit response across x with a 95 percent credible band"}
plot(nd$x, pr$fit, type = "n", xlab = "x", ylab = "predicted logit")
polygon(c(nd$x, rev(nd$x)), c(pr$lower, rev(pr$upper)),
        col = adjustcolor("steelblue", 0.25), border = NA)
lines(nd$x, pr$fit, lwd = 2)
```

For a probability instead of a logit, pass `type = "response"`; the
prediction and its bounds then pass through the inverse-logit and stay
inside $[0, 1]$. Region-specific predictions come from adding the
extracted `field_hat[r]` to the link-scale fit for that region's rows,
then mapping through the link. The field value carries the region's
departure from the average, so a region in the wave's peak predicts above
the marginal trend and one in the trough below it, by exactly the field
contrast you recovered earlier.

## Comparing against a non-spatial fit

Does the field earn its place? A high correlation with a known truth says
yes in simulation, but on real data there is no truth to check against.
What you can still ask is whether the field improves the model enough to
justify its parameters, and the marginal likelihood answers exactly that. It
is the probability of the data with every latent quantity integrated out,
so a model is not rewarded for the extra flexibility of a field unless
that flexibility buys real fit. Compare the same mean structure with and
without the field on this scale.

A nested fit reports its log marginal likelihood per grid point in
`$log_marginal`, one value for each $\tau$ on the integration grid. The
single integrated evidence is the log-sum-exp of that vector, which adds
the per-grid evidences on the probability scale before taking logs again.
The non-spatial Laplace fit has no grid and reports a scalar log marginal
likelihood directly through `logLik()`.

```{r compare}
m0 <- tulpa(y ~ x, data = ds, family = "binomial", mode = "laplace")

lse  <- function(z) { m <- max(z); m + log(sum(exp(z - m))) }
ev_spatial    <- lse(fit$log_marginal)
ev_nonspatial <- as.numeric(logLik(m0))
c(nonspatial = ev_nonspatial, spatial = ev_spatial)
```

```{r compare-diff}
ev_spatial - ev_nonspatial
```

The spatial fit carries the higher evidence by a wide margin, matching how
the data were built. The difference is the log Bayes factor for the field,
and a value this large is decisive: the data are vastly more probable with
a field than without one. The marginal likelihood is the right currency
because it already integrates over the field and its precision, so adding
the field is not rewarded for free. It has to pay for its parameters
through the integration and still come out ahead, which guards against the
overfitting that a raw goodness-of-fit comparison would invite.

`compare_models()` ranks fits by `criterion = "loglik"`, reading each fit's
`logLik()`. On a nested fit `logLik()` returns the integrated evidence, the
log-sum-exp of `$log_marginal` shown above, so `compare_models()` reports one row
per model and the areal-versus-flat ranking matches the manual differencing. The
same comparison runs for two spatial models, an ICAR against a BYM2 for instance,
by taking the log-sum-exp of each fit's `$log_marginal` and differencing.

## Practical guidance

A few rules of thumb for areal fields, with the numbers that make them
actionable.

- **At least 8 to 10 regions, ideally more.** The precision $\tau$ is
  learned from how much neighbouring regions differ, and a handful of
  units leaves it barely informed. Below roughly eight regions, a fixed
  effect per region is usually the honest choice over a smoothed field,
  since with so few units there is little to gain from borrowing and the
  smoothing parameter is mostly prior. The field repays the extra
  machinery when you have dozens to thousands of regions, many of them
  thin.

- **Check the graph is connected.** A field on a disconnected adjacency
  splits into independent pieces, each pinned to its own floating level,
  and the intercept can no longer anchor them jointly. The number of zero
  eigenvalues of the graph Laplacian `diag(rowSums(W)) - W` counts the
  connected components; more than one means the graph is split.

- **Read slopes, never the intercept, under ICAR.** The intercept
  confounds with the field level and its standard error runs to the
  prior scale, on the order of 100 on the link scale, against slope errors
  three orders of magnitude smaller. That gap is expected and is the
  signature of the intrinsic ridge, not a convergence problem. If you need
  a clean intercept, move to a proper CAR or a BYM2 field, whose IID part
  breaks the exact confounding.

- **ICAR versus BYM2.** Reach for `icar` when you want the simplest smooth
  field and do not need to separate structured from unstructured variation;
  it has one fewer parameter and a clean interpretation as pure smoothing.
  Reach for `bym2` when you want an interpretable spatial fraction
  $\lambda$, when some regional variation is plausibly independent noise
  rather than smooth signal, or when the Riebler scaling matters for
  comparing the field's magnitude across maps with different graphs. In
  disease mapping BYM2 is the common default for exactly these reasons; for
  a quick smoothing pass ICAR is the lighter choice.

- **When a spatial field is the wrong tool.** Three cases call for leaving
  it out. If the residuals from a non-spatial fit show no neighbour
  correlation, a field adds parameters and uncertainty for nothing, and the
  marginal-likelihood comparison above will report equal or lower evidence
  for the spatial model. If you have fewer than about eight regions, the
  smoothing parameter is too weakly informed to help. And if the units are
  not truly areal, raw points on a map rather than regions joined by
  borders, a continuous field fits the geometry better than an adjacency
  you would have to invent by binning the points into arbitrary cells.

## Continuous fields

The areal path assumes the world comes pre-divided into regions with
borders. Often it does not. Survey points, sensor locations, and trap
sites sit on continuous coordinates, and the spatial signal is smooth in
distance rather than tied to administrative units. When the signal lives
on coordinates rather than discrete regions, tulpa carries three
continuous fields. Each takes its coordinates in the spec, so there is no
`spatial(col)` term and no adjacency to build:

- `spatial_gp(~ lon + lat)` fits a nearest-neighbour Gaussian process
  (NNGP), an exact GP thinned to each point's nearest neighbours so it
  scales to large data.
- `spatial_hsgp(~ lon + lat)` fits a Hilbert-space approximate GP, a sum
  of basis functions that is fast for moderate ranges.
- `spatial_spde(~ lon + lat, data)` fits a Matern field through the SPDE
  representation on a mesh, routed to the dedicated `spde` backend.

```{r continuous-spec}
gp_spec <- spatial_gp(~ lon + lat)
gp_spec$type
```

The call shape mirrors the areal path: pass the spec through `spatial=`,
choose a mode, read the same accessors off the result. What changes is the
smoothing geometry. An areal field penalises differences between
neighbours on a graph; a continuous field penalises roughness as a
function of euclidean distance, with a range parameter that says how far
the correlation reaches before it decays. NNGP and HSGP trade a little
exactness for the speed needed on large coordinate sets, while the SPDE
representation links a Matern field to a sparse precision through a mesh,
which keeps the Gaussian-Markov advantages on continuous space. The choice
among the three is mostly a question of sample size and how smooth you
expect the surface to be.

The SPDE field has a dedicated fitter. `fit_spde()` takes the response, a
design matrix, and the `spatial_spde` spec directly, builds the Matern
precision on the mesh, and solves the field by sparse Laplace. Simulate a
smooth Poisson surface on scattered coordinates and fit it with the range
and marginal SD held fixed:

```{r spde-fit}
set.seed(20260531)
n      <- 300
coords <- data.frame(lon = runif(n), lat = runif(n))
field  <- 1.4 * (sin(2.5 * coords$lon) + cos(2.5 * coords$lat))
field  <- field - mean(field)
xcov   <- rnorm(n)
y      <- rpois(n, exp(1.0 + 0.5 * xcov + field))

spde <- spatial_spde(~ lon + lat, data = coords, max_edge = c(0.2, 0.5))
fit  <- fit_spde(y = y, X = model.matrix(~ xcov), spatial = spde,
                 family = "poisson", range = 0.4, sigma = 0.9)
round(fit$beta, 3)
```

The fixed-effect estimates land on the generating intercept and slope; the
mesh-node field effects sit in `fit$spatial_effects`, ready to project onto
a prediction grid. Leaving `range` and `sigma` at their default `NULL`
switches `fit_spde()` to nested Laplace over the two hyperparameters and
adds an outer Pareto-k-hat to the result (`fit$pareto_k`), the same
accuracy diagnostic the areal nested integration reports.

## See also

- `vignettes/quickstart.Rmd` for the engine basics: families, tiers,
  random effects, prediction.
- `vignettes/tgmrf.Rmd` for user-defined GMRF latent blocks, the
  general form behind the built-in fields.
- `?spatial_car`, `?spatial_bym2`, `?spatial_gp` for the spec
  constructors and their references.
