---
title: "Temporal random-walk effects with tulpa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Temporal random-walk effects with tulpa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260529)
```

```{r load, message = FALSE}
library(tulpa)
```

## Why a temporal field

Many datasets carry a clock. Counts of a species are taken year after year,
disease cases arrive month by month, a sensor logs a reading every hour. The
quantity you care about is usually a covariate effect: does abundance rise with
temperature, does risk fall with vaccination coverage. The clock sits in the
background, and the background moves.

When the background moves smoothly and you ignore it, the covariate pays for
the omission. Suppose abundance drifts upward over a decade for reasons you
never measured, and your covariate also happens to trend upward over the same
decade. A model with no time term will read the shared drift as covariate
signal and report a slope that is too steep. The trend and the covariate are
confounded through time. This is the temporal twin of spatial confounding, and
the fix has the same shape: give the model a flexible function of time, let it
soak up the smooth background, and read the covariate effect off what remains.

The confounding has a mechanism worth stating plainly. A regression slope is
the covariation of the response and the covariate divided by the variance of the
covariate. If both response and covariate share a slow drift, part of that
covariation comes from the drift rather than from any direct link between them.
The slope cannot tell the two sources apart on its own. A flexible time term
gives it a way to: anything that varies smoothly across time can be carried by
the curve, so the slope is left to explain only the part of the covariate that
wiggles faster than the background can follow. The price is that a covariate
which is itself very smooth in time becomes hard to disentangle, and its
interval widens to admit the doubt. That widening is honest. A model with no time
term would have reported a narrow interval around the wrong value.

The choice of how flexible to make the time term is the modelling decision. Too
stiff and the curve cannot absorb the background, leaving the bias in the slope.
Too loose and it chases noise, stealing signal that belonged to the
covariate. A random walk handles this by learning its own flexibility from the
data through a single smoothing parameter, rather than asking you to pick the
number of knots or the bandwidth by hand. The data decide how much the trend is
allowed to bend.

tulpa offers that flexible function as a first-order random walk. You declare
it once, hand it to `tulpa()` through the `temporal=` argument, and the engine
threads it through the same nested-Laplace machinery that integrates a spatial
field. The covariate slope comes back with a marginalized standard error, the
trend itself can be pulled out time point by time point, and the smoothing
strength is summarised by a single hyperparameter. This vignette walks the full
path: write down the model, simulate a known trend, fit, check the recovery,
read the uncertainty honestly, predict, and compare against a model with no
time term.

A note on scope before we start. The front-door route through `tulpa()` carries
the first- and second-order random walks and the autoregressive process,
`temporal_rw1()`, `temporal_rw2()`, and `temporal_ar1()`. A panel (per-group)
trend runs through the same door, and a temporal field can sit beside an areal
(ICAR, BYM2, CAR) spatial field. Pairing a panel trend with a spatial field, or a
temporal field with a continuous (GP, NNGP, HSGP) or SPDE spatial field, raises a
clean error and points you at the direct fitters. The main example below uses
`temporal_rw1()`; the closing section runs the others.

## The model

Write the linear predictor for observation $i$ as

$$
\eta_i \;=\; \beta_0 + x_i^\top \beta + \phi_{t(i)},
$$

where $t(i)$ is the time index of observation $i$ and $\phi = (\phi_1, \dots,
\phi_T)$ is a vector of one effect per time point. The covariate part $\beta_0 +
x_i^\top \beta$ is the usual fixed-effect structure. The new piece is $\phi$,
and the model is its prior.

The prior on $\phi$ encodes a belief about time: nearby time points are alike.
Rather than impose a parametric shape such as a line or a polynomial, a random
walk imposes only local smoothness and lets the global shape emerge. This is why
it suits a background you cannot name. The claim is modest: the trend does not
jump, and beyond that it can take whatever shape the data support, quadratic or
sinusoidal or neither.

A first-order random walk says that consecutive time points should be close.
Formally, the increments are independent and Gaussian,

$$
\phi_t - \phi_{t-1} \;\sim\; \mathcal{N}\!\left(0,\; \tau^{-1}\right),
\qquad t = 2, \dots, T,
$$

with precision $\tau$. Large $\tau$ pins neighbouring time points together and
forces a nearly flat trend; small $\tau$ lets the walk wander. The joint prior
on $\phi$ is a Gaussian Markov random field with precision matrix $Q(\tau) =
\tau R$, where $R$ is the structure matrix

$$
R_{tt} =
\begin{cases}
1 & t \in \{1, T\} \\
2 & 1 < t < T,
\end{cases}
\qquad
R_{t,\,t-1} = R_{t-1,\,t} = -1,
$$

and all other entries zero. Reading $R$ row by row reveals what it is: the
penalty on $\phi_t$ ties it to its two chain neighbours, exactly the intrinsic
conditional-autoregressive (ICAR) penalty you would write for a spatial field on
a graph whose graph happens to be a line. An RW1 is an ICAR on a chain. That
identity is what lets tulpa reuse one integration path for both: the chain
adjacency (time $t$ neighbours $t-1$ and $t+1$) produces the precision above,
and the nested-Laplace driver integrates it.

The conditional form of the prior makes the smoothing concrete. Conditional on
its neighbours, an interior time point has mean equal to the average of the two
adjacent points,

$$
\mathbb{E}\!\left[\phi_t \mid \phi_{-t}\right]
= \tfrac{1}{2}\left(\phi_{t-1} + \phi_{t+1}\right),
\qquad
\mathrm{Var}\!\left[\phi_t \mid \phi_{-t}\right]
= \tfrac{1}{2}\,\tau^{-1}.
$$

Each point is pulled toward the midpoint of its neighbours, and the strength of
the pull is set by $\tau$. This is the same conditional structure an ICAR field
imposes on a spatial graph, with two neighbours instead of however many a map
region happens to have. The endpoints, having one neighbour each, are pulled
toward that single neighbour with twice the conditional variance, which is why
the corner entries of $R$ are $1$ rather than $2$.

Why an intrinsic prior rather than a proper one. The random walk fixes how the
trend changes, not where it sits, and that is exactly the property you want from
a confounding sponge: the field should be free to take whatever level the data
imply without competing with the covariates for it. The cost is the rank
deficiency, handled below by a constraint. The benefit is that the prior says
nothing about the absolute level of the trend, only its shape, which is the
honest position when the background is something you never measured.

The matrix $R$ has rank $T - 1$. The walk fixes increments, not levels, so the
whole trend can slide up or down by a constant without changing any increment.
That constant is the unidentified direction, and it collides with the intercept
$\beta_0$. tulpa resolves it with a sum-to-zero constraint on $\phi$, which has a
consequence we return to when reading the intercept: under an intrinsic field
the intercept absorbs the free level and its standard error sits near the prior
width by design.

One switch changes the geometry. With `cyclic = TRUE` the chain closes into a
ring, so time $T$ becomes a neighbour of time $1$ and the increment $\phi_1 -
\phi_T$ joins the prior. A ring is the right structure for phase-wrapped data
where the last bin really does sit next to the first: month of year, hour of
day, compass bearing. The default is the open chain, which suits a calendar year
sequence where 2010 and 2030 are not neighbours.

The difference shows up at the boundary. On an open chain the endpoints are
weakly tied, each to a single neighbour, so the trend can swing freely at the
ends where the data are thinnest. On a ring every point has two neighbours, the seam included,
so a December estimate borrows strength from January and the trend joins up
without a discontinuity at the wrap. For seasonal data fitted on an open chain,
a spurious jump between the last and first bin is a common artefact; the ring
removes it by construction. The rank deficiency stays at one in both cases, so
the same single constraint identifies the level either way.

## Simulating data

To check that the fit recovers what it should, simulate with a trend you choose.
Lay down $T = 30$ time points and a smooth seasonal-looking trend, a single sine
wave centred at zero so it does not fight the intercept.

```{r sim-trend}
T_pts <- 30L
trend <- as.numeric(scale(sin(2 * pi * seq_len(T_pts) / T_pts)))
```

Scaling to mean zero and unit variance keeps the trend on a comparable footing
with the covariate effect, so neither dwarfs the other in the linear predictor.
Now draw $n = 900$ observations, assign each to a time point, give each a
covariate $x$, and build a binary response. The true intercept is $-0.2$, the
true slope is $0.9$, and each observation picks up the trend value at its time
point.

```{r sim-binomial}
set.seed(42)
n     <- 900L
time  <- sample(seq_len(T_pts), n, replace = TRUE)
x     <- rnorm(n)
eta   <- -0.2 + 0.9 * x + trend[time]
y     <- rbinom(n, 1, plogis(eta))
df    <- data.frame(y = y, x = x, time = time)
```

The `time` column holds integer time indices, but any orderable variable works.
`validate_temporal()` sorts the unique values and maps them to a chain in that
order, so calendar years, dates, or ordered factor levels all land on the same
$1, \dots, T$ index internally. The trend enters every observation that shares a
time point, which is what gives the random walk enough replication per node to
be informed.

A few details in the setup are deliberate. The trend is centred, so it carries
shape but no level, which keeps it from quarrelling with the intercept and
mirrors the sum-to-zero constraint the model will apply. The covariate is drawn
independently of time here, so this simulation tests recovery rather than
confounding correction; the trend and the slope describe genuinely separate
features of the data and a good fit should return both. The thirty time points
with nine hundred observations leave each time point about thirty observations on
average, comfortably enough to inform each $\phi_t$ and the smoothing precision
above them. If you wanted to study confounding directly, you would make $x$
correlate with `trend[time]` and watch the no-trend slope inflate while the
temporal model holds near $0.9$.

The single sine wave is a convenient truth because it is smooth, bounded, and
has no preferred level, so the centred trend the model recovers should overlay it
cleanly. The walk knows nothing of sine waves; it learns only that neighbours
are close, and the sine wave is a fair test of whether that local smoothness is
enough to reconstruct a global shape from noisy binary data.

## Fitting

Declare the temporal structure with `temporal_rw1()`, naming the time column,
and pass the spec through `temporal=`. The spec carries the time variable
itself, the same way a continuous spatial spec carries its coordinates, so there
is no `temporal(col)` term in the formula.

```{r fit-binomial}
tspec <- temporal_rw1("time")
fit <- tulpa(y ~ x, data = df, family = "binomial",
             temporal = tspec, mode = "auto")
coef(fit)
```

The slope lands near the true $0.9$. The intercept reads near zero with a very
wide interval, the sum-to-zero constraint at work, covered below. How was the
fit routed?

```{r routing}
c(backend = fit$backend, tier = fit$inference_tier)
fit$selection_reason
```

`mode = "auto"` recognised the temporal field, turned the RW1 into its chain
ICAR block, and sent it to the nested-Laplace backend (Tier 2). That backend
puts a grid over the smoothing precision $\tau$, runs an inner Laplace fit at
each grid point, and integrates over the grid. The conditional `mode = "laplace"`
path is not wired for temporal fields yet, so every temporal selection routes to
the nested integrator.

The two-layer structure is worth picturing. The inner layer is an ordinary
Laplace fit: hold $\tau$ fixed, find the mode of the fixed effects and the trend
together, and read the curvature there. The outer layer treats $\tau$ as the
unknown it is, places a grid of candidate values across a sensible range, runs
the inner fit at each, and weights the results by how well each $\tau$ explains
the data. Nothing is fixed at a single best smoothness. The reported slope and
its uncertainty are an average across the grid, so a fit that is genuinely unsure
how smooth the trend should be will say so through a wider slope interval rather
than a falsely confident one. `summary()` gives the fixed-effect table with
marginalized standard errors and credible bounds.

```{r summary-binomial}
summary(fit)
```

`confint()` returns the same bounds as a matrix.

```{r confint-binomial}
confint(fit)
```

### Extracting the temporal trend

The fixed-effect accessors report $\beta$, but the trend $\phi$ is the reason
you added the field. A nested fit carries the per-grid posterior modes in
`$modes`, a matrix with one row per grid point and one column per latent
coordinate, ordered as `[fixed effects, field]`. It also carries the grid
`$weights`. The grid-marginalized value of the field at time $u$ is the
weight-averaged mode in the column for that time point.

```{r extract-trend}
w     <- fit$weights / sum(fit$weights)
nf    <- fit$n_fixed
phi_hat <- vapply(seq_len(T_pts),
                  function(u) sum(w * fit$modes[, nf + u]),
                  numeric(1))
round(head(phi_hat), 3)
```

The offset `nf + u` skips the `nf` fixed-effect columns and lands on the field
column for time $u$. The weighted sum integrates over the smoothing precision
rather than fixing it at a point, which is the whole reason for the grid. This
recipe is identical for a spatial field; only the meaning of the index changes.

The column ordering is the key to reading `$modes`. Each row is one grid point's
joint posterior mode for the latent vector, laid out as the fixed effects first
and the field after, so column `nf + u` is always time point $u$ regardless of
how many grid points the integrator chose. Averaging down the rows with the grid
weights collapses the smoothing uncertainty into a single trend value per time
point. If you wanted a credible band on the trend rather than a point, you would
carry the per-grid curvature for the field columns through the same law of total
variance the fixed effects use, combining within-grid variance and
between-grid spread; the point estimate above is enough for the recovery check
this vignette runs.

## Checking the fit

A recovery check plots the estimated trend against the truth. Because the
random walk is identified only up to an additive constant, centre both before
comparing.

```{r trend-plot, fig.alt = "Estimated RW1 trend across time points overlaid on the true simulated trend"}
plot(seq_len(T_pts), trend - mean(trend), type = "l", lwd = 2,
     xlab = "time point", ylab = "centred temporal effect")
lines(seq_len(T_pts), phi_hat - mean(phi_hat), col = "darkorange", lwd = 2)
legend("topright", c("truth", "estimate"),
       col = c("black", "darkorange"), lwd = 2, bty = "n")
```

The estimate tracks the sine wave it was built from, smoothed a touch where the
prior pulls neighbours together. The correlation between the two confirms it.

```{r trend-cor}
round(cor(phi_hat, trend), 3)
```

The smoothing precision $\tau$ controls how tightly the walk hugs a straight
line, and the fit reports its posterior summary directly. `$theta_mean` is the
posterior mean of the field's hyperparameter, with `$theta_ci_lo` and
`$theta_ci_hi` bounding it.

```{r tau-summary}
c(mean = fit$theta_mean,
  lower = fit$theta_ci_lo,
  upper = fit$theta_ci_hi)
```

A larger $\tau$ means a stiffer walk. Here the interval excludes the very large
values that would flatten the trend to a line, so the data have spoken for some
curvature, consistent with the sine wave underneath.

Reading the precision interval is a useful diagnostic in its own right. If the
posterior for $\tau$ piled up against the upper edge of the grid, that would say
the data prefer a trend so smooth it is nearly flat, a hint that the field is
absorbing little and might be dropped. If it piled against the lower edge, the
trend would be straining to wiggle as much as the grid allows, a hint that the
prior is too restrictive or that something faster than a smooth trend is at work.
An interval comfortably inside the grid, as here, says the smoothing strength is
identified and the trend is doing real work without overfitting. The grid bounds
themselves come from the spec, so a precision pinned at an edge is also a cue to
widen them and refit.

## Interpreting the fit

Two points deserve care when reading the table.

First, the standard error on the slope is marginalized. The nested driver does
not report the slope uncertainty at a single best $\tau$; it folds the per-grid
fixed-effect covariances together through the law of total variance, so the
interval already accounts for not knowing the smoothing strength exactly.
`tulpa()` keeps the per-grid Hessians for this purpose. The practical reading is
that the slope interval is honest about the trend being estimated rather than
known.

The law of total variance is the right tool because the slope uncertainty has
two sources. Within any single grid point the inner Laplace fit gives a
covariance for the fixed effects, the curvature of the log-posterior at that
$\tau$. Across grid points the slope estimate itself moves, because a stiffer or
looser trend leaves a slightly different residual for the covariate to explain.
The total variance adds two pieces: the average of the within-grid covariances,
plus the spread of the per-grid estimates. Both the conditional uncertainty and
the smoothing uncertainty land in one interval. A naive fit that picked the single
best $\tau$ and reported its covariance would capture only the first term and
understate the interval, sometimes badly when the trend is hard to pin down.
The coverage of the marginalized interval is tested at its nominal rate, so the
ninety-five percent bounds mean what they say.

Second, the intercept. Under an intrinsic field the level of the trend is not
separately identified from $\beta_0$, so the sum-to-zero constraint hands the
intercept a near-flat posterior and its standard error sits at roughly the prior
width.

```{r intercept-se}
summary(fit)["(Intercept)", c("estimate", "std.error")]
```

That large standard error is a feature of the parameterisation, not a sign the
fit failed. Interpret slopes, which are identified and marginalized; do not read
the intercept as an estimate of the overall level. If you need a level, it lives
in the trend $\phi$, which you extract and centre as shown above.

The reason traces straight back to the intrinsic prior. The random walk
constrains increments, so the data inform the shape of $\phi$ and the height of
$\beta_0 + \bar\phi$ taken together, but not the split between them. tulpa pins
$\phi$ to sum to zero and lets $\beta_0$ float, which is one valid way to break
the tie, and under that choice the intercept inherits the unidentified level.
A different software might pin the intercept and let the trend float; the slopes,
which depend on neither convention, come out the same. The lesson is to anchor
interpretation on quantities that do not depend on how the tie was broken.
Slopes qualify. Differences in the trend between two time points qualify too,
since the shared level cancels. The bare intercept stands apart, tied to a
convention rather than to the data.

## Prediction

`predict()` gives the fixed-effect prediction at new covariate values, with the
field held at its population average. On the link scale the intercept's wide
posterior flows straight into the standard error, so a raw link-scale band is
dominated by that unidentified level rather than by covariate uncertainty.

```{r predict-link}
nd <- data.frame(x = seq(-2, 2, length.out = 50))
pr <- predict(fit, newdata = nd, type = "link", se.fit = TRUE)
round(head(pr, 2), 3)
```

The informative quantity is how the prediction *changes* with $x$, which is the
slope and its tight interval, not the absolute height of the line. To show the
covariate effect cleanly, plot the link-scale prediction centred at its own
mean, so the intercept's free level drops out and the slope's uncertainty is
what remains.

```{r predict-plot, fig.alt = "Centred link-scale prediction across x with the covariate slope and its credible band"}
ctr <- pr$fit - mean(pr$fit)
plot(nd$x, ctr, type = "l", lwd = 2,
     xlab = "x", ylab = "centred linear predictor")
band <- (pr$upper - pr$lower) / 2
polygon(c(nd$x, rev(nd$x)),
        c(ctr - band, rev(ctr + band)),
        col = adjustcolor("steelblue", 0.25), border = NA)
lines(nd$x, ctr, lwd = 2)
```

The band is narrow because the slope is well identified, even though the line's
absolute position is not. Centring is the right move precisely because the
quantity of interest is a contrast: the difference in linear predictor between
two covariate values, which the slope controls and the intercept cancels out of.
The raw `se.fit` is not wrong, it is answering a different question, namely how
uncertain the absolute height of the line is, and under an intrinsic field that
height is barely identified.

`predict()` holds the temporal field at its population average. What you see is
the covariate effect with the trend marginalised out, not a forecast at a
particular time. To predict at a specific time point, add that time's $\phi$
value from the extracted trend to the linear predictor before applying the
inverse link, and set the intercept to whatever level you want to anchor on.
The engine's population-level `predict()` deliberately stays on the marginal
trend, so the default answer is the covariate relationship rather than a
time-specific one.

## Comparing against no temporal field

Does the trend earn its place? Fit the same model without a time term and
compare evidence. On a Laplace-tier fit `logLik()` returns the approximate log
marginal likelihood, the quantity to compare across specifications.

```{r fit-notemporal}
m_nt <- tulpa(y ~ x, data = df, family = "binomial", mode = "laplace")
as.numeric(logLik(m_nt))
```

A nested fit integrates over a grid of $\tau$, so its `$log_marginal` is a vector
with one entry per grid point. The model evidence is the log of the
grid-summed marginal likelihood, a log-sum-exp over that vector.

```{r evidence}
lse <- function(v) { m <- max(v); m + log(sum(exp(v - m))) }
evidence_temporal <- lse(fit$log_marginal)
c(no_temporal = as.numeric(logLik(m_nt)), temporal = evidence_temporal)
```

The temporal model carries substantially higher evidence, matching the fact that
a trend really was in the data. `compare_models()` ranks fits by a shared
criterion, reading each one's `logLik()`.

```{r compare}
cmp <- compare_models(no_temporal = m_nt,
                      temporal = fit,
                      criterion = "loglik")
cmp
```

On the nested temporal fit `logLik()` returns the integrated evidence, the
log-sum-exp of `$log_marginal` computed above, so `compare_models()` carries one
row per model and reproduces the manual comparison alongside the non-temporal
`logLik()`.

A word on what this comparison settles. The marginal likelihood already
integrates over the walk and its smoothing precision, so it charges the temporal
model for the extra flexibility it brought, and a higher value therefore means
the trend paid for itself once that charge was levied. The gain in fit outweighed
the added complexity. That is the property that makes the marginal likelihood the
right yardstick for nested specifications, where one model is the other with a
structural piece added. The raw maximised likelihood cannot do this job, since it
always rises with more parameters and would crown the temporal model even when
the curve was only chasing noise. Predictive scores such as WAIC and LOO answer a
related but separate question about out-of-sample accuracy; they need a pointwise
log-likelihood the base engine does not carry, so they live in the model packages
built on it.

## A second family

The same call shape works for a Gaussian response; only `family` and the
dispersion `phi` change. Reuse the trend and slope, swap the binary draw for a
Gaussian one.

```{r fit-gaussian}
set.seed(7)
yg  <- -0.2 + 0.9 * x + trend[time] + rnorm(n, sd = 0.5)
dfg <- data.frame(y = yg, x = x, time = time)
fitg <- tulpa(y ~ x, data = dfg, family = "gaussian",
              temporal = tspec, mode = "auto", phi = 0.5)
coef(fitg)["x"]
```

The slope again recovers near $0.9$, and the same field-extraction recipe lifts
the trend out of `fitg$modes`. For the Gaussian family the dispersion `phi` is
the residual variance, conditioned on rather than estimated on this path, so the
value you pass should reflect the noise you expect; here it matches the `sd =
0.5` the data were drawn with. The mechanics are otherwise unchanged. The same
chain ICAR block, the same grid over $\tau$, the same marginalized slope. Only
the inner likelihood differs, which is the point of routing every family through
one integrator.

## Practical guidance

A handful of rules of thumb for everyday use.

- **Give the walk enough time points.** An RW1 needs at least 2 points to be
  defined, but a trend with fewer than about 8 to 10 time points has little room
  to bend, and the smoothing precision is then weakly informed. Below that,
  consider treating time as a small set of fixed effects instead. Replication
  matters as much as count: many observations per time point inform each $\phi_t$
  far better than one observation each.

- **Pick the smoothness to match the trend you expect.** RW1 penalises first
  differences, so it favours piecewise-flat trends and can look a little blocky.
  RW2 penalises second differences and favours trends that are locally straight,
  giving a smoother curve. If the underlying process is a gentle long-term drift,
  RW2 is usually the better prior; for trends that genuinely change level in
  steps, RW1 fits the shape. RW2 needs at least 3 time points.

- **Use `cyclic = TRUE` for phase-wrapped data.** Month of year, hour of day, and
  compass bearing all close on themselves, so December sits next to January and
  the ring's wrap-around increment lets the trend join up across the seam. Leave
  it `FALSE` for an open calendar sequence whose endpoints are not neighbours.

- **Read slopes, not the intercept.** Under the intrinsic RW1 the intercept's
  standard error sits near the prior width because the trend level is not
  separately identified. The covariate slopes are identified and their intervals
  are marginalized over $\tau$. If you need the overall level, read it from the
  extracted, centred trend.

- **Skip the field when time carries no shared signal.** If observations are not
  clustered in time, or the response has no plausible smooth temporal component,
  a random walk only adds parameters and softens the slope. Fit with and without
  the term and compare evidence, as above; let the data say whether the trend
  earns its place.

### What is wired, and what is not yet

The front door through `tulpa()` carries all three temporal priors. `temporal_rw2()`
penalizes second differences for a smoother trend than RW1:

```{r rw2}
fit_rw2 <- tulpa(y ~ x, data = df, family = "binomial",
                 temporal = temporal_rw2("time"), mode = "auto")
coef(fit_rw2)
```

`temporal_ar1()` is a stationary, full-rank alternative with an estimated
correlation parameter rather than an intrinsic walk. Panel trends through
`group_var=` fit a separate walk per group sharing one hyperparameter, and a
temporal field combined with an **areal** (`icar`/`car`/`bym2`/`car_proper`)
spatial field forms an additive space-time joint prior over the
`[spatial, temporal]` block stack:

```{r ar1-panel, eval = FALSE}
# AR1 temporal trend
tulpa(y ~ x, data = df, family = "binomial",
      temporal = temporal_ar1("time"), mode = "auto")

# Panel: one walk per site, shared smoothness
tulpa(y ~ x, data = panel_df, family = "binomial",
      temporal = temporal_rw1("time", group_var = "site"), mode = "auto")

# Additive space-time
tulpa(y ~ x + spatial(region), data = st_df, family = "binomial",
      spatial = list(type = "icar", adjacency = W),
      temporal = temporal_rw1("time"), mode = "auto")
```

Still off the front door: a panel temporal field alongside a spatial or
`latent()` block, and continuous (`gp`/`nngp`/`hsgp`) or SPDE space-time. Those
report a clear error rather than dropping the term; the dedicated `cpp_nested_laplace_st_*`
kernels exist for later wiring.

## See also

- `?temporal_rw1`: constructor reference.
- The spatial vignette: the same nested-Laplace path on an areal ICAR field.
- The `tgmrf()` vignette: building a custom GMRF latent block, including a
  periodic AR(1), end to end.
- The inference-modes vignette: what the Tier-2 nested-Laplace guarantee means.
