---
title: "User-defined GMRF latent blocks with tgmrf()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User-defined GMRF latent blocks with tgmrf()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260518)
```

```{r load, message = FALSE}
library(tulpa)
library(Matrix)
```

## What `tgmrf()` is

`tgmrf()` lets a *user script* define a Gaussian Markov Random Field
(GMRF) latent block and plug it into tulpa's inference layers as a
first-class latent term. It is the latent-side dual of `LikelihoodSpec`:
that interface lets a downstream package own the observation model;
`tgmrf()` lets a single R script own one latent block.

The contract is two closures and an init vector:

```r
my_block <- tgmrf(
  Q     = function(theta) { ... returns dgCMatrix ... },
  prior = function(theta) { ... returns scalar log-density ... },
  init  = c(...)
)
```

The closures are called numerically: no autodiff, no DSL, no codegen.
For a Gaussian latent block

$$
\log p(z \mid \theta) \;=\; \tfrac{1}{2}\,\log\det Q(\theta) - \tfrac{1}{2}(z-\mu)^\top Q(\theta)(z-\mu) + \text{const}
$$

the score and Hessian in $z$ are closed-form:

$$
\partial_z \log p(z \mid \theta) = -Q(\theta)(z-\mu), \qquad
\partial_z^{\,2} \log p(z \mid \theta) = -Q(\theta).
$$

So tulpa's Laplace inner step only needs `Q(theta)` and `mu(theta)` at
numeric `theta`. Hyperparameter gradients (`d/d theta`) come from
finite differences, typically 2-5 extra `Q` evaluations per outer
step, which is cheap when `Q` is sparse. The same block reaches every
inference tier: Laplace, IMH-Laplace, NUTS, VI.

## How a block becomes a fitted model: the algorithm

A `tgmrf` object carries everything the engine needs and nothing it
does not. The construct is a *templated GMRF latent block*: templated
because the same machinery integrates any precision family the closures
describe, latent because the block contributes a vector $z$ that lives
under the linear predictor rather than in the likelihood. When you call
`tgmrf()`, the constructor evaluates `Q(init)` exactly once. That single
call does three jobs at registration time. It catches user errors early
(a `Q` that throws, returns a dense matrix, or returns a non-square or
asymmetric matrix all stop here with a specific message). It infers
`n_latent` from `nrow(Q(init))`. And it captures the sparsity pattern,
the set of stored nonzero positions, which the inner solver reuses at
every later `theta` so the symbolic Cholesky factorisation is computed
once, not per grid point.

The block records `theta_dim` and `theta_names` from `init` (named
elements become the canonical hyperparameter names; unnamed ones get
`theta_1`, `theta_2`, ...), validates the optional `mu`, `graph`,
`bounds`, and `obs_idx` arguments against the inferred shape, and stores
the two closures by reference. Nothing is differentiated, compiled, or
parsed. The returned object has class `c("tgmrf", "tulpa_latent_block")`
and a `type = "tgmrf"` registry key that the nested-Laplace dispatcher
recognises.

At fit time the integration follows a fixed sequence. The steps below
describe what `tulpa_nested_laplace()` does with a `tgmrf` block; the
IMH, VI, and NUTS adapters share the inner machinery and differ only in
how they explore `theta`.

1. **Build the outer grid over `theta`.** If you supplied `bounds`, the
   driver lays a per-axis grid between `lower` and `upper`. The default
   resolution is 5 points per axis; the IMH, VI, and NUTS adapters expose
   this as `pilot_axis_points`. With $d$ hyperparameters the default grid
   is $5^d$ cells, so a two-parameter block is a 25-cell grid. You can
   also hand the driver an explicit grid matrix.
2. **For each grid point `theta_k`, run the inner Laplace solve.** The
   driver assembles the joint mode of $(\beta, z)$ given `theta_k` by
   Newton iteration. Each Newton step needs the score and Hessian in
   $(\beta, z)$. The $z$-part comes from `Q(theta_k)` and `mu(theta_k)`
   in closed form; the likelihood part comes from the chosen `family`
   (or a model-supplied `likelihood`). Because the sparsity pattern was
   captured at registration, the Cholesky of the system reuses the same
   symbolic factor at every grid point.
3. **Score the grid point.** At the inner mode the driver evaluates the
   Laplace approximation to the log marginal likelihood
   $\log p(y, \hat{z}_k \mid \theta_k)$ and adds the user `prior(theta_k)`.
   This single scalar per grid point is what `$log_marginal` holds.
4. **Integrate over the grid.** The per-cell log-marginals become
   normalised integration weights (`$weights`), and the driver reports
   posterior moments of every hyperparameter as weighted means and SDs
   (`$theta_mean`, `$theta_sd`) plus weighted quantiles.
5. **Diagnose the outer approximation.** With the grid integrated, the
   driver fits a Gaussian proposal to the hyperparameter posterior and
   importance-samples it to compute the outer Pareto-$\hat{k}$
   (`$pareto_k`); see the convergence section.

The adapters change only step 1's exploration of `theta`. IMH localises
the grid argmax, builds a finite-difference Hessian there, and proposes
$\theta' \sim N(\text{mode}, \text{scale}^2 H^{-1})$ with an exact
Metropolis accept/reject against `log_marginal(theta) + prior(theta)`.
VI runs L-BFGS on `log_marginal(theta)` and fits a Gaussian at the
optimum. NUTS runs leapfrog integration on the same target with a
finite-difference gradient. Every one of them calls the same inner
Laplace solve to evaluate a candidate `theta`. The inner solve is the
shared kernel, the outer loop is the only thing that varies. That is why
a block written once reaches all four tiers without modification: the
closures describe the latent prior, and the engine owns everything from
the Newton system up.

## A worked example: periodic AR1

The smallest non-trivial GMRF tulpa does not ship is a *periodic* AR1
(useful for diurnal, seasonal, or phase-wrapped data). The precision
is tridiagonal with wrap-around:

$$
Q_{ii} \;=\; \tfrac{1+\rho^2}{\sigma^2}, \qquad
Q_{ij} \;=\; -\,\tfrac{\rho}{\sigma^2}\;\text{ for } |i-j| = 1 \;(\mathrm{mod}\,n).
$$

Reparameterise to unconstrained `theta = (log_sigma, atanh_rho)`:

```{r block}
periodic_ar1 <- function(n) {
  tgmrf(
    Q = function(theta) {
      sigma <- exp(theta[1]); rho <- tanh(theta[2])
      d <- rep((1 + rho^2) / sigma^2, n)
      o <- rep(-rho / sigma^2, n)
      M <- Matrix::bandSparse(n, k = c(-1, 0, 1),
                              diagonals = list(o, d, o))
      M[1, n] <- M[n, 1] <- -rho / sigma^2     # wrap
      methods::as(methods::as(M, "generalMatrix"), "CsparseMatrix")
    },
    prior = function(theta) {
      dnorm(theta[1], 0, 1, log = TRUE) +     # weak on log_sigma
      dnorm(theta[2], 0, 1, log = TRUE)       # weak on atanh_rho
    },
    init   = c(log_sigma = 0, atanh_rho = atanh(0.3)),
    bounds = list(lower = c(log(0.3), atanh(0.0)),
                  upper = c(log(3.0), atanh(0.9))),
    name   = "periodic_ar1"
  )
}
```

The optional `bounds` argument sets the outer grid for nested-Laplace
and the hard-wall region for IMH and NUTS. The `prior` closure is a
log-density on `theta`. If you don't have prior beliefs, `function(theta) 0`
is fine; it's a flat prior on the unconstrained scale.

The closure shape is the contract. `Q` takes one numeric vector `theta`
and returns a square symmetric sparse matrix; `prior` takes the same
`theta` and returns one finite scalar; `init` is the numeric vector both
closures expect. The reparameterisation to `log_sigma` and `atanh_rho`
is deliberate: it maps the natural constraints $\sigma > 0$ and
$|\rho| < 1$ to the whole real line, so the outer grid, the L-BFGS
search, and the leapfrog steps all operate on unconstrained coordinates
and never propose an invalid precision. The `bounds` you give are in
those unconstrained units too. Use names on `init` and they propagate
through every fit as the reported parameter names.

### Simulate from the block

```{r sim}
n <- 40L
theta_true <- c(log(0.8), atanh(0.6))
blk        <- periodic_ar1(n)

# z | theta_true ~ N(0, Q(theta_true)^{-1})
Q_true <- blk$Q(theta_true)
L      <- Matrix::Cholesky(Q_true)
z_true <- as.numeric(Matrix::solve(L, rnorm(n), system = "Lt"))

# y_i ~ Poisson(exp(beta_0 + z_i))
beta0 <- 0.3
y     <- rpois(n, exp(beta0 + z_true))
X     <- matrix(1, n, 1L)
```

### Tier 2 (structured): nested Laplace

```{r laplace}
fit_lap <- tulpa_nested_laplace(
  y = y, n_trials = rep(1L, n), X = X,
  prior  = blk,
  family = "poisson"
)
fit_lap$theta_mean    # posterior mean over the outer grid
```

### Tier 2 (structured): Pathfinder VI

`tulpa_tgmrf_vi()` runs L-BFGS on `log_marginal(theta)` and fits a
Gaussian at the optimum. Same tier as the nested-Laplace posterior but
gives you a full covariance matrix rather than a discrete grid.

```{r vi}
fit_vi <- tulpa_tgmrf(
  y = y, n_trials = rep(1L, n), X = X, block = blk,
  family  = "poisson", mode = "vi",
  n_draws = 800L
)
fit_vi$mode_theta
fit_vi$elbo
```

### Tier 1 (exact): IMH-Laplace

The Tier-1 upgrade composes the Laplace body with an
independence-MH bias correction. The VI fit makes a near-optimal
proposal: high acceptance, low MC error.

```{r imh}
fit_imh <- tulpa_tgmrf(
  y = y, n_trials = rep(1L, n), X = X, block = blk,
  family = "poisson", mode = "imh",
  n_iter = 1200L, warmup = 400L
)
fit_imh$means
fit_imh$mean_accept
```

### Tier 1 (exact): NUTS over `theta`

For higher-dim `theta` (5+), the gradient-based sibling
`tulpa_tgmrf_nuts()` mixes better than independence-MH. For
`theta_dim = 2` IMH is generally faster; NUTS is the right call when
the posterior is multi-modal or strongly non-Gaussian.

```{r nuts, eval = FALSE}
fit_nuts <- tulpa_tgmrf(
  y = y, n_trials = rep(1L, n), X = X, block = blk,
  family    = "poisson", mode = "nuts",
  n_iter    = 200L, warmup = 100L,
  max_depth = 4L
)
fit_nuts$means
```

## Comparing posteriors

The three estimators target the same posterior `p(theta | y)`:

```{r compare}
data.frame(
  parameter = blk$theta_names,
  true      = theta_true,
  laplace   = fit_lap$theta_mean,
  vi_mode   = fit_vi$mode_theta,
  imh_mean  = fit_imh$means
)
```

Laplace and VI agree to within grid resolution / Gaussian-fit
curvature; IMH lifts the structured-tier bias at the cost of `n_iter`
inner Laplace solves.

## The mathematics

Fix notation first. The model has three layers. The data is
$y = (y_1, \dots, y_N)$ with a per-observation likelihood from `family`.
A fixed-effects design matrix $X$ of dimension $N \times p$ carries
coefficients $\beta \in \mathbb{R}^p$. The latent block contributes a
vector $z \in \mathbb{R}^{n_{\mathrm{lat}}}$ where
$n_{\mathrm{lat}}$ is `n_latent`. An index vector `obs_idx` maps each
observation to a latent slot; when it is `NULL` the driver assumes
$N = n_{\mathrm{lat}}$ and uses row order. The linear predictor for
observation $i$ is
$$
\eta_i \;=\; (X\beta)_i \;+\; z_{\,\mathrm{obs\_idx}[i]},
$$
and the mean is $g^{-1}(\eta_i)$ for the family's link $g$. The
hyperparameter vector is $\theta \in \mathbb{R}^d$ with $d$ equal to
`theta_dim`, living on the unconstrained scale your closures define.

### The GMRF prior

A Gaussian Markov Random Field is a multivariate normal whose precision
matrix $Q$ is sparse, and the sparsity is the conditional-independence
structure: $z_i$ and $z_j$ are conditionally independent given the rest
exactly when $Q_{ij} = 0$. The block density is
$$
p(z \mid \theta) \;=\; (2\pi)^{-n_{\mathrm{lat}}/2}\,
|Q(\theta)|^{1/2}\,
\exp\!\Big(-\tfrac{1}{2}(z-\mu(\theta))^\top Q(\theta)(z-\mu(\theta))\Big),
$$
where $\mu(\theta)$ is the mean vector (your `mu` closure, or zero when
`mu` is `NULL`) and $|Q(\theta)|$ is the determinant. Taking logs,
$$
\log p(z \mid \theta) \;=\;
\tfrac{1}{2}\log|Q(\theta)|
- \tfrac{1}{2}(z-\mu)^\top Q(\theta)(z-\mu)
+ \text{const}.
$$
The const absorbs the $(2\pi)$ factor, which does not depend on $z$ or
$\theta$ and drops out of every gradient. Differentiating in $z$ gives
the closed-form score and Hessian quoted above:
$\partial_z \log p = -Q(z-\mu)$ and $\partial_z^2 \log p = -Q$. This is
why you never write gradient code for the latent block. The Hessian in
$z$ is just $-Q(\theta)$, a constant in $z$, so the latent part of the
Newton system is exact, not an approximation.

### The precision parameterisation

The closure `Q(theta)` is the precision family. Whatever structure you
encode (AR1, periodic AR1, a random walk, a custom spatial GMRF) lives
entirely in how `Q` maps $\theta$ to nonzero entries. The reparameterise
step matters here. The natural parameters of a precision are often
positive scales or bounded correlations, and the integrator works best
on the whole real line, so the convention is to define `theta` on an
unconstrained scale and transform inside `Q`. For the periodic AR1 above,
$\theta = (\log\sigma, \operatorname{atanh}\rho)$ so that $\sigma > 0$
and $|\rho| < 1$ hold automatically. The Jacobian of that transform is
absorbed into your `prior(theta)`: because the prior is a log-density on
the unconstrained $\theta$, you supply whatever density you want there
directly and the engine does not add a change-of-variables term. A flat
`function(theta) 0` is a flat prior on the unconstrained scale, which is
weakly informative on the natural scale.

### The inner Laplace solve

For a fixed $\theta$ the engine needs the marginal likelihood
$p(y \mid \theta) = \int p(y \mid \beta, z)\,p(\beta)\,p(z \mid \theta)\,
\mathrm{d}\beta\,\mathrm{d}z$. The Laplace approximation replaces the
integrand by a Gaussian centred at its mode. Let $w = (\beta, z)$ and
$\ell(w) = \log p(y \mid w) + \log p(z \mid \theta) + \log p(\beta)$ be
the log joint. Newton iteration finds the mode $\hat{w}$ by repeatedly
solving
$$
H(w)\,\Delta w \;=\; -\,\nabla \ell(w), \qquad
H(w) \;=\; -\,\nabla^2 \ell(w),
$$
where $H$ is the negative Hessian of the log joint. The $z$-block of $H$
is $Q(\theta)$ plus the likelihood's Fisher information, the
$\beta$-block is $X^\top W X$ for the family's working weights $W$, and
the cross term couples them through `obs_idx`. The Laplace approximation
to the marginal is
$$
\log p(y \mid \theta) \;\approx\;
\ell(\hat{w}) \;+\; \tfrac{1}{2}\,d_w\log(2\pi)
\;-\; \tfrac{1}{2}\log|H(\hat{w})|,
$$
with $d_w = p + n_{\mathrm{lat}}$ the joint dimension. The engine adds
your `prior(theta)` to this to form `log_marginal(theta) + prior(theta)`,
which is the unnormalised log posterior of $\theta$ that every outer
layer targets. The sparsity pattern captured at registration is what
makes the $\log|H|$ and the linear solve cheap: a sparse Cholesky on a
pattern that does not change across grid points reuses one symbolic
factor.

### Outer integration over `theta`

The nested-Laplace layer treats $\theta$ as the remaining integration
problem. It evaluates $L(\theta_k) = \log p(y \mid \theta_k) +
\log p(\theta_k)$ at each grid point $\theta_k$, normalises to weights
$w_k \propto \exp(L(\theta_k))$ with $\sum_k w_k = 1$, and reports
posterior summaries as weighted statistics over the grid. The posterior
mean of hyperparameter $j$ is $\sum_k w_k\,\theta_{k,j}$ and the SD comes
from the weighted second moment. Derived quantities such as the
natural-scale $\sigma = e^{\theta_1}$ or $\rho = \tanh(\theta_2)$ are summarised by
evaluating the transform at each grid cell and taking weighted quantiles,
not by transforming the mean. The IMH, VI, and NUTS layers replace the
fixed grid by an adaptive exploration of the same $L(\theta)$ surface but
integrate against the identical inner Laplace solve. Every symbol here is
something your two closures and the family definition fully determine.

## Convergence

There are two convergence questions and they belong to different layers.
The first is whether the inner Newton solve found the joint mode of
$(\beta, z)$ at each $\theta$. The second is whether the outer
exploration of $\theta$ (the grid, the L-BFGS optimum, or the MCMC chain)
has actually characterised the hyperparameter posterior. These are
diagnosed separately.

For the inner solve, `tulpa_nested_laplace()` returns `$n_iter`, the
number of Newton iterations spent per grid point, and `$log_marginal`,
the per-cell scores. The inner budget is `control$max_iter` (default
`50L`) with tolerance `control$tol` (default `1e-6`). If a cell hits
`max_iter` without converging, or if `log_marginal` is non-finite at any
cell, the fit is unreliable at that $\theta$: tighten `bounds` so the
grid stays where $Q(\theta)$ is well-conditioned, or raise `max_iter`.
The IMH and NUTS adapters add an eager check: they evaluate the inner
solve at the pilot grid argmax before sampling and stop with a specific
message if it is non-finite, because a failure there is a bug in the
closures rather than numerical infeasibility at the edge.

For the outer approximation on a structured-tier fit, the relevant
number is the Pareto-$\hat{k}$, the iid-fit counterpart of Rhat. The
nested-Laplace fit reports `$pareto_k` and `$pareto_k_is_ess`: the
integrator fits a Gaussian proposal to the hyperparameter posterior,
importance-samples it (`control$k_samples`, default `200L`, each one
extra inner solve), and fits a generalised Pareto tail to the importance
ratios. A $\hat{k} < 0.7$ means the Gaussian grid characterises the
hyperparameter posterior well and the nested integration is reliable.
A $\hat{k} \geq 0.7$ means the posterior is too skewed or heavy-tailed
for the Gaussian grid, and you should escalate to the exact tier
(`tulpa_tgmrf_imh()` or `tulpa_tgmrf_nuts()`). The diagnostic is computed
for a single-block, single positive-scale-axis grid and left `NA` (with
the grid's quadrature effective sample size as the fallback) for
multi-axis or bounded grids. `diagnostic_summary(fit)` surfaces this for
any non-chain fit, and `tulpa_psis()` is the underlying PSIS core if you
want to compute it on your own importance ratios.

For the exact-tier MCMC fits the diagnostics are the usual ones. The IMH
fit reports `$mean_accept`, the post-warmup acceptance rate; a rate in a
healthy band (the VI fit makes a near-optimal proposal, so a good model
sees high acceptance) means the proposal matches the target. A collapsed
acceptance rate is the signal to switch to NUTS. The NUTS fit reports
`$mean_accept` plus `$tree_depth` per draw and the adapted `$epsilon`;
draws repeatedly hitting `max_depth` indicate a step size that is too
small for the posterior geometry. Both produce a `$draws` matrix you can
feed to `mcmc_diagnostics(fit)` for Rhat and effective sample size, or to
`check_diagnostics(fit)` for a pass/fail summary. The cleanest single
check is to run the structured-tier fit, read its Pareto-$\hat{k}$, and
only reach for the MCMC tiers when the diagnostic says the Gaussian
approximation is biased.

## Tuning

The split between statistical arguments and tuning knobs follows the
engine convention. `tulpa_nested_laplace()` carries statistical arguments
(`y`, `n_trials`, `X`, `prior`, `family`, `phi`) at the top level and
puts every numerical or performance knob in a single `control = list()`.
The `tgmrf` adapters expose their tuning as named arguments because each
adapter has a small, specific set.

The first decision is the tier. Start with `tulpa_nested_laplace()` for a
fast structured fit and read its `$pareto_k`. If the diagnostic is below
0.7 the structured posterior is trustworthy and you are done; if it is
above 0.7, escalate. `tulpa_tgmrf_imh()` is the cheapest exact tier and
the right default escalation for low `theta_dim` (2-4) with a roughly
Gaussian posterior, since its proposal is the VI Gaussian, so acceptance stays
high. `tulpa_tgmrf_nuts()` is for higher `theta_dim` (5+) or a strongly
non-Gaussian or multi-modal posterior, where independence-MH acceptance
collapses and gradient information pays for itself.

Inside the structured fit, the knobs that matter are the inner Newton
budget and the diagnostic cost. `control$max_iter` (default `50L`) and
`control$tol` (default `1e-6`) govern the inner solve; raise `max_iter`
only if cells fail to converge. `control$n_threads` (default `1L`) sets
OpenMP threads inside each inner solve, useful when `n_latent` is large.
`control$diagnose_k` (default `TRUE`) and `control$k_samples` (default
`200L`) control the Pareto-$\hat{k}$ computation; set `diagnose_k = FALSE`
to skip the extra inner solves when you do not need the accuracy gate.
The `bounds` you set on the block itself define the grid extent and the
hard wall for IMH and NUTS. Set them generously enough to contain the
posterior mass but tight enough to exclude $\theta$ where $Q$ goes
singular.

The adapters share `pilot_axis_points` (default `5L`), the per-axis
resolution of the pilot grid that initialises every adapter and feeds the
mode and Hessian estimates. Raise it when the default grid is too coarse
to locate the mode. `tulpa_tgmrf_imh()` adds `n_iter` (default `2000L`),
`warmup` (default `n_iter %/% 2L`), `thin`, `scale` (the proposal scale
multiplier, default `1.0`, lower it if acceptance is too high, raise it
if too low), and `fd_step` (default `0.05`, the finite-difference step
for the mode Hessian). `tulpa_tgmrf_vi()` adds `n_draws` (default `1000L`,
the variational draws returned), `max_lbfgs` (default `100L`), and
`lbfgs_tol` (default `1e-6`). `tulpa_tgmrf_nuts()` adds `epsilon` (default
`0.2 / sqrt(theta_dim)`, the initial leapfrog step), `max_depth` (default
`6L`, the tree-depth cap so each draw is at most $2^6$ leapfrog steps),
`target_accept` (default `0.65` for the warmup step-size adaptation), and
`fd_gradient_step` (default `0.02`, the central-difference step for the
gradient). The defaults are tuned for the common `theta_dim <= 5` case;
reach for the knobs when the diagnostics in the previous section flag a
specific failure, not pre-emptively.

## Which tier when

| Tier  | Function                  | Cost                            | When to reach for it |
|-------|---------------------------|---------------------------------|----------------------|
| 2 (structured) | `tulpa_nested_laplace()` | `n_grid` inner Laplace solves       | Quick fit, diagnostic, debias-target |
| 2 (structured) | `tulpa_tgmrf_vi()`       | L-BFGS + `n_draws` extra evals      | Want covariance instead of grid; IMH proposal |
| 1 (exact)      | `tulpa_tgmrf_imh()`      | `n_iter` Laplace solves             | Low theta dim, near-Gaussian posterior |
| 1 (exact)      | `tulpa_tgmrf_nuts()`     | `n_iter * 2^depth * 2 * theta_dim` Laplace solves | High theta dim or non-Gaussian |

The general rule: start with `tulpa_nested_laplace()` for a sanity
check, run `tulpa_tgmrf_imh()` for exact-tier posterior moments, and
reach for `tulpa_tgmrf_nuts()` only when IMH acceptance collapses.

## How this compares to other latent-block interfaces

Several frameworks let a user inject a custom latent structure, and they
trade off along the same axis: how much of the structure you write versus
which inference tiers you reach.

INLA's `rgeneric` interface lets you define a GMRF in R by supplying
functions for the precision and a few related quantities. The callback
runs in R and there is no autodiff, so the block is restricted to INLA's
Laplace machinery: the structured tier only, with no path to exact MCMC
and no debias step on a non-Gaussian residual. `tgmrf()` matches the
ergonomics (two R closures, no compilation) but the same block reaches
every tulpa tier, including the exact IMH and NUTS samplers, because the
outer layers all drive the same inner Laplace solve.

INLA's `cgeneric` interface moves the precision factory to a C function,
which is faster than `rgeneric` but still has no autodiff and still
supports the Laplace tier only: there is no exact-MCMC support for a
`cgeneric` block. The `tgmrf_cpp()` constructor (noted below) is the closer
analog on the speed axis, and it keeps the full-tier reach that `cgeneric`
lacks.

Stan takes the opposite approach: you write the entire model in its DSL,
which is then parsed and compiled to autodiffable C++. That gives exact
HMC over everything, but it asks you to specify the whole model: there
is no notion of plugging one latent block into an engine that owns the
rest, and you pay full HMC cost on every block including the
Gaussian-latent ones a Laplace approximation would handle cheaply.

TMB is the closest analog: you write a templated C++ snippet and get
autodiff through CppAD, with a Laplace approximation over the random
effects. The `tgmrf()` R path is lighter (no C++, closed-form latent
score from $Q$ rather than autodiff) and the `tgmrf_cpp()` path is the
direct counterpart for hot inner loops. The distinguishing move
across all four comparisons is tulpa's nested-approximation-plus-debias
design: the structured tier gives a cheap Laplace answer, and the same
block escalates to an exact-tier correction when the diagnostic says the
approximation is biased, without rewriting anything.

## What `tgmrf()` does not require

- No gradient code. The score in `z` is closed-form from `Q`.
- No Hessian code. The Hessian in `z` is just `-Q`.
- No autodiff. Hyperparameter gradients are finite-difference.
- No C++. The R closure path reaches every tier.

The `tgmrf_cpp()` constructor offers a templated-C++ backend with the
same S3 contract for cases where the R-to-C boundary on `Q(theta)` becomes
the bottleneck (large `n_latent`, hot CCD grids). It takes a `.cpp` file
that defines the `Q`, `mu`, and log-prior kernels as templated C++ and
registers them through the `TULPA_REGISTER_TGMRF` macro;
`inst/examples/tgmrf_periodic_ar1.cpp` is a worked example. The
user-visible API does not change; the same downstream consumers (formula
parser, inference layers, methods) treat the two paths identically.

## See also

- `?tgmrf`: constructor reference.
- `?tgmrf_cpp`: compiled-C++ backend with the same S3 contract.
- `?tulpa_nested_laplace`: Tier-2 grid integrator.
- `?tulpa_tgmrf_imh`, `?tulpa_tgmrf_nuts`, `?tulpa_tgmrf_vi`: adapter family.
- `inst/examples/tgmrf_periodic_ar1.R`: runnable script form of this vignette.
