---
title: "User-defined GMRF latent blocks with tgmrf()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User-defined GMRF latent blocks with tgmrf()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4
)
set.seed(20260518)
```

```{r load, message = FALSE}
library(tulpa)
library(Matrix)
```

## What `tgmrf()` is

`tgmrf()` lets a *user script* define a Gaussian Markov Random Field
(GMRF) latent block and plug it into tulpa's inference layers as a
first-class latent term. It is the latent-side dual of `LikelihoodSpec`:
that interface lets a downstream package own the observation model;
`tgmrf()` lets a single R script own one latent block.

The contract is two closures and an init vector:

```r
my_block <- tgmrf(
  Q     = function(theta) { ... returns dgCMatrix ... },
  prior = function(theta) { ... returns scalar log-density ... },
  init  = c(...)
)
```

The closures are called numerically — no autodiff, no DSL, no codegen.
For a Gaussian latent block

$$
\log p(z \mid \theta) \;=\; \tfrac{1}{2}\,\log\det Q(\theta) - \tfrac{1}{2}(z-\mu)^\top Q(\theta)(z-\mu) + \text{const}
$$

the score and Hessian in $z$ are closed-form:

$$
\partial_z \log p(z \mid \theta) = -Q(\theta)(z-\mu), \qquad
\partial_z^{\,2} \log p(z \mid \theta) = -Q(\theta).
$$

So tulpa's Laplace inner step only needs `Q(theta)` and `mu(theta)` at
numeric `theta`. Hyperparameter gradients (`d/d theta`) come from
finite differences — typically 2-5 extra `Q` evaluations per outer
step, which is cheap when `Q` is sparse. The same block reaches every
inference tier: Laplace, IMH-Laplace, NUTS, VI.

## A worked example: periodic AR1

The smallest non-trivial GMRF tulpa does not ship is a *periodic* AR1
— useful for diurnal, seasonal, or phase-wrapped data. The precision
is tridiagonal with wrap-around:

$$
Q_{ii} \;=\; \tfrac{1+\rho^2}{\sigma^2}, \qquad
Q_{ij} \;=\; -\,\tfrac{\rho}{\sigma^2}\;\text{ for } |i-j| = 1 \;(\mathrm{mod}\,n).
$$

Reparameterise to unconstrained `theta = (log_sigma, atanh_rho)`:

```{r block}
periodic_ar1 <- function(n) {
  tgmrf(
    Q = function(theta) {
      sigma <- exp(theta[1]); rho <- tanh(theta[2])
      d <- rep((1 + rho^2) / sigma^2, n)
      o <- rep(-rho / sigma^2, n)
      M <- Matrix::bandSparse(n, k = c(-1, 0, 1),
                              diagonals = list(o, d, o))
      M[1, n] <- M[n, 1] <- -rho / sigma^2     # wrap
      methods::as(methods::as(M, "generalMatrix"), "CsparseMatrix")
    },
    prior = function(theta) {
      dnorm(theta[1], 0, 1, log = TRUE) +     # weak on log_sigma
      dnorm(theta[2], 0, 1, log = TRUE)       # weak on atanh_rho
    },
    init   = c(log_sigma = 0, atanh_rho = atanh(0.3)),
    bounds = list(lower = c(log(0.3), atanh(0.0)),
                  upper = c(log(3.0), atanh(0.9))),
    name   = "periodic_ar1"
  )
}
```

The optional `bounds` argument sets the outer grid for nested-Laplace
and the hard-wall region for IMH and NUTS. The `prior` closure is a
log-density on `theta`. If you don't have prior beliefs, `function(theta) 0`
is fine — it's a flat prior on the unconstrained scale.

### Simulate from the block

```{r sim}
n <- 40L
theta_true <- c(log(0.8), atanh(0.6))
blk        <- periodic_ar1(n)

# z | theta_true ~ N(0, Q(theta_true)^{-1})
Q_true <- blk$Q(theta_true)
L      <- Matrix::Cholesky(Q_true)
z_true <- as.numeric(Matrix::solve(L, rnorm(n), system = "Lt"))

# y_i ~ Poisson(exp(beta_0 + z_i))
beta0 <- 0.3
y     <- rpois(n, exp(beta0 + z_true))
X     <- matrix(1, n, 1L)
```

### Tier 2 (structured): nested Laplace

```{r laplace}
fit_lap <- tulpa_nested_laplace(
  y = y, n_trials = rep(1L, n), X = X,
  prior  = blk,
  family = "poisson"
)
fit_lap$theta_mean    # posterior mean over the outer grid
```

### Tier 2 (structured): Pathfinder VI

`tulpa_tgmrf_vi()` runs L-BFGS on `log_marginal(theta)` and fits a
Gaussian at the optimum. Same tier as the nested-Laplace posterior but
gives you a full covariance matrix rather than a discrete grid.

```{r vi}
fit_vi <- tulpa_tgmrf_vi(
  y = y, n_trials = rep(1L, n), X = X, block = blk,
  family  = "poisson",
  n_draws = 800L
)
fit_vi$mode_theta
fit_vi$elbo
```

### Tier 1 (exact): IMH-Laplace

The Tier-1 upgrade composes the Laplace body with an
independence-MH bias correction. The VI fit makes a near-optimal
proposal — high acceptance, low MC error.

```{r imh}
fit_imh <- tulpa_tgmrf_imh(
  y = y, n_trials = rep(1L, n), X = X, block = blk,
  family = "poisson",
  n_iter = 1200L, warmup = 400L
)
fit_imh$means
fit_imh$mean_accept
```

### Tier 1 (exact): NUTS over `theta`

For higher-dim `theta` (5+), the gradient-based sibling
`tulpa_tgmrf_nuts()` mixes better than independence-MH. For
`theta_dim = 2` IMH is generally faster; NUTS is the right call when
the posterior is multi-modal or strongly non-Gaussian.

```{r nuts, eval = FALSE}
fit_nuts <- tulpa_tgmrf_nuts(
  y = y, n_trials = rep(1L, n), X = X, block = blk,
  family    = "poisson",
  n_iter    = 200L, warmup = 100L,
  max_depth = 4L
)
fit_nuts$means
```

## Comparing posteriors

The three estimators target the same posterior `p(theta | y)`:

```{r compare}
data.frame(
  parameter = blk$theta_names,
  true      = theta_true,
  laplace   = fit_lap$theta_mean,
  vi_mode   = fit_vi$mode_theta,
  imh_mean  = fit_imh$means
)
```

Laplace and VI agree to within grid resolution / Gaussian-fit
curvature; IMH lifts the structured-tier bias at the cost of `n_iter`
inner Laplace solves.

## Which tier when

| Tier  | Function                  | Cost                            | When to reach for it |
|-------|---------------------------|---------------------------------|----------------------|
| 2 (structured) | `tulpa_nested_laplace()` | `n_grid` inner Laplace solves       | Quick fit, diagnostic, debias-target |
| 2 (structured) | `tulpa_tgmrf_vi()`       | L-BFGS + `n_draws` extra evals      | Want covariance instead of grid; IMH proposal |
| 1 (exact)      | `tulpa_tgmrf_imh()`      | `n_iter` Laplace solves             | Low theta dim, near-Gaussian posterior |
| 1 (exact)      | `tulpa_tgmrf_nuts()`     | `n_iter * 2^depth * 2 * theta_dim` Laplace solves | High theta dim or non-Gaussian |

The general rule: start with `tulpa_nested_laplace()` for a sanity
check, run `tulpa_tgmrf_imh()` for exact-tier posterior moments, and
reach for `tulpa_tgmrf_nuts()` only when IMH acceptance collapses.

## What `tgmrf()` does not require

- No gradient code. The score in `z` is closed-form from `Q`.
- No Hessian code. The Hessian in `z` is just `-Q`.
- No autodiff. Hyperparameter gradients are finite-difference.
- No C++. The R closure path reaches every tier.

The forthcoming `tgmrf_cpp()` constructor will offer a templated-C++
backend with the same S3 contract for cases where the R↔C boundary on
`Q(theta)` becomes the bottleneck (large `n_latent`, hot CCD grids).
The user-visible API does not change — the same downstream consumers
(formula parser, inference layers, methods) treat the two paths
identically.

## See also

- `?tgmrf` — constructor reference.
- `?tulpa_nested_laplace` — Tier-2 grid integrator.
- `?tulpa_tgmrf_imh`, `?tulpa_tgmrf_nuts`, `?tulpa_tgmrf_vi` — adapter family.
- `inst/examples/tgmrf_periodic_ar1.R` — runnable script form of this vignette.
